---
title: "Bipartite Network Analysis"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bipartite Network Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>",
	fig.width = 7,
	fig.height = 5,
	dpi = 150,
	out.width = "100%",
	fig.align = "center"
)
```

```{r setup}
library(lame)
library(ggplot2)
set.seed(6886)
```

## Introduction

Not all networks are created equal. Many real-world networks are **bipartite**, meaning they have two distinct types of nodes, and ties only form between types, never within them. Students enroll in courses. Countries sign treaties. Legislators join committees. Users purchase products. In each case, the adjacency matrix is rectangular rather than square, and the row and column nodes play fundamentally different roles.

The `lame` package supports bipartite networks through both `ame()` (cross-sectional) and `lame()` (longitudinal). The key difference from the unipartite case is that the model uses separate latent spaces for the two node types, connected by an interaction matrix $G$ that captures how row-node and column-node characteristics relate to each other.

## The Bipartite AME Model

In a bipartite network, the probability of a tie between row node $i$ and column node $j$ is modeled as:

$$y_{ij} = \beta'x_{ij} + a_i + b_j + u_i'Gv_j + \epsilon_{ij}$$

The additive effects work just as in the unipartite case: $a_i$ captures how "active" row node $i$ is (e.g., a student who takes many courses), and $b_j$ captures how "popular" column node $j$ is (e.g., a course with high enrollment). But the multiplicative term $u_i'Gv_j$ is different from the unipartite $u_i'v_j$. Row and column nodes live in potentially different-dimensional latent spaces ($R_\text{row}$ and $R_\text{col}$), and $G$ is a $R_\text{row} \times R_\text{col}$ interaction matrix.

How $G$ is handled differs between the cross-sectional and longitudinal entry points:

- In **`ame()`** (cross-sectional bipartite), $G$ is initialised to a rank-truncated scaled identity and **held fixed** during MCMC. The rotation and scaling of the interaction are absorbed into $U$ and $V$, so $G$ itself is not estimated; what is identified is the overall multiplicative term $U G V'$.
- In **`lame()`** (longitudinal bipartite), the default MCMC estimator samples $G$ each iteration from its conditional Gaussian via the bipartite $G$-sampler. With `dynamic_G = FALSE` a single $G$ is shared across periods; with `dynamic_G = TRUE` a separate $G_t$ is drawn per period and returned as `fit$G_cube` (see the section on `dynamic_G` below). The fast ALS path, `lame(..., method = "als")`, also supports `dynamic_G = TRUE` for normal, binary, and Poisson bipartite panels, including named panels where row or column actors enter or exit; it estimates a penalized point path for $G_t$ rather than sampling posterior draws.

In either case, the bipartite parameterisation lets row-side latent positions ($U$, shape $n_\text{row} \times R_\text{row}$) and column-side latent positions ($V$, shape $n_\text{col} \times R_\text{col}$) be chosen at different ranks, which is the substantive flexibility users get.

A few terms are absent in the bipartite setting. There is no dyadic correlation parameter $\rho$, because ties are inherently directional from row to column nodes (you can't have a reciprocal tie in a student--course network). The additive effects $a$ and $b$ also have independent variances rather than a joint covariance, since they describe fundamentally different types of actors.

### Supported families in bipartite mode

Every family that works for a square network also works for a rectangular one. The rectangular Z-samplers live in `R/rZ_bipartite.R` and treat each cell as independent given the linear predictor (there is no reciprocal-cell coupling in a bipartite graph). The supported set:

| Family    | Bipartite supported? | Notes |
|-----------|----------------------|-------|
| `normal`  | yes | the unrestricted-Gaussian baseline. |
| `binary`  | yes | probit link on tie / no-tie. |
| `tobit`   | yes | left-censored continuous outcome. |
| `ordinal` | yes | rectangular ordinal-probit sampler (`rZ_ord_bip_fc`); use data-induced cutpoints (default) or pass `ordinal_cutpoints = "explicit"` for the Cowles MH variant on the unipartite path. |
| `cbin`    | yes | constrained-binary with `odmax` interpreted as the per-row outdegree cap. |
| `frn`     | yes | fixed-rank-nominations; each row's top-`odmax` ranks are observed. |
| `rrl`     | yes | row-ranked likelihood ranks each row over column nodes; scales linearly in the row length, so long rows (`nB = 50` or larger) are routine. |
| `poisson` | yes | rectangular MH on Z; supports the `period_exposure` offset on the longitudinal path. |

The same table applies to `lame()` in bipartite mode. `dynamic_G = TRUE` is supported in `lame()`. On the default MCMC path it runs a Carter-Kohn FFBS on `vec(G_t)` under an AR(1) state-space prior and returns posterior summaries plus a rotation-drift diagnostic. On the ALS path it estimates a point path for `G_t` for the normal, binary, and Poisson families, honors separate `R_row` and `R_col` ranks, and aligns named changing-composition panels to the union row and column actor sets. Use MCMC when you need posterior draws or the rank/censored families. We demonstrate the MCMC `dynamic_G` fit later in this vignette.

### Tobit and rrl in practice

Bipartite **tobit** is the right choice when your outcome is a continuous quantity that is censored at a known boundary, such as trade volume between a country and an IGO that records "no exchange" as zero rather than as missing, or a patient's reported dose of a drug they were not prescribed (also zero). The Z-sampler treats observed zeros as left-censored draws from the underlying latent intensity and propagates that censoring into the additive / multiplicative updates. Set `family = "tobit"` and supply the matrix of recorded values (zeros allowed).

Bipartite **rrl** (row-ranked likelihood) is supported in both `ame()` and `lame()`: the rectangular Z-sampler `rZ_rrl_bip_fc` in `R/rZ_bipartite.R` enforces the per-row rank cone (`Y_ij > Y_ik` implies `Z_ij > Z_ik` within row `i`) without a zero threshold and without `odmax`. The sampler scales linearly in the row length, so long rows (`nB = 50` or larger) are routine.

## Cross-Sectional Analysis with `ame()`

### Coming from a long-format tibble?

If your data lives in a long-format tibble (`patient_id`, `drug`, `prescribed`),
pivot it to a matrix once before calling `ame()`:

```{r tibble_bridge, eval = FALSE}
library(tidyr); library(tibble)
Y_df <- rx %>%                                        # long tibble
	pivot_wider(names_from = drug, values_from = prescribed)
Y <- as.matrix(Y_df[, -1])                            # numeric matrix
rownames(Y) <- Y_df$patient_id                        # row actor names
# colnames(Y) are already the drug names from pivot_wider
```

The matrix is what `ame(..., mode = "bipartite")` consumes. Rownames /
colnames carry through to `fit$APM` (one number per patient) and `fit$BPM`
(one number per drug) downstream.

### Simulating Data

To illustrate the workflow, we simulate a bipartite network of 30 students and 20 courses. Each student has a 2-dimensional latent position that captures their academic preferences, and each course has a 2-dimensional latent position that captures what kind of student it attracts. The interaction matrix $G$ determines how these latent dimensions combine to predict enrollment.

```{r simulate_cross_sectional}
# simulate a student-course enrollment network
n_students <- 30
n_courses <- 20

# true latent positions (unobserved in practice)
U_true <- matrix(rnorm(n_students * 2, 0, 0.8), n_students, 2)
V_true <- matrix(rnorm(n_courses * 2, 0, 0.8), n_courses, 2)

# interaction matrix: dimension 1 has positive affinity,
# dimension 2 has negative (students high on dim 2 avoid courses high on dim 2)
G_true <- matrix(c(1, 0.5, 0.5, -1), 2, 2)

# generate enrollment probabilities and binary outcomes
# negative intercept keeps enrollment rate realistic (~25-30%)
eta <- -0.8 + U_true %*% G_true %*% t(V_true)
prob <- pnorm(eta)
Y_bipartite <- matrix(rbinom(n_students * n_courses, 1, prob),
											n_students, n_courses)

rownames(Y_bipartite) <- paste0("Student", 1:n_students)
colnames(Y_bipartite) <- paste0("Course", 1:n_courses)

cat("Network dimensions:", dim(Y_bipartite), "\n")
cat("Enrollment rate:", round(mean(Y_bipartite), 2), "\n")
```

### Fitting the Model

Fitting a bipartite model requires setting `mode = "bipartite"` and specifying the latent dimensions for each node type separately via `R_row` and `R_col`.

```{r fit_cross_sectional, message=FALSE, warning=FALSE}
# burn and nscan are small so the vignette builds quickly
# for real analyses, use burn >= 1000 and nscan >= 5000.
fit_cross <- ame(
	Y = Y_bipartite,
	mode = "bipartite",
	R_row = 2,              # latent dimensions for students
	R_col = 2,              # latent dimensions for courses
	family = "binary",
	burn = 100,
	nscan = 500,
	odens = 5,
	verbose = FALSE
)

summary(fit_cross)
```

The summary shows the regression coefficient (just an intercept here, since we included no covariates) and the variance components. The fitted object also stores the student latent positions (`U`), course latent positions (`V`), the (fixed) interaction matrix (`G`), and additive effects (`APM` for student activity, `BPM` for course popularity), which we'll visualize next.

### Visualizing the Latent Space

The `uv_plot` function with `layout = "biplot"` places both row and column nodes in the same latent space, making it easy to see which students cluster near which courses. Students and courses that appear close together in this plot have a higher predicted probability of a tie.

```{r visualize_cross_sectional, fig.height=5, fig.alt="Bipartite latent-space biplot placing students and courses in the same 2D coordinate system, with proximity indicating a higher predicted enrollment probability."}
uv_plot(fit_cross, layout = "biplot") +
	ggtitle("Bipartite Latent Space: Students and Courses")
```

The interaction matrix $G$ mediates how the latent dimensions of students and
courses combine to predict ties. In **`ame()`** (the cross-sectional path used in
this fit) $G$ is held **fixed** at a rank-truncated scaled identity: the rotation
and scaling of the interaction are absorbed into $U$ and $V$, so $G$ itself is
not estimated and its entries carry no information. What *is* estimated is the
overall multiplicative term $U G V'$, which captures the latent structure of the
network. (In `lame()` $G$ is sampled rather than fixed; see the next section.)
We visualise the fitted latent structure $U G V'$ directly:

```{r interaction_heatmap, fig.height=3, fig.alt="Heatmap of the fitted multiplicative structure U G V' across student-course pairs using a diverging colour-blind-safe ColorBrewer RdBu palette: red tiles mark positive latent affinity (predicted enrollment above baseline), blue tiles mark negative latent affinity (predicted below baseline), and near-white tiles mark a posterior estimate near zero."}
G_mult <- fit_cross$U %*% fit_cross$G %*% t(fit_cross$V)
G_df <- data.frame(
	Student = rownames(G_mult)[row(G_mult)],
	Course = colnames(G_mult)[col(G_mult)],
	Fitted = as.vector(G_mult)
)
# rdbu reversed (direction = -1): positive => red, negative => blue,
# zero => white. By default scale_fill_distiller() centres white at the
# midpoint of the DATA range, not at zero -- so we pass symmetric limits
# c(-mx, mx) to force the white midpoint onto zero, which is the right
# choice for a signed effect. RdBu is colour-blind-safe per ColorBrewer's
# protanopia / deuteranopia checks.
mx <- max(abs(G_df$Fitted), na.rm = TRUE)
ggplot(G_df, aes(x = Course, y = Student, fill = Fitted)) +
	geom_tile() +
	scale_fill_distiller(palette = "RdBu", direction = -1, limits = c(-mx, mx)) +
	labs(title = "Fitted Multiplicative Structure (UGV')",
			subtitle = "Red = positive latent affinity, Blue = negative, near-white = ~0",
			x = "Course", y = "Student",
			fill = "U G V'") +
	theme_bw() +
	theme(panel.border = element_blank())
```

## Longitudinal Analysis with `lame()`

When bipartite networks are observed over multiple time periods (say, student enrollments across semesters, or country--treaty memberships across decades), `lame()` lets you pool information across time while optionally allowing the latent structure to evolve.

### Simulating Longitudinal Data

We simulate a panel of user--item interaction networks over 5 periods in which the latent positions drift via an AR(1) process with $\rho = 0.9$. This is a smooth, mean-reverting evolution: each period's latent position is mostly the previous one plus a small innovation. The simulation chunk below uses the **stationary-variance parameterisation** ($U_t = \rho\,U_{t-1} + \sqrt{1 - \rho^2}\,\epsilon$ with $\epsilon \sim N(0, I)$, so the marginal variance of $U_t$ stays equal to 1 for every $t$). The MCMC sampler internally uses the **innovation-variance parameterisation** ($U_t = \rho\,U_{t-1} + \epsilon$ with $\epsilon \sim N(0, \sigma_{uv}^2)$); the two are the same process when $\sigma_{uv}^2 = 1 - \rho^2$, but the variance lives on different objects, so don't mix them when writing your own simulations. The data-generating interaction matrix $G$ that combines $U$ and $V$ is held constant across periods, so when we fit `lame()` with `dynamic_G = FALSE` the model's static $G$ is matched to the truth; we show the time-varying alternative (`dynamic_G = TRUE`) further down.

```{r simulate_longitudinal}
n_periods <- 5
n_users <- 20
n_items <- 15

true_rho <- 0.9
G_long <- matrix(c(1, 0.3, 0.3, -0.8), 2, 2)

U_t <- vector("list", n_periods)
V_t <- vector("list", n_periods)
U_t[[1]] <- matrix(rnorm(n_users * 2), n_users, 2)
V_t[[1]] <- matrix(rnorm(n_items * 2), n_items, 2)
for(t in 2:n_periods) {
	U_t[[t]] <- true_rho * U_t[[t-1]] +
		sqrt(1 - true_rho^2) * matrix(rnorm(n_users * 2), n_users, 2)
	V_t[[t]] <- true_rho * V_t[[t-1]] +
		sqrt(1 - true_rho^2) * matrix(rnorm(n_items * 2), n_items, 2)
}

Y_list <- list()
for(t in 1:n_periods) {
	eta_t <- U_t[[t]] %*% G_long %*% t(V_t[[t]])
	prob_t <- pnorm(eta_t)
	Y_list[[t]] <- matrix(rbinom(n_users * n_items, 1, prob_t),
												n_users, n_items)
	rownames(Y_list[[t]]) <- paste0("User", 1:n_users)
	colnames(Y_list[[t]]) <- paste0("Item", 1:n_items)
}
names(Y_list) <- paste0("T", 1:n_periods)

cat("Time periods:", length(Y_list), "\n")
cat("Dimensions per period:", dim(Y_list[[1]]), "\n")
cat("Average density:", round(mean(sapply(Y_list, mean)), 2), "\n")
cat("True latent AR(1) coefficient:", true_rho, "\n")
```

### Static Model

The simplest longitudinal model assumes that the latent positions and additive effects are constant over time. This pools all time periods together to get a single estimate of each actor's latent position, useful when you believe the underlying structure is stable and you just want more data to estimate it precisely.

```{r fit_longitudinal_static, message=FALSE, warning=FALSE}
# iterations are small so the vignette builds quickly; use burn >= 1000
# and nscan >= 5000 for real analyses.
fit_static <- lame(
	Y = Y_list,
	mode = "bipartite",
	R_row = 2,
	R_col = 2,
	family = "binary",
	dynamic_uv = FALSE,
	dynamic_ab = FALSE,
	burn = 100,
	nscan = 500,
	odens = 5,
	verbose = FALSE,
	plot = FALSE
)

summary(fit_static)
```

### Dynamic Model

When you suspect the latent structure changes over time (users' tastes shift, items gain or lose popularity), the dynamic model lets the latent positions and additive effects drift via AR(1) processes. A high autoregressive parameter ($\rho$ close to 1) means slow, gradual change; a low one means the structure is more volatile from period to period.

```{r fit_longitudinal_dynamic, message=FALSE, warning=FALSE}
# same reduced iterations as above for vignette speed.
fit_dynamic <- lame(
	Y = Y_list,
	mode = "bipartite",
	R_row = 2,
	R_col = 2,
	family = "binary",
	dynamic_uv = TRUE,
	dynamic_ab = TRUE,
	burn = 100,
	nscan = 500,
	odens = 5,
	verbose = FALSE,
	plot = FALSE
)

summary(fit_dynamic)
```

The dynamic model reports estimated AR(1) persistence parameters for the latent positions (`rho_uv`) and additive effects (`rho_ab`). Values close to 1 indicate strong period-to-period persistence; values close to 0 indicate near-independent latent positions across periods.

```{r dynamic_rho_recovery}
cat(sprintf("True rho_uv = %.2f | posterior mean = %.3f | 95%% CI = [%.3f, %.3f]\n",
						true_rho,
						mean(fit_dynamic$rho_uv),
						quantile(fit_dynamic$rho_uv, 0.025),
						quantile(fit_dynamic$rho_uv, 0.975)))
```

The posterior recovers the true `rho_uv = 0.9` within the 95% credible interval, indicating that `lame()` correctly identifies that the latent positions drift smoothly rather than refreshing each period.

### Time-varying Interaction Matrix (`dynamic_G`)

When you suspect that the way the row and column latent spaces interact is itself shifting over time, not just the actor positions, set `dynamic_G = TRUE`. By default, `lame()` runs a Carter-Kohn FFBS on `vec(G_t)` under an AR(1) state-space prior (the persistence and innovation variance are sampled jointly inside the loop) and attaches three outputs:

- `fit$G_cube` -- the final draw's per-period $G_t$ ($R_\text{row} \times R_\text{col} \times T$).
- `fit$G_cube_post_mean` -- the posterior-mean $G_t$ averaged across all stored draws, same shape; this is the recommended summary.
- `fit$G_cube_post_sd` -- the elementwise posterior SD across draws, same shape.

The hyperparameter chains live at `fit$RHO_G` and `fit$SIGMA_G2`. A rotation-drift diagnostic (`fit$G_rotation_drift`) compares the variance of the raw $G_t$ entries to their SVD-canonicalised counterparts; when the ratio is large the apparent time variation is dominated by latent-rotation drift rather than real change in the interaction matrix.

With `method = "als"`, `dynamic_G = TRUE` estimates the same per-period bilinear surface as a penalized point path for normal, binary, and Poisson panels. Named changing-composition panels are aligned to the union row and column actor sets before fitting. The object still stores `G_cube`, `G_cube_post_mean`, and `G_cube_post_sd` for compatibility with MCMC code; `G_cube_post_mean` is the point path and `G_cube_post_sd` is zero because ALS has no posterior draws. The smoothing settings are reported through `rho_G`, `RHO_G`, `SIGMA_G2`, and `lambda_G_als`.

```{r fit_dynamic_G, message=FALSE, warning=FALSE}
fit_dynG <- lame(
	Y = Y_list,
	mode = "bipartite",
	R_row = 2,
	R_col = 2,
	family = "binary",
	dynamic_uv = TRUE,
	dynamic_ab = TRUE,
	dynamic_G = TRUE,
	burn = 100,
	nscan = 500,
	odens = 5,
	verbose = FALSE,
	plot = FALSE
)

# fit$G_cube_post_mean has dimensions R_row x R_col x n_periods
if (!is.null(fit_dynG$G_cube_post_mean)) {
	cat("G_cube_post_mean dimensions:", dim(fit_dynG$G_cube_post_mean), "\n")
	cat("Frobenius norm of posterior-mean G_t per period:\n")
	print(round(apply(fit_dynG$G_cube_post_mean, 3,
	                  function(g) sqrt(sum(g^2))), 3))
	cat("Rotation-drift ratio (>= 5 flags latent-rotation domination):",
	    round(fit_dynG$G_rotation_drift$ratio, 2), "\n")
} else {
	cat("G_cube was not attached (no bipartite + RA>0 + RB>0 case).\n")
}
```

If the Frobenius norms across periods are nearly identical (and the static-G fit GOF is comparable), there is no real evidence of time-varying interaction structure and you should stick with `dynamic_G = FALSE`. If they drift substantially and the static-G fit shows worse GOF coverage, the time-varying $G_t$ is doing meaningful work. The rotation-drift diagnostic is the second sanity check: if the ratio is above ~5, the apparent variation across $G_t$ slices is mostly U/V rotation rather than real temporal change, and the canonical reporting in `G_cube_post_mean` is the right summary.

### Visualizing Temporal Evolution

With dynamic effects, `uv_plot` can show trajectories, illustrating how each actor's latent position moves through the space over time. Each line traces one actor's path from the first to the last period, revealing which actors' roles are stable and which are shifting.

```{r visualize_longitudinal, fig.height=5, fig.alt="Trajectory plot showing the path of each actor's latent position from the first to the last period; coherent curved paths indicate smooth temporal evolution."}
uv_plot(fit_dynamic, plot_type = "trajectory") +
	ggtitle("Dynamic Latent Positions: Trajectories Over Time")
```

Convergence diagnostics are especially important for dynamic models, since the additional parameters ($\rho$, innovation variances) need adequate samples to be well-estimated.

```{r trace_dynamic, fig.height=6, fig.alt="MCMC trace plots and posterior densities for the dynamic bipartite model parameters; well-mixed caterpillar traces indicate adequate sampling."}
trace_plot(fit_dynamic)
```

## Comparing Static and Dynamic Models

A natural question is whether the dynamic model is worth the added complexity. The goodness-of-fit plots provide one way to compare: if the dynamic model's posterior-predictive distributions better cover the observed statistics, the extra flexibility is paying off. In each panel below the **observed bipartite statistic at each period appears as a solid Okabe-Ito orange (`#D55E00`) line with filled points**, the **posterior-predictive median is a dashed dark grey line**, and the **grey ribbon is the 95% posterior-predictive credible interval**; the three series are dual-encoded on colour and linetype so the comparison stays legible in greyscale and for colour-blind readers. For bipartite fits the three panels are labelled "Sender Degree Heterogeneity" (row-mean SD), "Receiver Degree Heterogeneity" (column-mean SD), and "Four Cycles", the bipartite analogue of transitivity.

```{r model_comparison, fig.height=4, fig.alt="Side-by-side goodness-of-fit panels comparing posterior predictive coverage of bipartite degree heterogeneity and four-cycle counts for the static versus dynamic model."}
# gof plots for each model
gof_plot(fit_static)
gof_plot(fit_dynamic)
```

We can also compare the GOF statistics directly. For bipartite networks, the key statistics are the standard deviation of row means (how much users vary in activity), the standard deviation of column means (how much items vary in popularity), and the four-cycle count (a measure of clustering in bipartite networks, analogous to transitivity in unipartite networks).

```{r model_comparison_boxplot, fig.height=4, fig.alt="Boxplots of posterior predictive bipartite statistics (row-mean SD, column-mean SD, four cycles) faceted by static and dynamic models, with horizontal reference lines at observed values."}
gof_static <- fit_static$GOF
gof_dynamic <- fit_dynamic$GOF

# extract posterior predictive samples (exclude column 1, which is observed)
# each element is a matrix [n_time x n_mcmc], so colMeans averages across time
extract_ppc <- function(gof, stat) {
	mat <- gof[[stat]]
	colMeans(mat[, -1, drop = FALSE])
}

# observed values (column 1, averaged across time periods)
obs_vals <- c(
	mean(gof_static$sd.rowmean[, 1]),
	mean(gof_static$sd.colmean[, 1]),
	mean(gof_static$four.cycles[, 1])
)

gof_df <- data.frame(
	value = c(
		extract_ppc(gof_static, "sd.rowmean"),
		extract_ppc(gof_dynamic, "sd.rowmean"),
		extract_ppc(gof_static, "sd.colmean"),
		extract_ppc(gof_dynamic, "sd.colmean"),
		extract_ppc(gof_static, "four.cycles"),
		extract_ppc(gof_dynamic, "four.cycles")
	),
	model = rep(rep(c("Static", "Dynamic"),
							each = ncol(gof_static$sd.rowmean) - 1), 3),
	statistic = rep(c("Row Mean SD", "Column Mean SD", "Four-Cycles"),
									each = 2 * (ncol(gof_static$sd.rowmean) - 1))
)

obs_df <- data.frame(
	statistic = c("Row Mean SD", "Column Mean SD", "Four-Cycles"),
	observed = obs_vals
)

ggplot(gof_df, aes(x = model, y = value)) +
	geom_boxplot() +
	geom_hline(data = obs_df, aes(yintercept = observed),
						linetype = 2) +
	facet_wrap(~statistic, scales = "free_y") +
	labs(title = "GOF Comparison: Static vs Dynamic",
			subtitle = "Dashed Line = Observed Value",
			x = "Model", y = "Simulated Statistic") +
	theme_bw() +
	theme(
		panel.border = element_blank(),
		strip.background = element_rect(fill = "black", color = "black"),
		strip.text = element_text(color = "white", hjust = 0)
	)
```

The simulated interaction matrix $G$ is constant, while actor positions drift via an AR(1) with $\rho = 0.9$ -- moderate persistence (the lag-4 autocorrelation is $0.9^4 \approx 0.66$, so the positions at the first and last period correlate only about 0.6-0.8 and do move). What is genuinely fixed is $G$, not the positions; combined with the short panel and short chains, that is why the static and dynamic models fit similarly here rather than the dynamic one clearly winning. If the boxplots overlap substantially and both cover the dashed line, the dynamic model's extra flexibility is not paying off here. In real applications where user preferences or item popularity genuinely shift over time, the dynamic model would show tighter coverage of the observed statistics.

### Choosing Latent Dimensions

The number of latent dimensions (`R_row`, `R_col`) controls how rich the multiplicative structure is. Too few dimensions and the model can't capture the latent clustering; too many and you risk overfitting (and slower computation). A practical approach is to fit models with different dimensions and compare their GOF statistics.

```{r dimension_selection, message=FALSE, warning=FALSE}
# include R=0 as a no-latent-space baseline
dims_to_test <- list(
	c(0, 0),
	c(1, 1),
	c(2, 2)
)

# compute observed GOF statistics from the data
obs_gof <- gof_stats(Y_bipartite, mode = "bipartite")

gof_results <- list()
for(i in seq_along(dims_to_test)) {
	fit_temp <- ame(
		Y = Y_bipartite,
		mode = "bipartite",
		R_row = dims_to_test[[i]][1],
		R_col = dims_to_test[[i]][2],
		family = "binary",
		burn = 200,
		nscan = 2500,
		odens = 5,
		verbose = FALSE
	)
	# posterior predictive tail probability: proportion of simulated
	# statistics at or above the observed value (one-sided)
	gof_results[[i]] <- c(
		R_row = dims_to_test[[i]][1],
		R_col = dims_to_test[[i]][2],
		pval_rowmean = mean(fit_temp$GOF[, "sd.rowmean"] >= obs_gof["sd.rowmean"]),
		pval_colmean = mean(fit_temp$GOF[, "sd.colmean"] >= obs_gof["sd.colmean"]),
		pval_fourcycles = mean(fit_temp$GOF[, "four.cycles"] >= obs_gof["four.cycles"])
	)
}

do.call(rbind, gof_results)
```

These are one-sided posterior predictive tail probabilities, not formal test
p-values. Values near 0.5 indicate the observed statistic lies in the bulk of
the model's posterior predictive distribution; values near 0 or 1 indicate
the model systematically over- or under-predicts that feature. (Note that for
`ame()`, `fit$GOF` stores the observed statistic in row 1 alongside the
posterior-predictive draws, so the tail probability here includes the
observed point itself — a one-row bias that is negligible for the chain
lengths we use, but worth knowing.) The
relationship to `R` is not monotonic; adding latent dimensions can move
some tail probabilities away from 0.5 (e.g. by making the predictive
distribution more dispersed), so use this together with the trajectory and
GOF plots rather than as a single number to optimise. A sensible heuristic
is to pick the smallest `R` at which all three tail probabilities sit
comfortably away from 0 and 1. On this particular simulated network the
three tail probabilities all stay high (near 1) at every `R` -- the fit
over-predicts both row-degree heterogeneity and four-cycles (the simulated
networks carry more of each than the observed network) regardless of
rank -- which is itself a useful verdict: no value of `R` rescues these
statistics, so the limitation is structural rather than a matter of
adding latent dimensions. That is exactly the kind of read this diagnostic
is for.

## Practical Guidance

**When to use bipartite models.** Use bipartite mode whenever your network has two fundamentally different types of nodes and ties only form between types. If your adjacency matrix is rectangular, you almost certainly need it. Even if it happens to be square (e.g., 20 students and 20 courses), set `mode = "bipartite"` if the row and column nodes represent different kinds of entities.

**Interpreting the output.** The additive effects (`APM`, `BPM`) have the most direct interpretation: they tell you which row nodes are unusually active and which column nodes are unusually popular, after controlling for covariates and latent structure. The latent positions (`U`, `V`) and interaction matrix (`G`) together capture residual patterns of association, answering the question "beyond the covariates and overall activity levels, which row nodes tend to connect to which column nodes?"

## Convergence Diagnostics

As with any Bayesian model, always check convergence before interpreting results. The `trace_plot` function shows MCMC traces and posterior densities for the regression coefficients and variance components.

```{r convergence, fig.height=6, fig.alt="MCMC trace and density plots for the cross-sectional bipartite fit; well-mixed traces and smooth unimodal densities support convergence."}
trace_plot(fit_cross)
```

Look for traces that mix well (no long flat stretches or slow drifts) and densities that are smooth and unimodal. With the short chains used in this vignette, convergence is not guaranteed. In practice, run longer chains and consider multiple starting values.

For numerical diagnostics use the same Stan-era stack the cross-sectional vignette demonstrates. `posterior::as_draws(fit_cross)` returns a `draws_array` keyed by chain, and `posterior::summarise_draws()` reports split-$\hat R$, `ess_bulk`, and `ess_tail` per coefficient and variance component. For honest between-chain $\hat R$ run `ame_parallel(..., n_chains = 4, fitter = "ame", combine_method = "pool")` so the pooled fit carries a `chain_indicator` column; `summarise_draws()` then computes Rhat across chains. `prior_summary(fit_cross)` prints the hyperparameters actually in effect (g-prior on $\beta$, variance-component priors), and `loo::loo(fit_cross)` works after refitting with `save_log_lik = TRUE` (Gaussian density for `normal`, probit Bernoulli for `binary`, log-Poisson for `poisson`; pair with `loo::loo_compare()` for model selection). See the [cross-sectional vignette](cross_sec_ame.html) for the worked end-to-end stack.

## Extracting Positions for Custom Analysis

If you need the latent positions in a tidy format (e.g., for merging with external data or building custom plots), the `latent_positions()` function returns a data frame with one row per actor-dimension-time combination:

```{r}
lp <- latent_positions(fit_cross)
head(lp)

# filter to just the row nodes (students)
lp_students <- lp[lp$type == "U", ]
cat("Student positions:", nrow(lp_students), "rows\n")
```

This is especially useful for bipartite networks where U (row nodes) and V (column nodes) have different numbers of actors.
