---
title: "Bipartite Network Analysis"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bipartite Network Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>",
	fig.width = 7,
	fig.height = 5,
	dpi = 150,
	out.width = "100%",
	fig.align = "center"
)
```

```{r setup}
library(lame)
library(ggplot2)
library(reshape2)
set.seed(6886)
```

## Introduction

Not all networks are created equal. Many real-world networks are **bipartite**, meaning they have two distinct types of nodes, and ties only form between types, never within them. Students enroll in courses. Countries sign treaties. Legislators join committees. Users purchase products. In each case, the adjacency matrix is rectangular rather than square, and the row and column nodes play fundamentally different roles.

The `lame` package supports bipartite networks through both `ame()` (cross-sectional) and `lame()` (longitudinal). The key difference from the unipartite case is that the model uses separate latent spaces for the two node types, connected by an interaction matrix $G$ that captures how row-node and column-node characteristics relate to each other.

## The Bipartite AME Model

In a bipartite network, the probability of a tie between row node $i$ and column node $j$ is modeled as:

$$y_{ij} = \beta'x_{ij} + a_i + b_j + u_i'Gv_j + \epsilon_{ij}$$

The additive effects work just as in the unipartite case: $a_i$ captures how "active" row node $i$ is (e.g., a student who takes many courses), and $b_j$ captures how "popular" column node $j$ is (e.g., a course with high enrollment). But the multiplicative term $u_i'Gv_j$ is different. Instead of $u_i'v_j$ with shared latent dimensions, the row and column nodes live in potentially different-dimensional latent spaces ($R_\text{row}$ and $R_\text{col}$), and the interaction matrix $G$ maps between them. This is more flexible, letting the model discover that, say, two dimensions of student preferences map onto three dimensions of course characteristics.

A few things drop away in the bipartite setting. There is no dyadic correlation parameter $\rho$, because ties are inherently directional from row to column nodes (you can't have a reciprocal tie in a student--course network). The additive effects $a$ and $b$ also have independent variances rather than a joint covariance, since they describe fundamentally different types of actors.

## Cross-Sectional Analysis with `ame()`

### Simulating Data

To illustrate the workflow, we simulate a bipartite network of 30 students and 20 courses. Each student has a 2-dimensional latent position that captures their academic preferences, and each course has a 2-dimensional latent position that captures what kind of student it attracts. The interaction matrix $G$ determines how these latent dimensions combine to predict enrollment.

```{r simulate_cross_sectional}
# simulate a student-course enrollment network
n_students <- 30
n_courses <- 20

# true latent positions (unobserved in practice)
U_true <- matrix(rnorm(n_students * 2, 0, 0.8), n_students, 2)
V_true <- matrix(rnorm(n_courses * 2, 0, 0.8), n_courses, 2)

# interaction matrix: dimension 1 has positive affinity,
# dimension 2 has negative (students high on dim 2 avoid courses high on dim 2)
G_true <- matrix(c(1, 0.5, 0.5, -1), 2, 2)

# generate enrollment probabilities and binary outcomes
# negative intercept keeps enrollment rate realistic (~25-30%)
eta <- -0.8 + U_true %*% G_true %*% t(V_true)
prob <- pnorm(eta)
Y_bipartite <- matrix(rbinom(n_students * n_courses, 1, prob),
											n_students, n_courses)

rownames(Y_bipartite) <- paste0("Student", 1:n_students)
colnames(Y_bipartite) <- paste0("Course", 1:n_courses)

cat("Network dimensions:", dim(Y_bipartite), "\n")
cat("Enrollment rate:", round(mean(Y_bipartite), 2), "\n")
```

### Fitting the Model

Fitting a bipartite model requires setting `mode = "bipartite"` and specifying the latent dimensions for each node type separately via `R_row` and `R_col`.

```{r fit_cross_sectional, message=FALSE, warning=FALSE}
# note: burn/nscan are kept small here for fast vignette building.
# for real analyses, use burn >= 1000 and nscan >= 5000.
fit_cross <- ame(
	Y = Y_bipartite,
	mode = "bipartite",
	R_row = 2,              # latent dimensions for students
	R_col = 2,              # latent dimensions for courses
	family = "binary",
	burn = 100,
	nscan = 500,
	odens = 5,
	verbose = FALSE
)

summary(fit_cross)
```

The summary shows the regression coefficient (just an intercept here, since we included no covariates) and the variance components. The fitted object also stores the student latent positions (`U`), course latent positions (`V`), the interaction matrix (`G`), and additive effects (`APM` for student activity, `BPM` for course popularity), which we'll visualize next.

### Visualizing the Latent Space

The `uv_plot` function with `layout = "biplot"` places both row and column nodes in the same latent space, making it easy to see which students cluster near which courses. Students and courses that appear close together in this plot have a higher predicted probability of a tie.

```{r visualize_cross_sectional, fig.height=5}
uv_plot(fit_cross, layout = "biplot") +
	ggtitle("Bipartite Latent Space: Students and Courses")
```

The interaction matrix $G$ mediates how the latent dimensions of students and courses combine to predict ties. Keep in mind that $G$, $U$, and $V$ are only identified up to rotation and scaling: the model can absorb rotations of $U$ and $V$ into $G$, so the individual entries of $G$ do not have a stable interpretation across runs. What is identified is the overall product $UGV'$, which captures the latent structure of the network. The diagonal entries of $G$ indicate the overall strength of the latent factor association, while off-diagonal entries indicate cross-dimension interactions.

```{r interaction_heatmap, fig.height=3}
G_melt <- melt(fit_cross$G)
ggplot(G_melt, aes(x = Var2, y = Var1, fill = value)) +
	geom_tile() +
	scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
	labs(title = "Estimated Interaction Matrix G",
			x = "Column Dimension", y = "Row Dimension") +
	theme_bw() +
	theme(panel.border = element_blank())
```

## Longitudinal Analysis with `lame()`

When bipartite networks are observed over multiple time periods (say, student enrollments across semesters, or country--treaty memberships across decades), `lame()` lets you pool information across time while optionally allowing the latent structure to evolve.

### Simulating Longitudinal Data

We simulate a panel of user--item interaction networks over 5 periods. The underlying latent structure is held fixed (for now), with small random shocks to the overall baseline at each period.

```{r simulate_longitudinal}
n_periods <- 5
n_users <- 20
n_items <- 15

U_long <- matrix(rnorm(n_users * 2), n_users, 2)
V_long <- matrix(rnorm(n_items * 2), n_items, 2)
G_long <- matrix(c(1, 0.3, 0.3, -0.8), 2, 2)

Y_list <- list()
for(t in 1:n_periods) {
	eta_t <- U_long %*% G_long %*% t(V_long) + rnorm(1, 0, 0.2)
	prob_t <- pnorm(eta_t)
	Y_list[[t]] <- matrix(rbinom(n_users * n_items, 1, prob_t),
												n_users, n_items)
	rownames(Y_list[[t]]) <- paste0("User", 1:n_users)
	colnames(Y_list[[t]]) <- paste0("Item", 1:n_items)
}
names(Y_list) <- paste0("T", 1:n_periods)

cat("Time periods:", length(Y_list), "\n")
cat("Dimensions per period:", dim(Y_list[[1]]), "\n")
cat("Average density:", round(mean(sapply(Y_list, mean)), 2), "\n")
```

### Static Model

The simplest longitudinal model assumes that the latent positions and additive effects are constant over time. This pools all time periods together to get a single estimate of each actor's latent position, useful when you believe the underlying structure is stable and you just want more data to estimate it precisely.

```{r fit_longitudinal_static, message=FALSE, warning=FALSE}
# note: iterations are kept small for vignette speed; use burn >= 1000
# and nscan >= 5000 for real analyses.
fit_static <- lame(
	Y = Y_list,
	mode = "bipartite",
	R_row = 2,
	R_col = 2,
	family = "binary",
	dynamic_uv = FALSE,
	dynamic_ab = FALSE,
	burn = 100,
	nscan = 500,
	odens = 5,
	verbose = FALSE,
	plot = FALSE
)

summary(fit_static)
```

### Dynamic Model

When you suspect the latent structure changes over time (users' tastes shift, items gain or lose popularity), the dynamic model lets the latent positions and additive effects drift via AR(1) processes. A high autoregressive parameter ($\rho$ close to 1) means slow, gradual change; a low one means the structure is more volatile from period to period.

```{r fit_longitudinal_dynamic, message=FALSE, warning=FALSE}
# same reduced iterations as above for vignette speed.
fit_dynamic <- lame(
	Y = Y_list,
	mode = "bipartite",
	R_row = 2,
	R_col = 2,
	family = "binary",
	dynamic_uv = TRUE,
	dynamic_ab = TRUE,
	burn = 100,
	nscan = 500,
	odens = 5,
	verbose = FALSE,
	plot = FALSE
)

summary(fit_dynamic)
```

The dynamic model reports estimated AR(1) persistence parameters for the latent positions (`rho_uv`) and additive effects (`rho_ab`). Since our simulated data uses the same latent structure across all time periods, these parameters reflect the interplay between the prior (which favors persistence) and the lack of temporal variation in the data. In real applications where user preferences or item popularity genuinely shift over time, you would see these parameters more clearly informed by the data.

### Visualizing Temporal Evolution

With dynamic effects, `uv_plot` can show trajectories, illustrating how each actor's latent position moves through the space over time. Each line traces one actor's path from the first to the last period, revealing which actors' roles are stable and which are shifting.

```{r visualize_longitudinal, fig.height=5}
uv_plot(fit_dynamic, plot_type = "trajectory") +
	ggtitle("Dynamic Latent Positions: Trajectories Over Time")
```

Convergence diagnostics are especially important for dynamic models, since the additional parameters ($\rho$, innovation variances) need adequate samples to be well-estimated.

```{r trace_dynamic, fig.height=6}
trace_plot(fit_dynamic)
```

## Comparing Static and Dynamic Models

A natural question is whether the dynamic model is worth the added complexity. The goodness-of-fit plots provide one way to compare: if the dynamic model's posterior predictive distributions better cover the observed statistics, the extra flexibility is paying off.

```{r model_comparison, fig.height=4}
# GOF plots for each model
gof_plot(fit_static)
gof_plot(fit_dynamic)
```

We can also compare the GOF statistics directly. For bipartite networks, the key statistics are the standard deviation of row means (how much users vary in activity), the standard deviation of column means (how much items vary in popularity), and the four-cycle count (a measure of clustering in bipartite networks, analogous to transitivity in unipartite networks).

```{r model_comparison_boxplot, fig.height=4}
gof_static <- fit_static$GOF
gof_dynamic <- fit_dynamic$GOF

# extract posterior predictive samples (exclude column 1, which is observed)
# each element is a matrix [n_time x n_mcmc], so colMeans averages across time
extract_ppc <- function(gof, stat) {
	mat <- gof[[stat]]
	colMeans(mat[, -1, drop = FALSE])
}

# observed values (column 1, averaged across time periods)
obs_vals <- c(
	mean(gof_static$sd.rowmean[, 1]),
	mean(gof_static$sd.colmean[, 1]),
	mean(gof_static$four.cycles[, 1])
)

gof_df <- data.frame(
	value = c(
		extract_ppc(gof_static, "sd.rowmean"),
		extract_ppc(gof_dynamic, "sd.rowmean"),
		extract_ppc(gof_static, "sd.colmean"),
		extract_ppc(gof_dynamic, "sd.colmean"),
		extract_ppc(gof_static, "four.cycles"),
		extract_ppc(gof_dynamic, "four.cycles")
	),
	model = rep(rep(c("Static", "Dynamic"),
							each = ncol(gof_static$sd.rowmean) - 1), 3),
	statistic = rep(c("Row Mean SD", "Column Mean SD", "Four-Cycles"),
									each = 2 * (ncol(gof_static$sd.rowmean) - 1))
)

obs_df <- data.frame(
	statistic = c("Row Mean SD", "Column Mean SD", "Four-Cycles"),
	observed = obs_vals
)

ggplot(gof_df, aes(x = model, y = value)) +
	geom_boxplot() +
	geom_hline(data = obs_df, aes(yintercept = observed),
						linetype = 2) +
	facet_wrap(~statistic, scales = "free_y") +
	labs(title = "GOF Comparison: Static vs Dynamic",
			subtitle = "Dashed Line = Observed Value",
			x = "Model", y = "Simulated Statistic") +
	theme_bw() +
	theme(
		panel.border = element_blank(),
		strip.background = element_rect(fill = "black", color = "black"),
		strip.text = element_text(color = "white", hjust = 0)
	)
```

Since we simulated data with fixed latent structure across all time periods, the static and dynamic models should fit similarly. If the boxplots overlap substantially and both cover the dashed line, the dynamic model's extra flexibility is not paying off here. In real applications where user preferences or item popularity genuinely shift over time, the dynamic model would show tighter coverage of the observed statistics.

### Choosing Latent Dimensions

The number of latent dimensions (`R_row`, `R_col`) controls how rich the multiplicative structure is. Too few dimensions and the model can't capture the latent clustering; too many and you risk overfitting (and slower computation). A practical approach is to fit models with different dimensions and compare their GOF statistics.

```{r dimension_selection, message=FALSE, warning=FALSE}
# include R=0 as a no-latent-space baseline
dims_to_test <- list(
	c(0, 0),
	c(1, 1),
	c(2, 2)
)

# compute observed GOF statistics from the data
obs_gof <- gof_stats(Y_bipartite, mode = "bipartite")

gof_results <- list()
for(i in seq_along(dims_to_test)) {
	fit_temp <- ame(
		Y = Y_bipartite,
		mode = "bipartite",
		R_row = dims_to_test[[i]][1],
		R_col = dims_to_test[[i]][2],
		family = "binary",
		burn = 200,
		nscan = 2500,
		odens = 5,
		verbose = FALSE
	)
	# posterior predictive p-values: proportion of simulated
	# statistics as or more extreme than the observed value
	gof_results[[i]] <- c(
		R_row = dims_to_test[[i]][1],
		R_col = dims_to_test[[i]][2],
		pval_rowmean = mean(fit_temp$GOF[, "sd.rowmean"] >= obs_gof["sd.rowmean"]),
		pval_fourcycles = mean(fit_temp$GOF[, "four.cycles"] >= obs_gof["four.cycles"])
	)
}

do.call(rbind, gof_results)
```

The posterior predictive p-values tell you how well each model reproduces the observed network structure. Values near 0.5 indicate good fit (the observed statistic falls in the middle of the simulated distribution). Values near 0 or 1 indicate the model consistently over- or under-predicts that feature. Look for the point where adding more dimensions stops improving the p-values.

## Practical Guidance

**When to use bipartite models.** Use bipartite mode whenever your network has two fundamentally different types of nodes and ties only form between types. If your adjacency matrix is rectangular, you almost certainly need it. Even if it happens to be square (e.g., 20 students and 20 courses), set `mode = "bipartite"` if the row and column nodes represent different kinds of entities.

**Interpreting the output.** The additive effects (`APM`, `BPM`) have the most direct interpretation: they tell you which row nodes are unusually active and which column nodes are unusually popular, after controlling for covariates and latent structure. The latent positions (`U`, `V`) and interaction matrix (`G`) together capture residual patterns of association, answering the question "beyond the covariates and overall activity levels, which row nodes tend to connect to which column nodes?"

## Convergence Diagnostics

As with any Bayesian model, always check convergence before interpreting results. The `trace_plot` function shows MCMC traces and posterior densities for the regression coefficients and variance components.

```{r convergence, fig.height=6}
trace_plot(fit_cross)
```

Look for traces that mix well (no long flat stretches or slow drifts) and densities that are smooth and unimodal. With the short chains used in this vignette, convergence is not guaranteed. In practice, run longer chains and consider multiple starting values.

## Extracting Positions for Custom Analysis

If you need the latent positions in a tidy format (e.g., for merging with external data or building custom plots), the `latent_positions()` function returns a data frame with one row per actor-dimension-time combination:

```{r}
lp <- latent_positions(fit_cross)
head(lp)

# filter to just the row nodes (students)
lp_students <- lp[lp$type == "U", ]
cat("Student positions:", nrow(lp_students), "rows\n")
```

This is especially useful for bipartite networks where U (row nodes) and V (column nodes) have different numbers of actors.
