---
title: "Your First AME Model"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Your First AME Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>",
	fig.align = "center"
)
```

## Why Network Models?

Imagine you're studying friendships in a high school. You have data on who nominated whom as a friend, plus information about each student (gender, race, grade). A natural first instinct is to run a logistic regression: does sharing the same gender predict friendship?

The problem is that friendships aren't independent observations. Some students are just more social than others (they nominate lots of friends regardless of who). Some students are more popular (they receive lots of nominations). And friendships tend to be reciprocated: if Alice names Bob, Bob is more likely to name Alice. A standard regression ignores all of this, and your standard errors will be wrong.

The **Additive and Multiplicative Effects (AME)** model handles these dependencies directly. It gives each actor a sender effect ($a_i$, how social they are), a receiver effect ($b_j$, how popular they are), and a position in a latent space ($u_i$, $v_j$) that captures who tends to connect with whom beyond what the covariates explain. The name reflects the two kinds of actor-level terms: the sender and receiver effects enter the model *additively* ($a_i + b_j$), while the latent positions enter *multiplicatively* ($u_i'v_j$, a dot product). Think of it as a regression that takes network structure seriously.

## The Data

We'll analyze a friendship network from the [Add Health study](https://addhealth.cpc.unc.edu/), a longitudinal study of adolescents in the United States. The `addhealthc3` dataset in `lame` contains a directed friendship nomination network along with student characteristics.

```{r load-data}
library(lame)
library(ggplot2)
set.seed(6886)

# load the Add Health friendship network
data(addhealthc3)

# convert valued network to binary (any nomination = friendship)
Y <- (addhealthc3$Y > 0) * 1
X_nodes <- addhealthc3$X

n <- nrow(Y)
cat("Students:", n, "\n")
cat("Friendships:", sum(Y, na.rm = TRUE), "\n")
cat("Network density:", round(mean(Y, na.rm = TRUE), 3), "\n")
```

Notice the `na.rm = TRUE` calls: network data often has missing entries (the diagonal is `NA` because self-ties are undefined, and some dyads may be unobserved). The model handles missing values internally via data augmentation, so `NA`s can stay in the matrix. The same applies to missing *covariates*: a few `NA`s in the node attributes (here `race` and `grade` have a handful) propagate to the dyads built from them, and `ame()` treats those dyads as unobserved and reports how many it data-augmented.

Before modeling, let's look at the basic structure. How much do students vary in their number of friends?

```{r viz-network, fig.width=10, fig.height=4, fig.alt="Two faceted histograms showing the distribution of nominations sent and nominations received per student in the Add Health friendship network, illustrating heterogeneity in sociality and popularity."}
out_degree <- rowSums(Y, na.rm = TRUE)
in_degree <- colSums(Y, na.rm = TRUE)

degree_df <- data.frame(
	Degree = c(out_degree, in_degree),
	Type = rep(c("Nominations sent", "Nominations received"), each = n)
)

ggplot(degree_df, aes(x = Degree)) +
	geom_histogram(binwidth = 1) +
	facet_wrap(~Type, ncol = 2) +
	labs(x = "Number of Ties", y = "Count") +
	theme_bw() +
	theme(
		panel.border = element_blank(),
		strip.background = element_rect(fill = "black", color = "black"),
		strip.text = element_text(color = "white", hjust = 0)
	)
```

There's real variation here: some students are much more social than others, and some are much more popular. This is exactly what the sender and receiver random effects will capture.

## Building Covariates

The classic question in friendship networks is **homophily**: do birds of a feather flock together? We'll test whether students are more likely to be friends with others of the same gender, race, and grade.

We deliberately include both `same_grade` (binary: same grade or not) and `grade_diff` (continuous: how many grades apart) to illustrate a common modeling pitfall: collinearity. These two covariates measure nearly the same thing, and the model cannot cleanly separate their effects. We'll see the consequences in the results below.

```{r create-covariates}
# `outer(x, x, FUN)` builds an n x n matrix whose (i,j) entry is FUN(x[i], x[j]).
# So `outer(female, female, "==")` returns TRUE where students i and j share the
# same gender. Multiplying by 1 converts TRUE/FALSE to 1/0.
#
# homophily indicators: 1 if same, 0 if different
same_female <- outer(X_nodes[,"female"], X_nodes[,"female"], "==") * 1
same_race <- outer(X_nodes[,"race"], X_nodes[,"race"], "==") * 1
same_grade <- outer(X_nodes[,"grade"], X_nodes[,"grade"], "==") * 1
# absolute grade difference; same shape, but a continuous covariate
grade_diff <- abs(outer(X_nodes[,"grade"], X_nodes[,"grade"], "-"))

# pack into a 3D array (n x n x p)
Xdyad <- array(NA, dim = c(n, n, 4))
Xdyad[,,1] <- same_female
Xdyad[,,2] <- same_race
Xdyad[,,3] <- same_grade
Xdyad[,,4] <- grade_diff
dimnames(Xdyad)[[3]] <- c('same_female', 'same_race', 'same_grade', 'grade_diff')
for(k in 1:4) diag(Xdyad[,,k]) <- NA

# nodal covariates (sender and receiver characteristics)
Xrow <- X_nodes[, c("female", "grade")]
Xcol <- X_nodes[, c("female", "grade")]
```

## Fitting the Model

Now the fun part. We fit a binary probit AME model with:

- **Covariates** for homophily effects (do similar students become friends?)
- **Sender and receiver random effects** (some students are more social/popular)
- **Dyadic correlation** (reciprocity: friendships tend to be mutual)
- **2-dimensional latent space** (residual clustering beyond what covariates explain)

### A note for ERGM users: what AME assumes

If you're coming from `ergm` / statnet, the most important conceptual difference is the dependence assumption. ERGM specifies a joint distribution over the entire adjacency matrix and uses change statistics (e.g. `gwesp`, `triangle`, `kstar`) to encode *unconditional* higher-order dependence: changing one tie shifts the probability of other ties through those terms directly. AME does not. AME assumes **conditional dyadic independence**: given the additive effects $(a_i, b_j)$, the latent positions $(u_i, v_j)$, the dyadic correlation $\rho$ for $(y_{ij}, y_{ji})$, and the covariates, all dyads are independent. Higher-order structure (clustering, transitivity, degree heterogeneity) is captured *indirectly* by integrating over these latent random effects; actors with similar $u_i$ tend to form ties to overlapping neighborhoods, which induces clustering and partial transitivity at the marginal level.

The practical consequences for an ERGM user:

- **No degeneracy.** Because the likelihood factors over dyads conditional on the random effects, AME does not exhibit the model-degeneracy failure modes of ERGMs with `triangle` or unconstrained `kstar` terms.
- **No change-statistic interpretation of $\beta$.** A coefficient on `same_grade` is the partial association on the probit-latent scale, not a log-odds change conditional on the rest of the network.
- **Some triangle structure is missed.** When transitive closure pressure is the substantive target (`gwesp` is a leading example), AME typically under-predicts triangles in the GOF and ERGM is the right tool. If your target is structural pattern + actor heterogeneity + covariate effects with calibrated uncertainty, AME is generally a more stable estimator.

```{r fit-model, message=FALSE, warning=FALSE}
fit <- ame(Y,
					Xdyad = Xdyad,
					Xrow = Xrow,
					Xcol = Xcol,
					R = 2,              # 2D latent space
					family = "binary",  # probit model for 0/1 data
					rvar = TRUE,        # sender random effects
					cvar = TRUE,        # receiver random effects
					dcor = TRUE,        # dyadic correlation (reciprocity)
					burn = 500,         # burn-in (lengthen for a final run)
					nscan = 2000,       # post-burn-in iterations
					odens = 25,         # thinning
					verbose = FALSE,
					gof = TRUE)
```

## What Did We Find?

```{r print-summary}
summary(fit)
```

Let's unpack the key results:

**Homophily effects.** Look at the regression coefficients and their credible intervals. The `grade_diff` coefficient is strongly negative and its credible interval excludes zero, telling us that students further apart in grade are much less likely to be friends. The `same_female` coefficient is positive (gender homophily) but its 95% credible interval includes zero, so it is not well-identified at this sample size -- read the sign as suggestive, not the effect as established. The `same_race` coefficient captures race homophily after controlling for grade and gender.

**The collinearity pitfall.** `same_grade` and `grade_diff` measure nearly the same thing -- a grade difference of 0 *is* the same-grade event -- so the model cannot cleanly *split* the grade signal between them. It still identifies the dominant effect: `grade_diff` is strongly negative and stable from run to run (about -0.59, never near zero). What suffers is the partition: `same_grade` is left to absorb a small, noisy residual bump on top of the linear grade-distance trend. The lesson is about *interpretation*, not estimate instability: include one or the other, not both, so the grade effect lands on a single, clean coefficient.

**Variance components.** The sender variance (`va`) and receiver variance (`vb`) quantify how much students differ in their sociality and popularity. Larger values mean more heterogeneity. The dyadic correlation (`rho`) captures reciprocity: values near 1 mean that if A nominates B, B almost always nominates A back. In this network, reciprocity is strong.

For binary networks with a probit link, the coefficients are on the latent scale. The marginal effect of a covariate on the response probability is $\phi(x'\beta) \cdot \beta_k$, where $\phi$ is the standard normal density. That density is maximized at $\phi(0) \approx 0.399$, which is the source of the textbook "$0.4 \times \beta$" rule of thumb for probit (see e.g. Wooldridge 2010, *Econometric Analysis of Cross Section and Panel Data*, ch. 15, or Greene 2018, *Econometric Analysis*, ch. 17). This approximation is only sharp when the baseline predicted probability is near 0.5. The Add Health friendship network has density `r round(mean(Y, na.rm = TRUE), 3)`, well below 0.5, so $\phi(\Phi^{-1}(0.128)) \approx 0.21$ -- the marginal effect of a unit change in a covariate is closer to $0.21 \times \beta$ than to the $0.4 \times \beta$ midpoint rule (and it would fall further, to about $0.18 \times \beta$, on a sparser $p \approx 0.1$ network). Two further wrinkles specific to AME: (i) the random effects $a_i, b_j, u_i'v_j$ contribute extra variance to the latent linear predictor, so the marginal effect after integrating them out is smaller again by a factor of $1/\sqrt{1 + V_\text{RE}}$; (ii) the right number to report depends on whether you want the effect at the sample mean, the average effect across observed dyads, or the effect at specific covariate values. For interpretation, compute the quantity you want with `predict(fit, type = "response")` and contrast predicted probabilities under counterfactual covariate settings.

## Did the Sampler Converge?

Since the model is estimated via MCMC (Markov chain Monte Carlo), trace plots are the first check that the sampler explored the posterior distribution thoroughly.

```{r trace-plots, fig.width=10, fig.height=8, fig.alt="MCMC trace and density plots for each regression coefficient; stable traces fluctuate around a steady mean and the density plots are smooth and unimodal."}
trace_plot(fit, params = "beta", ncol = 3)
```

**What to look for:** The trace plots (top) should fluctuate around a stable mean. If you see long trends, the chain has not converged. The density plots (bottom) should be smooth and unimodal. With `r nrow(fit$BETA)` post-burn-in samples most coefficients mix well; the slow ones at this seed are `grade_diff`, whose effective sample size drops to single digits, and -- typically the worse of the two across runs -- `rho`, the reciprocity parameter that is chronically the slowest-mixing term in these fits. Read this as a *mixing* problem, not an *estimate* problem: `grade_diff`'s point estimate is stable from run to run even when its ESS is low, and its near-twin `same_grade` mixes fine, so the poor mixing here comes from this short run and seed rather than a collinearity breakdown. Both `grade_diff` and `rho` call for a longer or multi-chain run before quoting their uncertainty.

### Numerical convergence diagnostics: `posterior::as_draws()`

A trace plot is a sanity check, not a diagnostic. The Stan-era summary triple is **split-$\hat R$**, **bulk ESS**, and **tail ESS**, and the `posterior` package computes all three from a `draws_array`. `lame` registers an `as_draws()` method that reshapes `BETA` and `VC` for any fit, so the same workflow you would run on a `stanfit` works here:

```{r as-draws-rhat-ess, eval = requireNamespace("posterior", quietly = TRUE)}
library(posterior)
draws <- posterior::as_draws(fit)              # draws_array [iter, chain, var]
posterior::summarise_draws(draws)              # mean, sd, q5, q95, rhat,
                                               # ess_bulk, ess_tail per param
```

The conventional thresholds: split-$\hat R$ < 1.01 for every monitored parameter; `ess_bulk` and `ess_tail` $\ge$ 400 per chain (so $\ge$ 1600 with four chains). This demo stores only `r nrow(fit$BETA)` draws (one short chain), so it cannot reach the ESS target -- that needs the longer, multi-chain run below -- but the $\hat R$ column is still informative. Most coefficients clear the $\hat R$ < 1.01 bar; two do not: `grade_diff_dyad` ($\hat R \approx 1.12$, `ess_bulk` in the single digits at this seed -- its point estimate is stable regardless), and `rho` ($\hat R \approx 1.08$), the reciprocity parameter that is the chronically slow-mixing term across seeds (on a typical run it, not `grade_diff`, is the worst-mixing parameter -- this fit's seed simply pushes `grade_diff` to the very bottom). A single-chain fit like this one reports within-chain split-$\hat R$, which is informative about within-chain non-stationarity but not about between-chain disagreement; the four-chain fit below gives the cross-chain check.

### Between-chain $\hat R$: `ame_parallel(n_chains = 4)`

For real convergence assessment, run four chains from different seeds and let `summarise_draws()` compute Rhat across chains:

```{r ame-parallel-rhat, eval = FALSE}
fit_mc <- ame_parallel(
	Y, Xdyad = Xdyad, Xrow = Xrow, Xcol = Xcol,
	family = "binary", R = 2,
	burn = 500, nscan = 2000, odens = 25,
	n_chains = 4, cores = 1,            # cores = 1 keeps the chunk portable
	combine_method = "pool",            # pooled fit; chain_indicator is preserved
	verbose = FALSE
)
posterior::summarise_draws(posterior::as_draws(fit_mc))
```

The pooled object carries a `chain_indicator` so `as_draws()` keeps chain identity separate, which is what split-$\hat R$ needs. The same reshape feeds `bayesplot::mcmc_trace()` and `tidybayes::tidy_draws()` directly.

### Sampler-failure diagnostics (the Gibbs analog of HMC divergences)

`lame` is a Gibbs / Metropolis-Hastings sampler, not HMC, so it has no notion of an HMC divergence. The closest analog is the per-block Metropolis acceptance / `tryError` count, exposed on every fit as `fit$mh_counters` (alias `fit$tryErrorChecks`). For the cross-sectional `ame()` fit here these counts are yours to inspect (they should be all zero); the longitudinal `lame()` path additionally fires an automatic warning when any block exceeds a 5% failure rate. For routine reporting:

```{r mh-counters}
str(fit$mh_counters)
```

Treat any block with >5% failures as suspicious; pair `ess_bulk` / `ess_tail` from `summarise_draws()` with `mh_counters` for the Gibbs-side equivalent of the Stan triple ($\hat R$, ESS, divergences).

### Inspecting the priors actually in effect: `prior_summary()`

Modelled on `rstanarm::prior_summary()`, `prior_summary(fit)` prints the hyperparameters that were *actually used* (defaults filled in) for the g-prior on $\beta$ and the variance components. Always run it once per fit; it is the cheapest way to catch a typo in a `prior = list(...)` override.

```{r prior-summary-cross}
prior_summary(fit)
```

### Model comparison: `loo()` and `loo_compare()`

If you refit with `save_log_lik = TRUE`, the fit carries a `[n_stored, n_obs]` pointwise log-likelihood matrix and `loo::loo(fit)` works directly via the registered S3 method:

```{r loo-cross, eval = FALSE}
fit_ll <- ame(Y, Xdyad = Xdyad, Xrow = Xrow, Xcol = Xcol,
              family = "binary", R = 2,
              burn = 500, nscan = 2000, odens = 25,
              save_log_lik = TRUE, verbose = FALSE)

# alternative model: drop the dyadic covariate
fit_ll_null <- ame(Y, Xrow = Xrow, Xcol = Xcol,
                   family = "binary", R = 2,
                   burn = 500, nscan = 2000, odens = 25,
                   save_log_lik = TRUE, verbose = FALSE)

loo::loo_compare(loo::loo(fit_ll), loo::loo(fit_ll_null))
```

For the six families `normal`, `binary`, `cbin`, `tobit`, `poisson`, `ordinal` the stored log-likelihood is the **exact family-specific Y density** (Gaussian / probit Bernoulli / censored-Gaussian / log-Poisson / cumulative-probit), so `elpd_loo` is directly comparable to a `loo()` from a brms or rstanarm fit to the same family. For the two rank likelihoods `frn` and `rrl`, the exact marginal requires GHK Monte Carlo and is only available on the longitudinal path via `log_lik_method = "observed_ghk"`; the cross-sectional default falls back to the augmented-Z normal density and emits a one-time warning. Inspect `fit$log_lik_method` to see which branch was used.

## Does the Model Fit the Data?

Goodness-of-fit (GOF) checks whether the model can reproduce key structural features of the observed network. This goes beyond dyad-level prediction: it tests whether the model captures emergent properties like degree heterogeneity, reciprocity patterns, and clustering.

```{r gof-plot, fig.width=10, fig.height=6, fig.alt="Goodness-of-fit panels comparing observed network statistics (dashed orange vertical line) against the posterior predictive histograms (grey) for sender and receiver degree heterogeneity, dyadic dependence, triadic dependence, and transitivity. Observed and predicted differ on both colour and linetype so the cue survives grayscale printing and colour-blind viewing."}
gof_plot(fit)
```

Each panel shows a different network statistic. The grey histogram is the posterior-predictive distribution from networks simulated from the fitted model; the **dashed orange vertical line is the observed value** on the same scale. The figure encodes the observed-vs-predicted contrast on two channels at once: colour (Okabe-Ito orange `#D55E00` vs `grey60`) **and** linetype (a dashed vertical segment vs a filled histogram bar), so the cue survives both colour-blind viewers and grayscale printing. The legend strip at the top of each panel labels the two glyphs explicitly as "Observed" and "Posterior predictive".

The panel titles are the human-readable labels ("Sender Degree Heterogeneity", "Receiver Degree Heterogeneity", "Dyadic Dependence", "Triadic Dependence", and "Transitivity"). The user-facing aliases accepted by `gof_plot(..., statistics = ...)` are `sd.row`, `sd.col`, `dyad.dep`, `triad.dep` (the cyclic-triad measure), and `trans.dep` (the transitive-triad analogue). The actual column names stored in `fit$GOF` differ slightly: the row/column SDs are `sd.rowmean` / `sd.colmean` and the cyclic-triad statistic is `cycle.dep`, so `colnames(fit$GOF)` returns `sd.rowmean`, `sd.colmean`, `dyad.dep`, `cycle.dep`, `trans.dep`. The alias map applied inside `gof_plot()` rewrites `sd.row` → `sd.rowmean`, `sd.col` → `sd.colmean`, and `triad.dep` → `cycle.dep` before plotting.

In the plot above the observed values for Sender and Receiver Degree Heterogeneity fall inside their posterior-predictive histograms (although Receiver sits to the right of the mode), and Dyadic Dependence (the Y-vs-t(Y) correlation, essentially empirical reciprocity) is also covered. Those three features are explicitly modeled by the sender/receiver random effects, the row/column variance components, and the dyadic correlation $\rho$. The two triad-level statistics tell a different story: observed Triadic Dependence sits in the upper tail of its simulated distribution, and observed Transitivity sits at the far right edge. AME under-predicts triangle closure on this network, the expected signature of conditional dyadic independence. The latent space soaks up some triangle structure through clustering of $u_i$ vectors, but the absence of an explicit triangle-closure term means we cannot match observed transitivity. If transitivity is your substantive target, fit an ERGM; if it is a diagnostic concern only, increasing `R` sometimes helps modestly.

You can also compute GOF after the fact using the `gof()` function, which is useful if you want to test custom statistics. The built-in `trans.dep` panel is a *correlation-style* clustering measure; ERGM users will typically want raw counts that match the `triangle` / `ttriple` / `gwesp` family. Those are one-liners as custom stats:

```{r custom-gof}
# ERGM-style custom statistics:
#   triangles  : count of transitive triples i -> j, j -> k, i -> k,
#                the directed analogue of ergm's `ttriple` change statistic
#                (and what `gwesp(decay = 0, fixed = TRUE)` reduces to)
#   two_path   : count of two-paths i -> j -> k (the "potential" triangles)
#   trans_ratio: triangles / two_path -- the classic clustering coefficient
#                (sna::gtrans), the right scalar summary of what gwesp targets
#   deg_cor    : Pearson correlation of out-degree and in-degree
custom_stats <- function(Y) {
	Yb <- Y; Yb[is.na(Yb)] <- 0           # NA -> 0 for matrix multiply
	YY <- Yb %*% Yb
	triangles   <- sum(YY * Yb)            # transitive triples
	two_path    <- sum(YY) - sum(diag(YY)) # two-paths excluding length-2 reciprocal loops
	out_deg <- rowSums(Y, na.rm = TRUE)
	in_deg  <- colSums(Y, na.rm = TRUE)
	c(triangles   = triangles,
	  two_path    = two_path,
	  trans_ratio = triangles / max(two_path, 1),
	  deg_cor     = cor(out_deg, in_deg))
}

gof_custom <- gof(fit, custom_gof = custom_stats, nsim = 50, verbose = FALSE)
obs <- gof_custom[1, ]
sim <- gof_custom[-1, , drop = FALSE]

# posterior-predictive p-value: fraction of simulated networks at least as
# extreme (right tail) as the observed; values near 0 mean the model
# under-predicts the statistic, values near 1 mean it over-predicts
pp_right <- function(o, s) mean(s >= o, na.rm = TRUE)
data.frame(
	stat       = colnames(gof_custom),
	observed   = round(obs, 3),
	sim_mean   = round(colMeans(sim, na.rm = TRUE), 3),
	pp_p_right = round(mapply(pp_right, obs, as.data.frame(sim)), 3)
)[c("triangles", "two_path", "trans_ratio", "deg_cor"), ]
```

Read the table the same way you would read `ergm::gof()`. For each statistic a `pp_p_right` near 0 means the simulated networks rarely hit the observed value from below, i.e. the model under-predicts that quantity; a value near 1 means it over-predicts. On Add Health the raw `triangles` count comes back reasonably well covered (`pp_p_right` around 0.25 in our run): the AME fit absorbs much of the triangle pressure through clustering of $u_i$ and through the joint scale of the additive effects, so the raw count is in the predictive distribution even though the model has no `gwesp`-style change statistic in the likelihood. `two_path` is driven by the marginal degree distribution and the additive effects, so it is also usually well covered (`pp_p_right` around 0.4 here). The structural gap shows up in the ratio: the `trans_ratio` (triangles / two-paths) is the unweighted clustering coefficient ERGM users recognise from `sna::gtrans`, and on this network it comes back with `pp_p_right` close to zero (around 0.05 in our run). That is the scalar summary targeted by `gwesp(decay = 0, fixed = TRUE)`: the model gets the absolute scale of triangle counts roughly right but does not concentrate triangles per two-path the way the data do. The `deg_cor` is degree assortativity: AME captures some of it through the sender-receiver covariance (`cab`), but the rest reflects higher-order structure outside the dyadic likelihood.

## Visualizing the Latent Space

The multiplicative effects ($u_i'v_j$) place each student in a 2D latent space. Students near each other in sender space (triangles) tend to nominate similar friends; students near each other in receiver space (circles) tend to be nominated by similar students. This captures clustering that the covariates alone can't explain.

```{r latent-space, fig.width=8, fig.height=6, fig.alt="Circular latent-space layout of students with triangles marking sender positions and circles marking receiver positions; proximity indicates similar friendship patterns."}
uv_plot(fit, layout = "circle", show.edges = FALSE,
				label.nodes = TRUE, label.size = 2.5) +
	theme_bw() +
	theme(
		panel.border = element_blank(),
		axis.text = element_blank(),
		axis.title = element_blank(),
		panel.grid = element_blank()
	)
```

Students on opposite sides of the plot have dissimilar friendship patterns. Clusters of students placed near each other likely share some unobserved characteristic (same social group, same extracurricular activities) that drives their friendship choices beyond gender, race, and grade.

## Who Are the Most Social / Popular Students?

The additive effects decompose individual heterogeneity into sender effects (sociality) and receiver effects (popularity):

```{r individual-effects, fig.width=10, fig.height=8, fig.alt="Two stacked lollipop charts showing the posterior-mean sender effect (sociality) and receiver effect (popularity) for each student as a point connected by a stem to a dashed zero line, sorted from the most negative effect on the left to the most positive on the right; positive values sit above the line and negative below."}
p1 <- ab_plot(fit, effect = "sender", sorted = TRUE,
							title = "Sender Effects (Sociality)")
p2 <- ab_plot(fit, effect = "receiver", sorted = TRUE,
							title = "Receiver Effects (Popularity)")

library(patchwork)
p1 / p2
```

Students with large positive sender effects nominate more friends than expected given their covariates; those with large positive receiver effects receive more nominations.

## Making Predictions

The model gives you predicted probabilities for every possible tie, including unobserved ones. This is useful for link prediction (who is likely to become friends?) and for understanding the model's fit at the dyad level. One detail: for sparse networks (density below 0.1), raw accuracy can be misleadingly high because predicting "no tie" for every dyad already gets most predictions right. Focus on the off-diagonal cells of the confusion matrix to see how well the model distinguishes true ties from non-ties.

```{r predictions}
pred_resp <- predict(fit, type = "response")

# how well does the model classify?
Y_vec <- as.vector(Y)
pred_vec <- as.vector(pred_resp)
keep <- !is.na(Y_vec)

# simple classification at threshold 0.5
pred_binary <- (pred_vec[keep] > 0.5) * 1
confusion <- table(Actual = Y_vec[keep], Predicted = pred_binary)
knitr::kable(confusion, caption = "Confusion Matrix (threshold = 0.5)")

accuracy <- sum(diag(confusion)) / sum(confusion)
baseline <- max(mean(Y_vec[keep]), 1 - mean(Y_vec[keep]))  # always-predict-modal-class
cat("\nAccuracy:", round(accuracy, 3),
    " | Baseline (predict modal class):", round(baseline, 3),
    " | Lift:", round(accuracy - baseline, 3), "\n")
```

The accuracy by itself flatters the model: a "predict the modal class" baseline already gets the printed baseline value, and the model's lift over that baseline is what's meaningful. For this friendship network with density ~13% the baseline is ~0.87, so the relevant question is whether the model's accuracy is materially above 0.87, not whether it crosses 0.5.

For link prediction, the area under the ROC curve (AUC) evaluates the model's ability to rank true ties above non-ties across all thresholds rather than relying on a single cutpoint. The `pROC` and `precrec` packages compute both ROC and PR-AUC from `pred_resp`. For sparse networks the PR-AUC is often the more useful summary. A model that always predicts "no tie" already has AUROC near 0.5 but PR-AUC near the network density, so the gap between PR-AUC and density is the metric of substantive interest.

```{r auc-eval, eval = FALSE}
# pROC: ROC curve and AUROC
# install.packages("pROC")
roc_obj <- pROC::roc(response = Y_vec[keep], predictor = pred_vec[keep],
                     quiet = TRUE)
auroc   <- as.numeric(pROC::auc(roc_obj))

# precrec: both ROC and Precision-Recall AUCs (the latter is more
# informative when the positive class is rare, as in friendship networks)
# install.packages("precrec")
ev <- precrec::evalmod(scores = pred_vec[keep], labels = Y_vec[keep])
precrec::auc(ev)   # returns AUROC and PR-AUC side-by-side

cat("AUROC:", round(auroc, 3), "\n",
    "Baseline AUROC (random ranker):", 0.5, "\n",
    "Baseline PR-AUC (density of Y):", round(mean(Y_vec[keep]), 3), "\n")
```

### Held-out link prediction

Reporting accuracy / AUC on the same dyads the model was fit on is **in-sample** fit, not predictive performance. For a held-out evaluation, mask a random subset of dyads to `NA` before fitting (the AME sampler handles `NA` entries via data augmentation, so the masked dyads are excluded from the likelihood), then score `predict(fit, type = "response")` on the held-out indices. Stratify the split on `Y` so the test set has both classes; on a density-0.1 network a naive 10% mask leaves very few positive test dyads, so report the test-set positive count with the AUC.

```{r heldout, eval = FALSE}
# 80/20 stratified mask of off-diagonal dyads
set.seed(6886)
off_diag    <- which(row(Y) != col(Y) & !is.na(Y))
pos_idx     <- off_diag[Y[off_diag] == 1]
neg_idx     <- off_diag[Y[off_diag] == 0]
test_idx    <- c(sample(pos_idx, round(0.2 * length(pos_idx))),
                 sample(neg_idx, round(0.2 * length(neg_idx))))

Y_train               <- Y
Y_train[test_idx]     <- NA      # mask test dyads from the likelihood

fit_train <- ame(Y_train, Xdyad = Xdyad, Xrow = Xrow, Xcol = Xcol,
                 R = 2, family = "binary",
                 rvar = TRUE, cvar = TRUE, dcor = TRUE,
                 burn = 500, nscan = 2000, odens = 25,
                 verbose = FALSE, gof = FALSE)

pred_train <- predict(fit_train, type = "response")

# `evaluate_heldout()` is the shipped helper that vectorises the same scoring
# call, gates AUROC / PR-AUC on whether {precrec} (preferred) or {pROC} is
# installed, and falls back to Brier + log-loss otherwise. It accepts the
# observed Y, predictions, and a logical TEST mask of the same shape.
test_mask           <- matrix(FALSE, nrow(Y), ncol(Y),
                              dimnames = dimnames(Y))
test_mask[test_idx] <- TRUE
evaluate_heldout(y_obs = Y, y_pred = pred_train,
                 mask = test_mask, family = "binary")
# columns: n_eval, auroc, auprc, brier, logloss
cat("Test dyads:", length(test_idx),
    " | positives:", sum(Y[test_idx] == 1), "\n")
```

Two details on the ML split. (1) The actor identities are the same in train and test (only some of their dyads are masked); this is a *transductive* link-prediction split, not an inductive one. The latent positions $u_i, v_j$ for every actor are still informed by their non-masked dyads. For an **inductive** evaluation that holds out entire actors, drop a subset of rows/columns from $Y$ before fitting and score the model's predictions for those actors using only covariates (the held-out actors have no posterior $u_i, v_j$, so their predictions collapse to $X\beta + \hat a_{\text{mean}} + \hat b_{\text{mean}}$, which unsurprisingly performs worse). (2) The fit above is a single chain at modest length; for stable held-out AUC estimates run `ame_parallel(..., n_chains = 4)` and pool. A standard `lame` fit on a network with $n \lesssim 200$ takes seconds to a few minutes at `nscan = 2000`; for $n \approx 500$ expect tens of minutes, and beyond $n \approx 1000$ the per-iteration cost of the $O(n^2)$ dyadic operations starts to dominate and the fast point estimator (`ame_als()`, see the [fast estimation vignette](fast_estimation.html)) is the recommended exploration tool. There is no GPU backend in the current release.

## Simulating from the Model

You can generate new networks from the fitted posterior. A **posterior
predictive check** asks: if I draw a parameter set from the posterior
and then simulate a network with those parameters, does that simulated
network look like the observed one on whatever feature I care about
(here, overall density)? If the observed feature lies inside the
simulated distribution, the model reproduces that aspect of the data;
if it lies in a tail, the model is missing something there.

```{r simulation, fig.width=8, fig.height=4, fig.alt="Posterior predictive check histogram of simulated network densities (grey bars) with a dashed Okabe-Ito orange vertical line marking the observed density; the encoding is dual on colour and linetype so the cue survives greyscale printing and colour-blind viewing. Overlap indicates the model reproduces the observed density."}
sims <- simulate(fit, nsim = 100)

# compare simulated vs observed density
sim_densities <- sapply(sims$Y, function(y) mean(y, na.rm = TRUE))
obs_density <- mean(Y, na.rm = TRUE)

ggplot(data.frame(density = sim_densities), aes(x = density)) +
	geom_histogram(bins = 20, fill = "grey60") +
	# dual-encode the observed density as colour AND linetype so the cue
	# survives greyscale printing and colour-blind viewing (Okabe-Ito orange).
	geom_vline(xintercept = obs_density,
	           color = "#D55E00", linewidth = 1, linetype = "dashed") +
	labs(title = "Posterior Predictive Check: Network Density",
			subtitle = "Dashed orange line = observed; grey histogram = simulated from model",
			x = "Network Density (Mean Tie Probability)", y = "Replicate Count") +
	theme_bw() +
	theme(panel.border = element_blank())
```

## Practical Tips

**Choosing R (latent dimensions).** Start with R = 0 (no latent space), then try R = 1 and R = 2. Compare GOF statistics. If adding dimensions doesn't improve the GOF, the covariates and additive effects are sufficient. `ame()` / `lame()` will warn when `R > floor(n/3)` (or `min(nA, nB)/3` in bipartite) because at that rank the multiplicative effects start absorbing structure that belongs to the additive effects and the latent positions stop being interpretable; in practice `R = 2` or `R = 3` is the right default unless you have a specific reason to push higher.

**MCMC settings.** For exploratory work, `burn = 500, nscan = 2000, odens = 25` is fine. For a final run, use at least `burn = 2000, nscan = 10000, odens = 25` and check convergence with `trace_plot()`.

**Variance components.** Include `rvar = TRUE` and `cvar = TRUE` unless you have a specific reason not to. Include `dcor = TRUE` for directed networks where reciprocity is plausible.

**Interpreting coefficients.** For binary networks, the coefficients are on the probit scale. Use `predict(fit, type = "response")` to get predicted probabilities.

## What's Next?

- **Multiple time periods?** See the [lame overview](lame-overview.html) for longitudinal models
- **Two types of nodes?** See the [bipartite vignette](bipartite.html)
- **Evolving network structure?** See the [dynamic effects vignette](dynamic_effects.html)
- **Other data types?** The `family` argument supports normal, binary, ordinal, poisson, tobit, and more

## References

Hoff, PD (2021). Additive and Multiplicative Effects Network Models. *Statistical Science* 36, 34--50.

Minhas, S., Dorff, C., Gallop, M. B., Foster, M., Liu, H., Tellez, J., & Ward, M. D. (2022). Taking dyads seriously. *Political Science Research and Methods*, 10(4), 703--721.
