---
title: "Dynamic Effects in Longitudinal AME Models"
author: "Cassy Dorff, Tosin Salau, Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic Effects in Longitudinal AME Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Networks change. Countries that were close allies a decade ago may have drifted apart. A legislator's co-sponsorship patterns shift as they gain seniority or switch committee assignments. A user's purchasing habits evolve as their tastes change.

The standard AME model assumes that each actor has a fixed latent position and fixed additive effects across all time periods. This is a reasonable starting point (pooling across time gives you more data and more precise estimates), but it can miss important dynamics. The `lame` package provides two mechanisms for letting the model capture temporal change: `dynamic_uv` for time-varying latent positions and `dynamic_ab` for time-varying additive effects.

This vignette explains what these options do, when to use them, and how to interpret the results.

## What Are Dynamic Effects?

### The Static Baseline

In the standard AME model for longitudinal data, the tie between actors $i$ and $j$ at time $t$ is:

$$y_{ij,t} = \beta'x_{ij,t} + a_i + b_j + u_i'v_j + \epsilon_{ij,t}$$

The covariates ($x_{ij,t}$) can vary over time, but everything else (the sender effect $a_i$, the receiver effect $b_j$, and the latent positions $u_i$ and $v_j$) is constant. This means the model assumes that a country's tendency to sanction others, or its position in the latent "sanctioning space," is the same in 1993 as in 2000.

### Dynamic Latent Positions (`dynamic_uv = TRUE`)

When you set `dynamic_uv = TRUE`, each actor's latent position evolves over time according to an AR(1) process:

$$U_{i,k,t} = \rho_{uv} \, U_{i,k,t-1} + \epsilon_{i,k,t}$$
$$V_{j,k,t} = \rho_{uv} \, V_{j,k,t-1} + \eta_{j,k,t}$$

The autoregressive parameter $\rho_{uv}$ controls how persistent the positions are. A value close to 1 means positions change slowly, as last year's position is a strong predictor of this year's. A value close to 0 means positions are essentially re-drawn each period. In practice, $\rho_{uv}$ is estimated from the data, so you don't need to choose it yourself.

This is useful when you believe the underlying community structure is evolving: alliances shift, social groups re-form, trading blocs realign.

### Dynamic Additive Effects (`dynamic_ab = TRUE`)

When you set `dynamic_ab = TRUE`, the sender and receiver effects evolve over time:

$$a_{i,t} = \rho_{ab} \, a_{i,t-1} + \epsilon_{i,t}$$
$$b_{j,t} = \rho_{ab} \, b_{j,t-1} + \eta_{j,t}$$

This captures changes in actors' overall activity levels. A country might become a more active sanctioner after a change in government. A student might become more or less socially active across semesters. These are changes in how much an actor participates, not in who they connect with (that's what `dynamic_uv` captures).

You can use either option alone or combine them. In practice, `dynamic_ab` is often a good place to start, since changes in overall activity are common and relatively easy to estimate. `dynamic_uv` adds more flexibility but also more parameters.

## Prior Specifications

The AR(1) coefficients and innovation variances have sensible default priors:

- $\rho_{uv} \sim \text{TruncNormal}(0.9, 0.1, 0, 1)$, centered on high persistence
- $\rho_{ab} \sim \text{TruncNormal}(0.8, 0.15, 0, 1)$, slightly less persistent
- $\sigma_{uv}^2, \sigma_{ab}^2 \sim \text{InverseGamma}(2, 1)$

These can be customized via the `prior` argument if you have strong beliefs about the rate of change:

```{r}
prior_custom <- list(
  rho_uv_mean = 0.95,    # expect very slow change in latent positions
  rho_uv_sd = 0.05,      # tight prior
  sigma_uv_shape = 3,
  sigma_uv_scale = 2
)
```

## Practical Example

We simulate a simple longitudinal network and fit models with and without dynamic effects to see the difference.

```{r}
library(lame)

set.seed(6886)
n <- 25  # actors
n_periods <- 5

# Generate sparse binary networks
Y_list <- list()
for(t in 1:n_periods) {
  Y_t <- matrix(rbinom(n*n, 1, 0.1), n, n)
  diag(Y_t) <- NA
  rownames(Y_t) <- colnames(Y_t) <- paste0("Actor", 1:n)
  Y_list[[t]] <- Y_t
}

# Fit model with both dynamic effects
# Note: iterations are reduced for vignette speed.
# Use burn >= 1000 and nscan >= 5000 for real analyses.
fit_dynamic <- lame(
  Y = Y_list,
  R = 2,
  dynamic_uv = TRUE,
  dynamic_ab = TRUE,
  family = "binary",
  burn = 100,
  nscan = 500,
  odens = 25,
  print = FALSE,
  plot = FALSE
)
```

### Visualizing Dynamic Effects

The trajectory plot shows how each actor's latent position moves through the 2-dimensional space over time. Each line traces one actor's path from the first to the last period. Actors whose lines are short stayed in roughly the same position; actors with long, wandering lines experienced substantial shifts in their network role.

```{r, fig.width=7, fig.height=5}
uv_plot(fit_dynamic, plot_type = "trajectory")
```

The `ab_plot` with `plot_type = "trajectory"` shows how each actor's additive effect changes over time. Rising lines indicate actors who became more active (or more popular) over the observation window; falling lines indicate the opposite. This is particularly useful for identifying actors who experienced a structural break in their network behavior.

```{r, fig.width=7, fig.height=5}
ab_plot(fit_dynamic, effect = "sender", plot_type = "trajectory")
```

As always, check convergence. Dynamic models have more parameters, so adequate mixing is harder to achieve and requires longer chains.

```{r, fig.width=7, fig.height=5}
trace_plot(fit_dynamic)
```

### Comparing Model Specifications

A natural workflow is to fit several specifications and compare them:

```{r}
# Static model (no dynamic effects)
fit_static <- lame(Y_list, R = 2, family = "binary",
                   burn = 100, nscan = 400, odens = 20,
                   print = FALSE, plot = FALSE)

# Dynamic latent positions only
fit_uv <- lame(Y_list, R = 2, dynamic_uv = TRUE, family = "binary",
               burn = 100, nscan = 400, odens = 20,
               print = FALSE, plot = FALSE)

# Dynamic additive effects only
fit_ab <- lame(Y_list, R = 2, dynamic_ab = TRUE, family = "binary",
               burn = 100, nscan = 400, odens = 20,
               print = FALSE, plot = FALSE)

# Full dynamic
fit_full <- lame(Y_list, R = 2, dynamic_uv = TRUE, dynamic_ab = TRUE,
                 family = "binary",
                 burn = 100, nscan = 400, odens = 20,
                 print = FALSE, plot = FALSE)
```

Compare the GOF statistics across models. The model that best reproduces the observed network statistics (heterogeneity, transitivity, dyadic dependence) without unnecessary complexity is usually the best choice.

```{r}
if(!is.null(fit_static$GOF) && !is.null(fit_full$GOF)) {
  cat("Static model GOF (sample):\n")
  print(head(fit_static$GOF))
  cat("\nFull dynamic model GOF (sample):\n")
  print(head(fit_full$GOF))
}
```

### When to Use Dynamic Effects

**Use `dynamic_ab`** when you suspect actors' overall activity levels change over time. This is common in many settings: countries go through isolationist vs. interventionist phases, users churn in and out of platforms, students become more or less engaged across semesters.

**Use `dynamic_uv`** when you suspect the underlying community structure is shifting. This is a stronger claim, not just that actors are more or less active, but that the pattern of *who connects with whom* is changing. Examples include political realignment, market disruption, or generational turnover in a social network.

**Use both** when you believe both types of change are happening. This is the most flexible specification but also the most data-hungry. With short panels (few time periods) or sparse networks, the dynamic parameters may not be well-identified, and you might be better off with the static model.

**Stick with static** when you have few time periods, when the network structure is genuinely stable, or when you primarily care about the covariate effects (which are the same across specifications) rather than the latent structure.

## Implementation Notes

The dynamic effects are implemented in C++ via Rcpp and RcppArmadillo using a forward-filtering backward-sampling (FFBS) algorithm. This is the standard approach for state-space models: the forward pass computes predictive distributions $p(x_t | x_{1:t-1})$, and the backward pass samples from the smoothed distribution $p(x_{1:T} | y_{1:T})$.

## References

1. **Hoff, PD (2021)**. Additive and Multiplicative Effects Network Models. *Statistical Science* 36, 34--50.

2. **Sewell, D. K., & Chen, Y. (2015)**. Latent space models for dynamic networks. *Journal of the American Statistical Association*, 110(512), 1646-1657.

3. **Durante, D., & Dunson, D. B. (2014)**. Nonparametric Bayes dynamic modeling of relational data. *Biometrika*, 101(4), 883-898.
