---
title: "Dynamic Effects in Longitudinal AME Models"
author: "Cassy Dorff, Tosin Salau, Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic Effects in Longitudinal AME Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>",
	fig.width = 7,
	fig.height = 5
)
```

## Introduction

Networks change. Countries that were close allies a decade ago may have drifted apart. A legislator's co-sponsorship patterns shift as they gain seniority or switch committee assignments. A user's purchasing habits evolve as their tastes change.

The standard AME model assumes that each actor has a fixed latent position and fixed additive effects across all time periods. This is a reasonable starting point (pooling across time gives you more data and more precise estimates), but it can miss important dynamics. The `lame` package provides two mechanisms for letting the model capture temporal change: `dynamic_uv` for time-varying latent positions and `dynamic_ab` for time-varying additive effects.

This vignette explains what these options do, when to use them, and how to interpret the results.

## What Are Dynamic Effects?

### The Static Baseline

In the standard AME model for longitudinal data, the tie between actors $i$ and $j$ at time $t$ is:

$$y_{ij,t} = \beta'x_{ij,t} + a_i + b_j + u_i'v_j + \epsilon_{ij,t}$$

The covariates ($x_{ij,t}$) can vary over time, but everything else (the sender effect $a_i$, the receiver effect $b_j$, and the latent positions $u_i$ and $v_j$) is constant. This means the model assumes that a country's tendency to sanction others, or its position in the latent "sanctioning space," is the same in 1993 as in 2000.

### Dynamic Latent Positions (`dynamic_uv = TRUE`)

When you set `dynamic_uv = TRUE`, each actor's latent position evolves over time according to an AR(1) process:

$$U_{i,k,t} = \rho_{uv} \, U_{i,k,t-1} + \epsilon_{i,k,t}, \quad \epsilon_{i,k,t} \sim N(0, \sigma_{uv}^2)$$
$$V_{j,k,t} = \rho_{uv} \, V_{j,k,t-1} + \eta_{j,k,t}, \quad \eta_{j,k,t} \sim N(0, \sigma_{uv}^2)$$

Note the **innovation parameterisation** *and* the storage convention. The symbol $\sigma_{uv}^2$ is the variance of the period-to-period innovation $\epsilon_{i,k,t}$, not the marginal/stationary variance of $U_{i,k,t}$ — and the on-fit field `fit$sigma_uv` holds the per-iteration posterior draws of the innovation **standard deviation** $\sigma_{uv}$ (so `mean(fit$sigma_uv)` is the posterior-mean SD and `mean(fit$sigma_uv)^2` recovers the innovation variance). Under the stationary AR(1), the implied marginal variance is $\sigma_{uv}^2 / (1 - \rho_{uv}^2)$. This matters when simulation code supplies an SD to `rnorm()` in the recurrence: using the *marginal* SD makes the process too variable by a factor of $1/(1 - \rho_{uv}^2)$. The autoregressive parameter $\rho_{uv}$ controls how persistent the positions are. A value close to 1 means positions change slowly, as last year's position is a strong predictor of this year's. A value close to 0 means positions are essentially re-drawn each period. In practice, $\rho_{uv}$ is estimated from the data.

This is useful when you believe the underlying community structure is evolving: alliances shift, social groups re-form, trading blocs realign.

### Dynamic Additive Effects (`dynamic_ab = TRUE`)

When you set `dynamic_ab = TRUE`, the sender and receiver effects evolve over time:

$$a_{i,t} = \rho_{ab} \, a_{i,t-1} + \epsilon_{i,t}, \quad \epsilon_{i,t} \sim N(0, \sigma_{ab}^2)$$
$$b_{j,t} = \rho_{ab} \, b_{j,t-1} + \eta_{j,t}, \quad \eta_{j,t} \sim N(0, \sigma_{ab}^2)$$

The same innovation parameterisation applies as for `dynamic_uv`: $\sigma_{ab}^2$ is the *innovation* variance, and `fit$sigma_ab` holds the posterior draws of the innovation standard deviation $\sigma_{ab}$ (square the posterior mean to get the innovation variance); the implied stationary marginal variance is $\sigma_{ab}^2 / (1 - \rho_{ab}^2)$. The sampler in `sample_dynamic_ab_cpp` uses the stationary distribution $N(0, \sigma_{ab}^2 / (1 - \rho_{ab}^2))$ as the **prior on the initial state** $a_{i,1}$ — i.e. the $t = 0$ prior precision is $1 / (\sigma_{ab}^2 / (1 - \rho_{ab}^2))$ rather than the one-step innovation precision $1 / \sigma_{ab}^2$ — and then runs a single-pass Gibbs sweep across $t$ in which the interior full conditional combines both the AR(1) look-back from $a_{i,t-1}$ and the look-ahead from $a_{i,t+1}$ (a smoothing-style update), reducing to the look-back alone at $t = T - 1$. Substantively, `dynamic_ab` captures changes in actors' overall activity levels. A country might become a more active sanctioner after a change in government. A student might become more or less socially active across semesters. These are changes in how much an actor participates, not in who they connect with (that's what `dynamic_uv` captures).

You can use either option alone or combine them. In practice, `dynamic_ab` is often a good place to start, since changes in overall activity are common and relatively easy to estimate. `dynamic_uv` adds more flexibility but also more parameters.

### Dynamic Regression Coefficients (`dynamic_beta`)

The third dynamic option lets the regression coefficients themselves evolve over time. When you set `dynamic_beta = TRUE` (or pass a subset like `dynamic_beta = "dyad"` to make only the dyadic coefficients dynamic), each selected coefficient follows an independent AR(1):

$$\beta_{k,t} = \rho_{\beta,b(k)} \, \beta_{k,t-1} + \epsilon_{k,t}, \quad \epsilon_{k,t} \sim N(0, \sigma_{\beta,b(k)}^2 \Lambda_{kk})$$

where $b(k)$ is the block of coefficient $k$ (intercept / dyad / row / col), so all coefficients in the same block share the AR(1) persistence $\rho_{\beta,b(k)}$ and innovation scale $\sigma_{\beta,b(k)}$. The diagonal matrix $\Lambda$ is a per-coefficient scale built once from the (period-averaged) design `XtX` so that the innovation variance for coefficient $k$ is commensurate with the data-level information about $\beta_k$ (a $g$-prior analogue; see `build_beta_state_scales()` in the source); by construction $\Lambda$ is diagonal, so the innovations are independent within a block. The package draws the entire time path of $\beta$ jointly via forward-filter / backward-sample (FFBS; Carter & Kohn 1994, Frühwirth-Schnatter 1994), the efficient state-space sampler for conditionally-Gaussian dynamic linear models. The per-block quantity is an **innovation SD** $\sigma_{\beta,b}$ (not a variance); the conditional variance of the innovation for coefficient $k$ is $\sigma_{\beta,b(k)}^2 \Lambda_{kk}$. Note a storage difference from `dynamic_uv`/`dynamic_ab`: `fit$sigma_beta` and `fit$rho_beta` hold the **last-iteration** scalars, while the full posterior draws live in `fit$SIGMA_BETA` and `fit$RHO_BETA` -- summarise those (e.g. `colMeans(fit$SIGMA_BETA)`), do not square the last-iteration `fit$sigma_beta`.

This is useful when you believe a covariate's effect changes over time: maybe trade increases the probability of cooperation more strongly during a recession, or a peer's behavior matters more in certain semesters. The dynamic-coefficient path is reported in `coef(fit)` as a `p x T` matrix, and the per-period 95% CIs are available via `confint(fit)`.

You can pass `dynamic_beta` in five forms:

- `FALSE` (default): every coefficient is static.
- `TRUE`: every coefficient becomes dynamic.
- character vector of block shortcuts: `"intercept"`, `"dyad"`, `"row"`, `"col"`, or any coefficient name from `colnames(fit$BETA)`.
- integer vector: column indices into the BETA layout.
- logical vector of length `p`: per-coefficient mask.

When the intercept (or a nodal coefficient) is included in `dynamic_beta`, the additive-effects sampler imposes a sum-to-zero constraint on `a` (and `b`) so that the intercept and the additive effects remain jointly identified. This happens automatically.

The AR(1) hyperparameters $\rho_\beta$ and $\sigma_\beta^2$ have their own priors:

- $\rho_\beta \sim \text{TruncNormal}(0.8, 0.15, 0, 0.999)$
- $\sigma_\beta^2 \sim \text{InverseGamma}(2, 1)$

You can override these via `prior = list(rho_beta_mean = ..., rho_beta_sd = ..., sigma_beta_shape = ..., sigma_beta_scale = ...)`.

#### A worked `dynamic_beta` example

Simulate a small longitudinal network where a single dyadic covariate's effect grows linearly over time, and fit with `dynamic_beta = "dyad"`.

```{r dynamic-beta-example, message=FALSE}
library(lame)
set.seed(2026)

n_db <- 15
T_db <- 5
beta_t_true <- seq(-0.5, 1.0, length.out = T_db)   # truth: rising effect

X_db <- replicate(T_db, matrix(rnorm(n_db * n_db), n_db, n_db), simplify = FALSE)
Y_db <- vector("list", T_db)
for (t in seq_len(T_db)) {
	Yt <- beta_t_true[t] * X_db[[t]] + matrix(rnorm(n_db * n_db, 0, 0.4), n_db, n_db)
	diag(Yt) <- NA
	rownames(Yt) <- colnames(Yt) <- paste0("a", seq_len(n_db))
	Y_db[[t]] <- Yt
}
names(Y_db) <- paste0("t", seq_len(T_db))
X_db_arr <- lapply(X_db, function(x) array(x, c(n_db, n_db, 1)))

fit_db <- lame(Y_db, Xdyad = X_db_arr,
               family = "normal", R = 0,
               nscan = 200, burn = 50, odens = 5,
               dynamic_beta = "dyad", verbose = FALSE)

# fit$BETA is a 3-D array [n_stored x p x T]
dim(fit_db$BETA)

# coef() returns a p x T matrix of posterior means
coef(fit_db)

# confint() returns one row per coef[t] with 95% credible interval
head(confint(fit_db), 8)
```

**Migration warning for `amen` / static-`lame` users.** Under `amen::ame()` and any static `lame` fit, `fit$BETA` is a 2-D `[n_stored, p]` matrix and `apply(fit$BETA, 2, mean)` gives the posterior mean per coefficient. When `dynamic_beta` is active, `fit$BETA` becomes a 3-D `[n_stored, p, T]` array. The second margin is still the coefficient index in both shapes, so `apply(fit$BETA, 2, mean)` still returns a length-`p` vector and looks superficially correct, but it now silently averages across both iterations **and** time periods. The call does not error, the output length matches what the old code expected, and the silent collapse can survive into downstream plots as a single "average effect" that hides any time variation the user was trying to recover. Two safe patterns:

```{r dynamic-beta-migration, eval = FALSE}
# check the shape before summarising coefficients
if (length(dim(fit_db$BETA)) == 3L) {
	# 3-D: take posterior mean over the iteration margin -> [p, T]
	pm <- apply(fit_db$BETA, c(2, 3), mean)
} else {
	# 2-D amen-style fit
	pm <- apply(fit_db$BETA, 2, mean)
}

# or just use coef(), which returns a [p, T] matrix for dynamic fits
# and a length-p vector for static fits -- no shape sniffing required
coef(fit_db)
```

The per-period posterior mean for `X1_dyad` should track the linear ramp `beta_t_true = c(-0.5, -0.125, 0.25, 0.625, 1.0)`. The AR(1) prior provides mild shrinkage that reduces the per-period variance compared to fitting a separate `ame()` per period.

`summary(fit_db)` prints a *Dynamic coefficients per period* block with per-coefficient `Mean | Min | Max | Drift_pct | Dynamic` columns; the `Drift_pct` is `100 * (Max - Min) / |Mean|`, useful for spotting which coefficients move the most across periods.

#### Choosing a prior family: `dynamic_beta_kind`

`lame()` exposes a stable argument `dynamic_beta_kind` for selecting the prior family on the time-varying coefficients. **In this release the full decision tree is:**

> **Mean-reverting** coefficient? Use the default `dynamic_beta_kind = "ar1"`.
> **Permanent / unit-root drift** (no mean reversion)? Use `dynamic_beta_kind = "rw1"`.
> **Smooth with curvature** (acceleration matters; second-order smoothness)? Use `dynamic_beta_kind = "rw2"`.
> **Smooth with a known length-scale** of variation? Use `dynamic_beta_kind = "matern32"` (set the length scale via `prior = list(matern32_length_scale = ...)`; defaults to `max(2, T/4)`).
> If the stationarity warning fires after fitting AR(1), meaning the posterior on $\rho_\beta$ is concentrated at or above 0.97, refit with `"rw1"` (or `"rw2"` if acceleration matters).

`"random_walk"` is accepted as an alias for `"rw1"`. The `"rw2"` and `"matern32"` samplers use an R-level joint Gaussian update (slower than the C++ FFBS used for `"ar1"` / `"rw1"`, but produces full posterior draws and works on any T).

The same `fit_db` setup above can be refit under each kind. Each call swaps only the `dynamic_beta_kind` argument; `coef()` returns the same `p x T` matrix regardless.

```{r dynamic-beta-kind-rw1}
# rw1: random walk, rho pinned at 1
fit_rw1 <- lame(Y_db, Xdyad = X_db_arr, family = "normal", R = 0,
                nscan = 200, burn = 50, odens = 5,
                dynamic_beta = "dyad", dynamic_beta_kind = "rw1",
                verbose = FALSE)
round(coef(fit_rw1), 3)
```

```{r dynamic-beta-kind-rw2}
# rw2: smooth with curvature
fit_rw2 <- lame(Y_db, Xdyad = X_db_arr, family = "normal", R = 0,
                nscan = 200, burn = 50, odens = 5,
                dynamic_beta = "dyad", dynamic_beta_kind = "rw2",
                verbose = FALSE)
round(coef(fit_rw2), 3)
```

```{r dynamic-beta-kind-matern32}
# matern 3/2: smooth gp with a length-scale knob
fit_m32 <- lame(Y_db, Xdyad = X_db_arr, family = "normal", R = 0,
                nscan = 200, burn = 50, odens = 5,
                dynamic_beta = "dyad", dynamic_beta_kind = "matern32",
                prior = list(matern32_length_scale = 2),
                verbose = FALSE)
round(coef(fit_m32), 3)
```

When `time_index` is supplied with unequal gaps, the AR(1) / RW1 conditional variance is automatically scaled by the gap (the per-coefficient diagonal entry, suppressing $\Lambda_{kk}$ for readability):
$$q_t = \sigma_\beta^2 \cdot \frac{1 - \rho_\beta^{2\Delta_t}}{1 - \rho_\beta^2}$$
so quarterly-then-annual data, for example, gets the correct prior. The full diagonal entry is $\sigma_\beta^2 \Lambda_{kk} (1 - \rho_\beta^{2\Delta_t}) / (1 - \rho_\beta^2)$; the $\Lambda$ factor is the same period-invariant per-coefficient scale used in the equal-gap case. Equal-gap `time_index` is byte-identical to leaving it unspecified. For `dynamic_beta_kind = "rw2"` and `"matern32"`, `time_index` is consumed by the joint precision-matrix construction (RW2 uses the second-difference operator on the supplied positions; Matern 3/2 uses the continuous-time kernel $k(s, t) = (1 + \sqrt{3}|s-t|/\ell)\exp(-\sqrt{3}|s-t|/\ell)$), so unequal-gap support is uniform across kinds even though the closed-form $q_t$ above is the AR(1)/RW1 case.

```{r dynamic-beta-time-index}
# observations at t = 1, 2, 4, 8 (a doubling gap structure)
fit_gaps <- lame(Y_db, Xdyad = X_db_arr, family = "normal", R = 0,
                 nscan = 200, burn = 50, odens = 5,
                 dynamic_beta = "dyad",
                 time_index = c(1, 2, 4, 8, 16)[seq_len(T_db)],
                 verbose = FALSE)
fit_gaps$time_index
round(coef(fit_gaps), 3)
```

#### Hierarchical pooling: `dynamic_beta_pool`

When multiple coefficient blocks are dynamic (intercept / dyad / row / col), each block has its own AR(1) hyperparameters $(\rho_b, \sigma_b)$ by default, and short panels (small $T$) can leave one block's hyperparameters badly under-identified. `dynamic_beta_pool` adds a hierarchical layer that shrinks block-level hyperparameters toward a shared mean: use `"rho"` when you want blocks to share a common persistence (you believe e.g. dyadic and nodal coefficients all drift at similar speed but the data alone can't pin each $\rho_b$ down), `"sigma"` when blocks should share an innovation scale (a common case when blocks were standardised consistently), and `"both"` when you want full hierarchical pooling. The default `"none"` leaves each block independent. Pooling is implemented as a hierarchical MH step on the block-level hyperpriors and is most valuable for $T \le 5$ panels where the per-block posteriors otherwise collapse to their priors.

```{r dynamic-beta-pool}
fit_pool <- lame(Y_db, Xdyad = X_db_arr, family = "normal", R = 0,
                 nscan = 200, burn = 50, odens = 5,
                 dynamic_beta = "dyad", dynamic_beta_pool = "both",
                 verbose = FALSE)
fit_pool$dynamic_beta_pool
```

> **Diagnostic check.** With pooling on, compare the per-block posterior dispersion of `(rho_beta, sigma_beta)` against an identical fit with `dynamic_beta_pool = "none"`. For the pooling layer to be doing useful work the between-block sd-of-means should shrink under `"both"` relative to `"none"`. On very short panels with only a single dynamic block (`dynamic_beta = "dyad"` alone) the hierarchical layer has no other blocks to borrow strength from, and the pooled fit can actually look noisier in `RHO_BETA` because the hyperprior is then a near-flat mixing distribution over a single posterior draw; pooling earns its keep when **two or more** blocks (e.g. `dynamic_beta = c("intercept", "dyad", "row", "col")`) are dynamic at once.

#### Per-actor time-varying slopes: `dynamic_beta_per_actor`

For research questions that ask *whether one actor responds to a covariate differently than another, and whether that gap evolves*, set `dynamic_beta_per_actor = "row"` (or `"col"`) and choose the relevant covariate via `per_actor_covariate_idx`. The sampler imposes a per-period sum-to-zero constraint on the per-actor deviations, so the population coefficient remains in `coef(fit)` and the deviations are stored in `fit$THETA_ACTOR` (a `n_iter x n_actors x T` array under the default `keep_per_actor = "auto"`).

```{r dynamic-beta-per-actor, eval = FALSE}
fit_pa <- lame(Y_db, Xdyad = X_db_arr, family = "normal", R = 0,
               dynamic_beta_per_actor = "row",
               per_actor_covariate_idx = 1L,
               keep_per_actor = "summary",
               nscan = 200, burn = 50, odens = 5, verbose = FALSE)
str(fit_pa$theta_actor_mean)   # n_actors x T posterior-mean deviations

# check the per-period sum-to-zero constraint
colSums(fit_pa$theta_actor_mean)   # ~ 0 0 0 0 0
```

> **Known dimnames gap.** `theta_actor_mean` is currently returned without `rownames()`/`colnames()` even when the input `Y` has actor names and named periods. If you need labelled output, attach them manually after the fit using whatever object on the fit still carries the labels, e.g. `dimnames(fit_pa$theta_actor_mean) <- list(rownames(fit_pa$YPM[[1]]), names(fit_pa$YPM))`. The sum-to-zero identifiability constraint (`per_actor_identifiability = "center"`, the default) is enforced exactly: `colSums(fit_pa$theta_actor_mean)` is numerically zero at every period.

#### Diagnosing your dynamic fit

After fitting, two helpers summarize the coefficient path:

- `summary(fit)`: prints the *Dynamic coefficients per period* block (mentioned above) and the per-block posterior-mean `rho_beta`. The stationarity warning fires when, for any block, the **5th percentile** of the posterior on $\rho_\beta$ is $\ge 0.97$ **and** the IQR is below 0.1 — that is, at least 95% of the posterior mass sits above 0.97 with little spread. The print method then recommends switching to `dynamic_beta_kind = "rw1"`. The companion forecast warning (next section) uses a complementary trigger: the **upper bound** of the central 95% credible interval reaching 0.99.
- `detect_change_point(fit, threshold_bf = 5)`: heuristic regime-switch diagnostic. Computes the maximum first-difference $|\Delta \beta_t|$ scaled by $\sigma_\beta$, compares it to a prior-null reference built by resampling $(\rho, \sigma)$ from the posterior, and reports a tail-ratio score in the historical `bf` column. Large values (≥5) flag coefficients with abrupt breaks. This is a heuristic diagnostic, not a marginal-likelihood Bayes factor.

Worked example on the simulated linear-ramp data above (no real break is present, so we expect modest scores):

```{r change-point-demo}
detect_change_point(fit_db, threshold_bf = 5)
```

Each row reports the largest one-period change for one coefficient, the tail-ratio score, and a flag. The diagnostic is deliberately conservative: it fires only when a jump is large *relative to what the AR(1) prior can absorb*. A modest planted break such as `beta_t_true = c(rep(-0.5, 3), rep(1.0, 2))` does **not** reliably trip it: at $T = 5$ the score is small and noisy run to run (often 0-2, occasionally clearing the threshold of 5), because the sampler can widen $\sigma_\beta$ to absorb a single step. Only a stark break (e.g. `c(rep(-2, 3), rep(2, 2))`) pushes the scaled first-difference reliably past the prior reference (scores roughly 6-16 across runs, always clearing the threshold of 5). On the smooth linear ramp here the score is 0, as expected for a process with no abrupt jump.

Interpret the `bf` column as a bounded tail-ratio score: values below 3 are weak, values from 3 to 10 suggest a jump worth inspecting, and values above 10 indicate a jump the AR(1) prior has difficulty generating. The construction is `bf = Pr(M_post > m_star) / 0.05`, where `m_star` is the 95% quantile of the prior null, so the score is **bounded above by 20**. A value at or near 20 means the posterior path almost never looks as smooth as the prior reference, not that a separate change-point model has a formal Bayes factor of 20. The `t_hat` column points to *where* the largest scaled jump sits in posterior; treat it as a flag for visual inspection of the coefficient trajectory, not a formal break-time estimate. With very short panels, the score can be unstable run to run.

For prior elicitation *before* fitting, see `dynamic_beta_prior_summary(T = 10, kind = "ar1", rho_mean = 0.8, rho_sd = 0.15)`. It simulates paths from the prior so you can check the implied max first-difference, roughness, range, and trend correlation distributions for your chosen `T`.

#### Temporal-trend posterior-predictive check: `gof_temporal()`

The change-point diagnostic targets *coefficients*; the temporal-trend posterior-predictive check targets the *network statistic itself*. `gof_temporal()` fits a linear time trend to a chosen network statistic (density / mean / reciprocity / transitivity), computes the same slope on every `simulate(fit)` replicate, and returns a two-sided Gelman-style posterior-predictive p-value, `p_pp = 2 * min(p_up, 1 - p_up)`. This runs from 0 (observed slope in the extreme tail) to 1 (observed slope dead-centre). The print method's reading: small `p_pp` (below 0.05) means the observed temporal trend is *incompatible* with the fitted model; a large `p_pp` (toward 1) means the observed trend is *well covered* by the predictive distribution.

```{r gof-temporal-demo}
gof_temporal(fit_db, stat = "mean", n_rep = 200, seed = 1)
```

What is this telling us? `stat = "mean"` tracks the average edge weight across the network each period. In this simulation the dyadic covariate is zero-mean, so $\beta_t X_t$ averages to roughly zero *regardless* of how $\beta_t$ moves: the network mean has essentially no trend (its observed per-period slope is near 0). The high `p_pp` therefore confirms only that the fit reproduces a flat statistic; it is **not** a verdict on the (real, rising) trend in the coefficient. This is the key distinction from the change-point diagnostic above: `gof_temporal()` tests trends in a *network statistic*, while `detect_change_point()` and the per-period `coef(fit)` test trends in a *coefficient*. A static-`beta` fit also returns a high (well-covered) `p_pp` here, for the same reason -- there is no trend in the network mean for either model to miss.

`gof_temporal()` earns its keep when the substantive statistic itself drifts -- a network whose density or reciprocity climbs over time. There a static fit that cannot track the drift lands `p_pp` near 0, and the printed "incompatible" verdict points toward `dynamic_beta` (or `dynamic_ab` / `dynamic_uv`) when interpreting time-averaged summaries. To probe whether a *coefficient* trends, use the per-period `coef(fit)`, the change-point diagnostic, or a static-vs-dynamic LOO comparison.

#### Inspecting the priors actually in effect: `prior_summary()`

Modelled on `rstanarm::prior_summary()`, `prior_summary(fit)` prints the hyperparameters that were *actually used* (defaults filled in) for the AR(1) coefficient block, the variance components, and the g-prior on `beta`. Always run it once per fit; it is the cheapest way to catch a typo in a `prior = list(...)` override.

```{r prior-summary-demo}
prior_summary(fit_db)
```

#### Multi-chain $\hat R$ via `lame_parallel()` and `rhat_dynamic_beta()`

A single chain only ever buys you within-chain split-$\hat R$. For a between-chain diagnostic on the per-period coefficient path, run four parallel chains with different seeds and call `rhat_dynamic_beta()` on the per-chain fit list:

```{r rhat-dynamic-demo, eval = requireNamespace("posterior", quietly = TRUE)}
fit_list_db <- lame_parallel(
	Y_db, Xdyad = X_db_arr,
	family = "normal", R = 0,
	nscan = 200, burn = 50, odens = 5,
	dynamic_beta = "dyad",
	n_chains = 4, cores = 1,           # cores = 1 keeps the vignette portable
	combine_method = "list",            # one fit per chain
	verbose = FALSE
)

# multivariate (Brooks-Gelman) Rhat per coefficient on the length-T path,
# alongside the max univariate split-Rhat over the T per-period scalars.
rhat_dynamic_beta(fit_list_db)

# the pooled fit also routes through posterior::as_draws() with per-period
# coefficients flattened to "coef[t]" variable names, so summarise_draws()
# gives you ess_bulk / ess_tail per (coef, period).
fit_pool_db <- combine_ame_chains(fit_list_db)
posterior::summarise_draws(posterior::as_draws(fit_pool_db))
```

Use `rhat_mvt < 1.01` and `rhat_max_univariate < 1.01` as the joint pass condition for the path. The univariate column is what you compare against the Stan-era 1.01 threshold per scalar; the multivariate column catches the case where individual periods look fine but the *trajectory* disagrees across chains.

#### Sampler-failure diagnostics (the Gibbs analog of HMC divergences)

The `lame` sampler is Gibbs / Metropolis-Hastings, not HMC, so it does not produce HMC-style divergences. The closest analog is the per-block Metropolis acceptance / `tryError` count, which is exposed on every fit as `fit$mh_counters` (alias `fit$tryErrorChecks`). A post-MCMC warning fires automatically when any block exceeds a 10% failure rate; for routine reporting:

```{r mh-counters-demo, eval = FALSE}
# per-block failure counts and acceptance proxies for the just-fit chain
str(fit_db$mh_counters)
```

Treat any block with `>5%` failures as suspicious; combine that with the `rhat_dynamic_beta()` and `summarise_draws()` output above for a full diagnostic pass. Pairing `ess_bulk` / `ess_tail` from `summarise_draws()` with the `mh_counters` totals gives you the Gibbs-side equivalent of the Stan diagnostic triple (R-hat, ESS, divergences).

### Forecasting

When at least one dynamic component is active, `predict(fit, h = K)` propagates the AR(1) / RW1 state-space model forward $K$ periods and returns a list of $K$ matrices (one per future period). The propagation samples $(\rho_\beta, \sigma_\beta)$ from the posterior at each draw, so the resulting forecast accounts for hyperparameter uncertainty.

```{r forecast-example}
# h = 3-step forecast on the link scale
fc_link <- predict(fit_db, h = 3, type = "link")
length(fc_link)              # 3 (one matrix per future period)
dim(fc_link[[1]])            # n x n

# response scale applies the family inverse link per draw
fc_resp <- predict(fit_db, h = 3, type = "response")

# by_draw = TRUE returns an [n, n, h, n_draws] array of per-draw forecasts
fc_full <- predict(fit_db, h = 3, type = "response", by_draw = TRUE)
dim(fc_full)                 # n x n x 3 x n_draws

# interval = "credible" returns lower / median / upper per horizon
fc_ci <- predict(fit_db, h = 3, type = "response", interval = "credible")
str(fc_ci[[1]])              # list($lower, $median, $upper)
```

Forecast variance behavior depends on `dynamic_beta_kind`. For AR(1) with $|\rho_\beta| < 1$ the $h$-step conditional variance is $\sigma_\beta^2 (1 - \rho_\beta^{2h}) / (1 - \rho_\beta^2)$ and saturates at the stationary variance $\sigma_\beta^2 / (1 - \rho_\beta^2)$ as $h \to \infty$. The saturation level itself explodes as $\rho_\beta \to 1$, so AR(1) forecasts beyond a few periods become uninformative in that regime even though the variance technically converges. For RW1 (the unit-root case) the variance grows exactly linearly in $h$, contributing $\sigma_\beta^2 \cdot h$ per step (so the $h$-step conditional variance from the RW1 piece alone is $h \cdot \sigma_\beta^2$, with no saturation). `predict()` emits a one-per-fit warning when the **upper bound** of the central 95% posterior credible interval for $\rho_\beta$ reaches 0.99 (i.e. `quantile(RHO_BETA, 0.975) >= 0.99`); note this 0.99 forecast trigger is intentionally higher than the 0.97 stationarity flag used by `summary(fit)`. In that regime, prefer `dynamic_beta_kind = "rw1"` if the underlying process is genuinely non-stationary, and treat horizons beyond $h \approx 3$ cautiously in either case. **Forecasting under `"rw2"` / `"matern32"` is supported but currently uses the same AR(1) forward recursion** with $\rho_\beta$ pinned at 1 (since neither prior identifies an AR(1) persistence), so the forecast variance grows linearly in $h$ just like RW1; the smoothing or length-scale structure of these priors informs the *posterior path of $\beta$ in-sample* but does not shape the out-of-sample propagation.

The `predict()` warning copy is kind-aware: AR(1) reports the stationary-variance saturation explosion as $\rho_\beta \to 1$, RW1 / RW2 / Matern 3/2 report the linear variance growth, and the "refit with `dynamic_beta_kind = \"rw1\"`" suggestion is only printed under AR(1) (so it does not nag a user who is already on the unit-root branch).

You can pass `newdata = list_of_future_X_arrays` to combine counterfactual future covariates with the forecast (one $n \times n \times p$ array per future period).

For `family = "poisson"` fits that supplied `period_exposure` at fit time, the future-period exposure is passed through `predict(..., newexposure = ...)`. The forecast multiplies $\exp(\eta)$ by `newexposure[k]` at horizon $k$, so the response-scale matrix is reported in the same units as the original `Y`. If `newexposure` is omitted, the forecast reuses the last observed `period_exposure` value (and falls back to 1 for fits that did not use exposure offsets).

```{r forecast-newexposure, eval = FALSE}
# poisson fit with period_exposure = c(1, 5, 25, 100)
# pass the future exposure explicitly when h > 0:
fc_pois <- predict(fit_pois, h = 2, type = "response",
                   newexposure = c(200, 400))
```

### Visualising the coefficient path: `autoplot.lame`

For a fit with `dynamic_beta` active, `autoplot(fit)` returns a `ggplot` with a ribbon plot per coefficient (posterior median line, 95% interval ribbon, faceted by coefficient). The S3 method is registered against `ggplot2::autoplot` so plain `autoplot(fit)` dispatches when `ggplot2` is loaded.

```{r autoplot-example, fig.width = 6, fig.height = 4, fig.alt="Ribbon plot of posterior coefficient paths across time, faceted by coefficient, with the median as a line and the 95 percent credible interval as a shaded band."}
library(ggplot2)
autoplot(fit_db, probs = c(0.025, 0.5, 0.975)) +
	labs(title = "Posterior coefficient paths",
	     subtitle = "Median line, 95% credible interval ribbon",
	     x = "Time period", y = "Coefficient value")
```

### LOO / WAIC via `save_log_lik = TRUE`

Pass `save_log_lik = TRUE` and `lame` attaches a `[n_stored, n_obs]` pointwise log-likelihood matrix to `fit$log_lik`. From there, `loo::loo(fit)` and `loo::waic(fit)` work directly because the package registers S3 methods that hand the stored matrix to `loo::loo.matrix()`:

```{r save-log-lik-eval, eval = requireNamespace("loo", quietly = TRUE)}
fit_ll <- lame(Y_db, Xdyad = X_db_arr, family = "normal", R = 0,
               nscan = 200, burn = 50, odens = 5,
               dynamic_beta = "dyad",
               save_log_lik = TRUE,
               verbose = FALSE)

dim(fit_ll$log_lik)        # [n_stored, n_observations]
loo_db <- loo::loo(fit_ll)
print(loo_db)
```

One important detail for cross-software comparisons. For six families -- `normal`, `binary`, `cbin`, `tobit`, `poisson`, `ordinal` -- the stored pointwise log-likelihood is the **exact family-specific Y density** on the response scale (Gaussian density for `normal`, probit Bernoulli log-pmf for `binary`/`cbin`, censored-Gaussian for `tobit`, log-Poisson for `poisson`, and the cumulative-probit log-pmf for `ordinal`). For these six, `elpd_loo` is directly comparable to a `loo()` result from a Stan / brms fit to the same family on the same data, modulo the `lame` Gibbs sampler's $\hat R$ and ESS being satisfactory. For the two rank likelihoods `frn` and `rrl`, the exact marginal needs GHK Monte Carlo and is currently only available on the longitudinal path via `log_lik_method = "observed_ghk"`; the default `observed_exact` for those two families falls back to the augmented-Z normal density and emits a one-time warning. Inspect `fit$log_lik_method` on any fit to see which branch was used.

#### Memory-conscious variant: `save_log_lik = "chunked"`

If the per-iteration log-likelihood matrix would exceed RAM (large `n_stored x n_obs`), pass `save_log_lik = "chunked"` instead to stream it to per-column-chunk binary files. The chunks are recovered transparently when you call `loo(fit)` (or directly via `read_log_lik(fit)`). The on-disk layout is row-major within each chunk so one `readBin` per chunk suffices.

```{r save-log-lik-chunked, eval = FALSE}
fit_chk <- lame(Y_db, Xdyad = X_db_arr, family = "normal", R = 0,
                nscan = 1000, burn = 200, odens = 5,
                dynamic_beta = "dyad",
                save_log_lik = "chunked",
                log_lik_chunk_size = 5000L,
                verbose = FALSE)

loo::loo(fit_chk)   # reads the chunks back automatically
```


### Prior Specifications

The AR(1) coefficients and innovation variances have sensible default priors:

- $\rho_{uv} \sim \text{TruncNormal}(0.9, 0.1, 0, 1)$, centered on high persistence
- $\rho_{ab} \sim \text{TruncNormal}(0.8, 0.15, 0, 1)$, slightly less persistent
- $\sigma_{uv}^2, \sigma_{ab}^2 \sim \text{InverseGamma}(2, 1)$

These can be customized via the `prior` argument if you have strong beliefs about the rate of change.

## Genuine Dynamics: What They Look Like

We start with an example where dynamics are truly present, so you can see what the model recovers when there is a real signal. We simulate 25 actors over 5 periods where each actor's latent position evolves via an AR(1) process with $\rho = 0.85$ and innovation SD $\sigma_{uv} = 0.3$ (matching the **innovation parameterisation** the C++ sampler uses). Initial positions are drawn at SD 1.5, which is larger than the implied stationary SD $0.3/\sqrt{1 - 0.85^2} \approx 0.57$. This seeds the simulation with strong latent structure that the AR(1) then drifts down toward stationarity, producing a network with realistic density (roughly 30%) after a negative intercept.

```{r, message=FALSE}
library(lame)
set.seed(6886)

n_dyn <- 25
n_per <- 5
R_dyn <- 2

# evolving latent positions with AR(1) dynamics
U <- matrix(rnorm(n_dyn * R_dyn, 0, 1.5), n_dyn, R_dyn)
V <- matrix(rnorm(n_dyn * R_dyn, 0, 1.5), n_dyn, R_dyn)

Y_dyn <- list()
for(t in 1:n_per) {
	if(t > 1) {
		U <- 0.85 * U + matrix(rnorm(n_dyn * R_dyn, 0, 0.3), n_dyn, R_dyn)
		V <- 0.85 * V + matrix(rnorm(n_dyn * R_dyn, 0, 0.3), n_dyn, R_dyn)
	}
	eta <- -1 + U %*% t(V)
	Y_t <- matrix(rbinom(n_dyn * n_dyn, 1, pnorm(eta)), n_dyn, n_dyn)
	diag(Y_t) <- NA
	rownames(Y_t) <- colnames(Y_t) <- paste0("A", 1:n_dyn)
	Y_dyn[[t]] <- Y_t
}

# iterations are small so the vignette builds quickly
# use burn >= 1000 and nscan >= 5000 for real analyses.
fit_dyn_real <- lame(Y_dyn, R = 2,
	dynamic_uv = TRUE, dynamic_ab = TRUE,
	family = "binary",
	burn = 100, nscan = 1000, odens = 10,
	verbose = FALSE, plot = FALSE)

summary(fit_dyn_real)
```

The trajectory plot shows how actors move through the latent space over time. With genuine temporal structure, the paths should show coherent movement rather than random jumping:

```{r uv-traj-real, fig.width=7, fig.height=5, fig.alt="Trajectory plot of each actor's path through the 2D latent space across five periods under genuine AR(1) dynamics, showing coherent and gradual movement."}
uv_plot(fit_dyn_real, plot_type = "trajectory")
```

With more than eight actors the trajectory legend becomes a rainbow that loses meaning. Pass `highlight =` a small character vector of actor names to grey out the rest and colour only the highlighted set with the colour-blind-safe Okabe--Ito palette:

```{r uv-traj-highlight, fig.width=7, fig.height=5, fig.alt="Same trajectory plot as above but only four named actors (A1, A5, A10, A15) are coloured with the Okabe-Ito palette; remaining actors are drawn in light grey so the comparison reads at a glance."}
uv_plot(fit_dyn_real, plot_type = "trajectory",
        highlight = c("A1", "A5", "A10", "A15"))
```

Always check convergence. Dynamic models have more parameters than static models, so adequate mixing is harder to achieve and requires longer chains:

```{r trace-dyn-real, fig.width=7, fig.height=5, fig.alt="MCMC trace and density plots for the dynamic-effects fit on truly correlated data; well-mixed traces indicate the longer sampler ran long enough."}
trace_plot(fit_dyn_real)
```

## What Happens Without Temporal Structure?

To understand how the dynamic model behaves when there is no signal, we fit the same model to independent networks (no temporal correlation). This is a useful null baseline: the estimated `rho` values tell you what the prior alone produces when the data are uninformative.

```{r}
set.seed(6886)
n <- 30
n_periods <- 5

# independent binary networks with no temporal structure
Y_list <- list()
for(t in 1:n_periods) {
	Y_t <- matrix(rbinom(n*n, 1, 0.2), n, n)
	diag(Y_t) <- NA
	rownames(Y_t) <- colnames(Y_t) <- paste0("Actor", 1:n)
	Y_list[[t]] <- Y_t
}

# fit with a custom prior on rho_uv to illustrate prior sensitivity
prior_custom <- list(
	rho_uv_mean = 0.95,    # expect very slow change in latent positions
	rho_uv_sd = 0.05       # tight prior
)

fit_null <- lame(
	Y = Y_list,
	R = 2,
	dynamic_uv = TRUE,
	dynamic_ab = TRUE,
	family = "binary",
	prior = prior_custom,
	burn = 100,
	nscan = 1000,
	odens = 10,
	verbose = FALSE,
	plot = FALSE
)

summary(fit_null)
```

We can compare the estimated persistence parameters across the two settings:

```{r}
cat("Independent data rho_uv:", round(mean(fit_null$rho_uv), 3), "\n")
cat("Correlated data rho_uv:", round(mean(fit_dyn_real$rho_uv), 3), "\n")
```

With only 5 time periods you might expect the prior to dominate, and yet the
posterior means above can be substantially different from each other and from
the prior centre of 0.95. The correlated fit lands in the upper-AR(1) range
(the posterior mean is around 0.94 here; the true 0.85 actually falls *below*
the 95% credible interval, because at this short panel length the well-known
small-sample upward bias of AR(1) estimators is large enough that even the
interval misses the truth), while the null fit collapses well
below the prior centre because the data carry no period-to-period correlation
for the prior to anchor against. In other words: even a deliberately tight
prior centred near 1 will not always pull a clearly nonsensical posterior back
toward 1, and the strength of that pull depends on the data. With even a modest
panel the AR(1) parameter is identifiable enough to do real work, but the
posterior mean is not a substitute for inspecting the trajectory plots, which
are more informative than the point estimate of `rho_uv` alone
(compare the coherent paths above with the erratic paths below).

### Visualizing the Null Case

The trajectory plot for independent data looks different: paths are more erratic, without the smooth coherence of genuinely correlated positions.

```{r uv-traj-null, fig.width=7, fig.height=5, fig.alt="Trajectory plot of latent positions for the null model fit on independent networks; paths look erratic and lack the smooth coherence of genuine AR(1) data."}
uv_plot(fit_null, plot_type = "trajectory")
```

The additive effects show similar behavior. Without genuine temporal structure, the trajectories reflect noise rather than real shifts in actor activity.

```{r ab-traj-null, fig.width=7, fig.height=5, fig.alt="Trajectory plot of sender additive effects for the null model fit on independent networks; per-actor lines fluctuate without persistent trend, reflecting noise rather than temporal signal."}
ab_plot(fit_null, effect = "sender", plot_type = "trajectory")
```

## Comparing Model Specifications

A natural workflow is to fit several specifications and compare their goodness-of-fit. We use the independent data from above to show that dynamic effects add nothing when there is no signal.

```{r}
# four small fits to compare specifications. nscan is intentionally
# small here so the vignette stays under CRAN's vignette compute
# budget; for a real comparison use nscan >= 2000 per fit and run them
# under `ame_parallel()` with 4 chains.
fit_static <- lame(Y_list, R = 2, family = "binary",
									burn = 50, nscan = 300, odens = 5,
									verbose = FALSE, plot = FALSE)

fit_uv <- lame(Y_list, R = 2, dynamic_uv = TRUE, family = "binary",
							burn = 50, nscan = 300, odens = 5,
							verbose = FALSE, plot = FALSE)

fit_ab <- lame(Y_list, R = 2, dynamic_ab = TRUE, family = "binary",
							burn = 50, nscan = 300, odens = 5,
							verbose = FALSE, plot = FALSE)

fit_full <- lame(Y_list, R = 2, dynamic_uv = TRUE, dynamic_ab = TRUE,
								family = "binary",
								burn = 50, nscan = 300, odens = 5,
								verbose = FALSE, plot = FALSE)
```

Compare the GOF statistics across models. For each specification, we compute posterior predictive p-values: how often do the simulated network statistics exceed the observed ones? Values near 0.5 indicate good fit.

```{r}
# compute p-values for each model and each GOF statistic
compute_pvals <- function(fit) {
	gof <- fit$GOF
	sapply(names(gof), function(stat) {
		mat <- gof[[stat]]
		obs <- mat[, 1]           # first column = observed
		sims <- mat[, -1]         # remaining = posterior predictive
		mean(colMeans(sims) >= mean(obs))
	})
}

gof_comparison <- rbind(
	Static     = compute_pvals(fit_static),
	Dynamic_UV = compute_pvals(fit_uv),
	Dynamic_AB = compute_pvals(fit_ab),
	Full       = compute_pvals(fit_full)
)
round(gof_comparison, 3)
```

Since the data were generated without temporal structure, the dynamic specifications should not systematically out-fit the static model. With short chains and a small panel these posterior-predictive p-values are noisy and will not line up exactly across specifications -- so read the comparison as "no specification is clearly better here", rather than expecting identical numbers or values pinned near 0.5.

## When to Use Dynamic Effects

**Use `dynamic_ab`** when you suspect actors' overall activity levels change over time. This is common in many settings: countries go through isolationist vs. interventionist periods, users churn in and out of platforms, students become more or less engaged across semesters.

**Use `dynamic_uv`** when you suspect the underlying community structure is shifting. This is a stronger claim, not just that actors are more or less active, but that the pattern of *who connects with whom* is changing. Examples include political realignment, market disruption, or generational turnover in a social network.

**Use `dynamic_beta`** when you suspect a covariate's *effect* on tie formation changes over time, while the actors themselves and the community structure are stable. Example: trade's effect on alliance formation may strengthen during economic recessions and weaken otherwise; peer influence may matter more in some semesters than others. `dynamic_beta` is the right tool when the question is *how strongly X translates into Y*, not *who is connecting to whom* (which is `dynamic_uv`'s territory) or *who is generally active* (which is `dynamic_ab`'s).

**Use multiple together** when you believe several types of change are happening simultaneously. This is the most flexible specification but also the most data-hungry. With short panels (few time periods) or sparse networks, the dynamic parameters may not be well-identified, and you might be better off with a simpler specification.

**Stick with static** when you have few time periods, when the network structure is genuinely stable, or when you primarily care about the *average* covariate effects rather than how they evolve.

## Dealing with Rotational Indeterminacy

One subtlety of dynamic latent space models: the latent space is only identified up to an orthogonal rotation **at each time point** *and* **at each posterior draw**. Even if actor positions are evolving smoothly under the truth, the raw estimated $U_t$ matrices can appear to "jump" between periods due to arbitrary rotations, and posterior means across draws can shrink toward zero when different draws sit in different rotations of the latent space.

The `procrustes_align()` function applies the standard orthogonal-Procrustes rotation (Gower 1975) to align each time period's positions to the previous one:

```{r}
# align latent positions across time
aligned <- procrustes_align(fit_null)
str(aligned$U)  # 3D array: actors x dimensions x time
```

**Methodological note.** By default, `procrustes_align()` aligns the **posterior-mean** trajectory `fit$U` (an `n x R x T` array of posterior means) across time only. The Hoff (2005) / Sewell--Chen (2015) standard is to Procrustes-align each posterior draw to a reference (the first draw, or the posterior mean) *before* computing per-period summaries, because rotation indeterminacy is a per-draw property, not a per-mean one. The `per_draw = TRUE` argument is the hook for that workflow -- it looks for a 4-D per-draw trajectory cube `fit$U_full` / `fit$V_full` (`[n, R, T, S]`). **That cube is not currently persisted by either `ame()` or `lame()`**, so `per_draw = TRUE` emits an informational note and falls back to mean-trajectory alignment on every fit today; full per-draw alignment is a planned extension. In the meantime, for dynamic latent positions the trajectory view (`uv_plot(plot_type = "trajectory")`) is more appropriate than per-period credible intervals on the raw `U_t` entries.

You can also extract aligned positions as a tidy data frame using `latent_positions()` with `align = TRUE` (the default for `lame` objects):

```{r}
lp <- latent_positions(fit_null, align = TRUE)
head(lp)
```

This data frame is ready for `ggplot2` if you want to build custom trajectory plots (e.g., highlighting specific actors or overlaying external events).

## Parallel chains with `ame_parallel()` / `lame_parallel()`

Outside the dynamic-$\beta$ Rhat workflow above, the same parallel wrapper is the recommended way to (a) get a between-chain Rhat on **any** parameter and (b) cut wall-clock time roughly linearly in `n_chains` when you have idle cores. `ame_parallel()` auto-dispatches to `ame()` for cross-sectional data and `lame()` for a list / 3-D array; `lame_parallel()` is a thin wrapper that forces the longitudinal fitter. Both accept the full `ame()` / `lame()` argument list via `...`.

```{r parallel-worked-example, eval = FALSE}
# 4 chains, one per core, pooled into a single fit.
# `cores` defaults to `n_chains`; the wrapper clamps `cores` to
# `parallel::detectCores(logical = TRUE)` and emits a `cli_warn` when you
# over-request, so a stray `cores = 16` on a 4-core laptop will be capped
# (not silently honoured). For a *physical*-core cap (avoiding hyperthread
# oversubscription on BLAS-heavy workloads), pass it yourself:
n_phys <- max(1L, parallel::detectCores(logical = FALSE) - 1L)
data(YX_bin_list)

fit_pool <- ame_parallel(
  Y               = YX_bin_list$Y,
  Xdyad           = YX_bin_list$X,
  family          = "binary",
  R               = 2,
  rvar = TRUE, cvar = TRUE, dcor = TRUE,
  burn = 1000, nscan = 4000, odens = 25,
  n_chains        = 4,
  cores           = min(4, n_phys),   # set to 1 for sequential
  combine_method  = "pool",            # one merged fit; use "list" for per-chain
  verbose         = FALSE
)

summary(fit_pool)            # pooled posterior, 4x the draws of a single chain

# `compute_mcmc_diagnostics()` reports per-parameter between-chain Rhat and
# bulk/tail ESS off the pooled fit (when chains are preserved internally):
compute_mcmc_diagnostics(fit_pool)
```

A practical note on scale. Each chain runs in its own R worker, so the memory cost is roughly `n_chains x` the single-chain footprint; for a network with $n \approx 200$ and `nscan = 4000` that is a few hundred MB per chain, fine on a laptop; for $n \approx 1000$ plan on dozens of GB if you run `n_chains = 8` in parallel and prefer fewer, longer chains. **There is no GPU backend.** All numerical work is CPU via RcppArmadillo, so the right scaling lever is cores plus the fast point estimator (`ame_als()` / `lame_als()`, see the [fast estimation vignette](fast_estimation.html)) for $n \gtrsim 500$ exploratory passes; reserve the full MCMC fit for the final calibrated run.

## Implementation Notes

The dynamic effects are implemented in C++ via Rcpp and RcppArmadillo. The AR(1) updates use Gibbs sampling with block updates across all actors at each time point, which is efficient for large networks.

## References

1. **Hoff, PD (2021)**. Additive and Multiplicative Effects Network Models. *Statistical Science* 36, 34--50.

2. **Hoff, PD (2005)**. Bilinear mixed-effects models for dyadic data. *Journal of the American Statistical Association* 100(469), 286--295. The original AME formulation underlying `amen` and `lame`.

3. **Sewell, D. K., & Chen, Y. (2015)**. Latent space models for dynamic networks. *Journal of the American Statistical Association*, 110(512), 1646-1657.

4. **Durante, D., & Dunson, D. B. (2014)**. Nonparametric Bayes dynamic modeling of relational data. *Biometrika*, 101(4), 883-898.

5. **Carter, C. K., & Kohn, R. (1994)**. On Gibbs sampling for state space models. *Biometrika* 81(3), 541--553. The original forward-filter / backward-sample algorithm used by `dynamic_beta_kind = "ar1"` / `"rw1"`.

6. **Frühwirth-Schnatter, S. (1994)**. Data augmentation and dynamic linear models. *Journal of Time Series Analysis* 15(2), 183--202. Independent derivation of FFBS for dynamic linear models.

7. **Gower, J. C. (1975)**. Generalized Procrustes analysis. *Psychometrika* 40(1), 33--51. The orthogonal-Procrustes rotation underlying `procrustes_align()`.
