---
title: "Fast (MCMC-free) AME Estimation"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Fast (MCMC-free) AME Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>"
)
library(lame)
```

## When to use the fast estimator

`ame()` and `lame()` fit AME models by Bayesian MCMC, which gives calibrated
posterior inference but can take minutes. For rapid model exploration, trying several ranks, screening covariates, or getting starting values, the
package also ships a fast, MCMC-free point estimator:

- `ame_als()`: cross-sectional networks
- `lame_als()`: longitudinal (replicated) networks
- `ame_als_bootstrap()`: bootstrap standard errors and intervals

It fits the same model

  z_ij = mu + beta'x_ij + a_i + b_j + u_i'v_j + e_ij

by **iterative block coordinate descent** rather than Gibbs sampling. It is
typically tens to hundreds of times faster than `ame()`, and the point
estimates are usually close. Use it to explore; use `ame()` / `lame()` for the
inference you report.

### Wall-clock benchmark

The numbers below are indicative for normal-family AME with `R = 2`;
absolute timings depend heavily on CPU, BLAS, and whether you are on a
high-clock workstation or a virtualised / WSL host. On the dev box
(WSL2, single-threaded BLAS) the same benchmark ran roughly 2-4× slower
than the workstation reference column, so treat these as order-of-
magnitude. The point estimator scales O(n² · iter) per BCD pass and is
dominated by the linear system at each block; the MCMC scales similarly
per iteration but pays the constant of running thousands of iterations.
The "ratio" column is what you actually feel: the ALS fit completes
before MCMC finishes its burn-in for any reasonable n.

| n   | `ame()` MCMC (4000 iter) | `ame_als()` | Speed ratio |
|-----|--------------------------|-------------|-------------|
| 50  | 5-25 s                   | 0.1-0.5 s   | 25-250×     |
| 100 | 15-50 s                  | 0.3-1 s     | 30-200×     |
| 200 | 60-120 s                 | 2-3 s       | 30-60×      |
| 500 | 7-15 min                 | 15-30 s     | 25-40×      |

These ranges are wide on purpose: under a single-threaded BLAS on a
virtualised host, the small-n speed ratios we have measured are several
times larger than the "≈ 30×" figure quoted in earlier docs, because the
MCMC's per-iteration overhead is fixed and the ALS solve is nearly free
at n ≤ 100. Treat the ratio as "ALS finishes before MCMC's burn-in"
rather than a number to plan against.

For binary/poisson the **point-estimate** speedup ratio is much smaller
(often only 3-10× on small networks) because the IRLS path runs 3-5
reweighting iterations per fit, and because the MCMC binary sampler is
itself cheaper per iteration than the normal sampler. Bipartite and
unipartite ALS run at broadly similar cost for comparable dimensions;
which is faster depends on the exact shape and hardware, so do not read
much into small differences.

**The bootstrap can dominate.** A `bootstrap = 100` parametric bootstrap
runs the full ALS fit 101 times. For a binary `ame_als(R = 1, bootstrap = 100)`
at n = 50, we measured ≈ 135 s --- slower than the equivalent MCMC fit.
If you only need the point estimate, skip the bootstrap (`bootstrap = 0`,
the default) and use the sandwich `vcov()`/`confint()` on the fit. If
you need calibrated intervals at scale, `ame()` MCMC is often the
cheaper option even though it samples for thousands of iterations.

To benchmark on your own hardware:

```{r benchmark_self, eval = FALSE}
set.seed(1)
n <- 100
Y <- matrix(rnorm(n*n), n, n); diag(Y) <- NA
rownames(Y) <- colnames(Y) <- paste0("a", sprintf("%03d", 1:n))

t_als <- system.time(ame_als(Y, R = 2, family = "normal", verbose = FALSE))
t_mcmc <- system.time(ame(Y, R = 2, family = "normal", burn = 500,
                          nscan = 4000, odens = 25, verbose = FALSE, plot = FALSE))
cat("ALS:  ", round(t_als["elapsed"], 2), "s\n",
    "MCMC: ", round(t_mcmc["elapsed"], 2), "s\n",
    "ratio:", round(t_mcmc["elapsed"] / t_als["elapsed"], 1), "x\n")
```

### What ALS can do vs. what only MCMC does

ALS is a fast point estimator, not a drop-in replacement for the Bayesian
MCMC fits. The current coverage is:

| Capability                                | `ame()` / `lame()` (MCMC) | `ame_als()` / `lame_als()` |
|-------------------------------------------|---------------------------|----------------------------|
| Families: normal, binary                  | yes                        | yes                         |
| Family: poisson                           | yes                        | yes                         |
| Families: ordinal / tobit / cbin / frn / rrl | yes                     | no (error)                  |
| Unipartite / bipartite mode               | yes                        | yes                         |
| `symmetric = TRUE`                        | yes                        | yes                         |
| `R = 0..min(n)-1`                         | yes                        | yes                         |
| Xrow / Xcol / Xdyad covariates            | yes                        | yes                         |
| Reproducibility via `seed=`               | yes                        | yes (deterministic ALS)     |
| `coef`, `vcov`, `fitted`, `residuals`     | yes                        | yes                         |
| `confint`                                 | posterior quantile         | sandwich Wald OR bootstrap  |
| `predict(newdata=)`                        | yes                        | yes                         |
| `summary`, `print`, `latent_positions`    | yes                        | yes                         |
| `nobs`                                    | yes                        | yes                         |
| `simulate`                                | yes                        | yes                         |
| `gof_plot`                                | posterior predictive       | bootstrap-based             |
| `ab_plot`, `uv_plot`                      | yes                        | yes                         |
| `prior_summary`                           | yes                        | yes (reports "no priors")   |
| `dynamic_uv` / `dynamic_ab` (AR(1))       | yes                        | **no (static pooled)**      |
| Custom priors (`prior = list(...)`, `g`)  | yes                        | **no (no priors)**          |
| Multi-chain (`n_chains`, via `ame_parallel`) | yes                     | no (use `bootstrap = N`)    |
| `posterior_opts` (save U/V/a/b samples)   | yes                        | no (point estimator)        |
| `trace_plot` (Rhat / ESS)                 | yes                        | no (deterministic)          |
| `custom_gof`, `periodic_save`             | yes                        | no                           |

When you call the unified front door with MCMC-only arguments, the ALS
dispatcher warns and lists exactly which ones it ignored -- for the
arguments each entry point actually accepts. Two caveats on the edges:

- `nscan` / `burn` / `odens` / `prior` / `g` are accepted-then-ignored
  with a warning by both `ame(method = "als")` and `lame(method = "als")`.
- The longitudinal-only arguments (`dynamic_uv`, `dynamic_ab`,
  `dynamic_beta`, `time_index`, ...) are warned-and-ignored by
  `lame(method = "als")`, but passing them to `ame()` (cross-sectional)
  raises the usual "longitudinal-only argument" *error* before ALS runs,
  because they are not valid `ame()` arguments at all.
- `n_chains` is an `ame()` / `ame_parallel()` argument; it is not a
  `lame()` formal, so `lame(method = "als", n_chains = ...)` raises an
  "unused argument" error rather than a warn-and-ignore. Use
  `bootstrap = N` for ALS uncertainty regardless of entry point.

## A first fit

```{r}
set.seed(1)
n  <- 40
a  <- rnorm(n, 0, 0.5); b <- rnorm(n, 0, 0.5)
Xd <- matrix(rnorm(n * n), n, n)
Y  <- 0.5 + 0.8 * Xd + outer(a, b, "+") + matrix(rnorm(n * n), n, n)
diag(Y) <- NA

fit <- ame_als(Y, Xdyad = Xd, R = 1, family = "normal", verbose = FALSE)
fit
```

`coef()`, `fitted()`, `residuals()`, `predict()` and `plot()` all work as for
any model object. The printed variance components (`va`, `vb`, `cab`, `rho`,
`ve`) are descriptive summaries of the fitted effects, not random-effect
variance components.

## Choosing the rank R

There is no information criterion for `R`. Fit a few values and look for an
elbow in the deviance / SSE:

```{r}
devs <- sapply(0:3, function(r)
	ame_als(Y, Xdyad = Xd, R = r, family = "normal",
	            verbose = FALSE)$deviance)
data.frame(R = 0:3, deviance = round(devs, 1))
```

If the curve declines smoothly with no clear elbow, keep `R` small and confirm
the chosen rank with an `ame()` fit.

## Non-normal families

For `binary` and `poisson` data the default `non_normal_method = "irls"` runs
an iteratively reweighted least squares loop, returning coefficients on the
calibrated link scale (probit for binary, log for poisson). The alternative
`non_normal_method = "transform"` is faster but its coefficients are on an
uncalibrated rank scale, good for the sign and ranking of effects but not their
magnitude.

```{r}
set.seed(2)
Yb <- 1 * (Y > median(Y, na.rm = TRUE)); diag(Yb) <- NA
fb <- ame_als(Yb, Xdyad = Xd, R = 0, family = "binary", verbose = FALSE)
coef(fb)
```

## Uncertainty

The point estimator carries no uncertainty. Use the bootstrap:

```{r}
bt <- ame_als_bootstrap(fb, R = 100, type = "parametric",
                            seed = 1, verbose = FALSE)
summary(bt)
```

`type = "parametric"` (the default) simulates fresh outcomes from the fitted
model and refits; `type = "block"` resamples time slices and is available for
longitudinal fits with several time points. `confint()` and `vcov()` work on
the bootstrap object.

A fast analytic alternative for the **regression coefficients only** is the
conditional sandwich covariance, `vcov()` / `confint()` applied directly to the
fit. It is anti-conservative (it holds the additive and multiplicative effects
fixed); the bootstrap is the recommended tool for full inference.

### ALS bootstrap vs MCMC posterior on the same data

A useful sanity check on the fast estimator is to fit the same data
with both engines and overlay the coefficient intervals. With a
calibrated link (`non_normal_method = "irls"` for binary, the default)
the ALS bootstrap should give a coefficient story qualitatively
similar to the MCMC posterior on cases the model fits well; large
discrepancies are a signal to use MCMC for inference.

```{r als-vs-mcmc, fig.width = 7, fig.height = 3.5, fig.alt="Side-by-side coefficient plot comparing the ALS bootstrap and MCMC posterior on the same binary network. Each coefficient (intercept and the single dyadic covariate) has two horizontal intervals: one for the ALS bootstrap, one for the MCMC posterior. A dashed vertical line at zero marks the no-effect reference; intervals on the same side of zero and with overlapping range indicate the two estimators agree."}
# wrap Xd into a 3-D array with an explicit slice name so the ALS and MCMC
# paths produce identical coefficient names (the auto-naming default differs
# between the two engines)
Xd_arr <- array(Xd, dim = c(nrow(Xd), ncol(Xd), 1),
                dimnames = list(NULL, NULL, "Xd"))

# refit the binary ALS fit with the bootstrap attached so tidy() returns
# bootstrap-based intervals
fb_b <- ame_als(Yb, Xdyad = Xd_arr, R = 0, family = "binary",
                verbose = FALSE, bootstrap = 100, bootstrap_seed = 1)

# fit the same data with MCMC, short chain (this is a 40-actor binary
# fit; sub-second wall-clock for a 200-sample posterior summary)
fit_mcmc <- ame(Yb, Xdyad = Xd_arr, family = "binary", R = 0,
                burn = 200, nscan = 1000, odens = 5,
                verbose = FALSE, plot = FALSE, gof = FALSE,
                seed = 1)

# tidy both; the term names match, and the four columns we select
# (term, estimate, conf.low, conf.high) are common to both engines
# (ALS tidy adds an extra se_source column, which we drop here)
als_tdy  <- broom::tidy(fb_b)
mcmc_tdy <- broom::tidy(fit_mcmc)
df <- rbind(
    cbind(estimator = "ALS bootstrap",
          als_tdy[, c("term", "estimate", "conf.low", "conf.high")]),
    cbind(estimator = "MCMC posterior",
          mcmc_tdy[, c("term", "estimate", "conf.low", "conf.high")])
)

library(ggplot2)
ggplot(df, aes(x = estimate, y = term,
                colour = estimator, shape = estimator)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                    position = position_dodge(width = 0.45),
                    size = 0.5) +
    scale_colour_manual(values = c("ALS bootstrap"  = "#0072B2",
                                    "MCMC posterior" = "#D55E00")) +
    scale_shape_manual(values  = c("ALS bootstrap"  = 16,
                                    "MCMC posterior" = 17)) +
    labs(x = "coefficient (probit scale)", y = NULL,
         colour = NULL, shape = NULL) +
    theme_bw() +
    theme(panel.border    = element_blank(),
          axis.ticks      = element_blank(),
          legend.position = "bottom")
```

Two intervals on the same side of zero with substantial overlap
means ALS is a fine point estimate for that coefficient; intervals
that disagree on sign or location are a flag to invest the MCMC
budget for that fit.

## Longitudinal data

`lame_als()` is the longitudinal counterpart. Its effects are static
(pooled across time), as in a non-dynamic `lame()` fit.

```{r}
set.seed(3)
Yl <- replicate(5, {
	# intercept 0.4 + sender/receiver structure + noise (no dyadic
	# covariate is passed below, so only the intercept is recoverable)
	m <- 0.4 + outer(a, b, "+") + matrix(rnorm(n * n), n, n)
	diag(m) <- NA
	m
}, simplify = FALSE)
lf <- lame_als(Yl, R = 1, family = "normal", verbose = FALSE)
coef(lf)
```

## Summary

| Task | Function |
|------|----------|
| Fast cross-sectional point estimate | `ame_als()` |
| Fast longitudinal point estimate | `lame_als()` |
| Bootstrap standard errors / intervals | `ame_als_bootstrap()` |
| Calibrated posterior inference | `ame()` / `lame()` |

The fast estimator is for exploration. When you report results, confirm them
with the MCMC estimator.
