---
title: "Forecasting Longitudinal Networks with lame"
author: "Cassy Dorff, Tosin Salau, Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasting Longitudinal Networks with lame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>",
	fig.width = 7,
	fig.height = 5,
	eval = TRUE
)
```

## What this vignette covers

`lame()` with at least one dynamic component (`dynamic_beta`,
`dynamic_ab`, `dynamic_uv`) gives you a state-space model that is
exactly the right object for forecasting future periods. This vignette
walks through:

1. **Fit a model with dynamic coefficients.**
2. **Forecast `h` periods ahead** on the link scale and the response
   scale.
3. **Compute counterfactual forecasts** by feeding in alternative
   future covariates.
4. **Visualise** the forecast and the per-coefficient ribbon plot.
5. **Diagnose** whether the forecast variance has exploded (near-
   unit-root coefficients).

## Step 1: Fit

Forecasting needs a fit where the dynamic coefficient is actually
*identified* at each period, so we simulate a panel that delivers that:
30 actors over 6 periods, a single dyadic covariate `trade`, and a
directed binary outcome `cooperation`. The effect of `trade` on
cooperation is genuinely time-varying but **mean-reverting** -- it
follows an AR(1) around 0.5 with persistence 0.7, starting high (0.9)
and settling toward its long-run level. The network is moderately dense
(~37%), which is what makes per-period coefficients estimable; this is
the regime in which forecasting is well-posed.

A note on data choice: `dynamic_beta` needs both a reasonable number of
periods and enough ties per period to pin down a coefficient *for that
period*. Very sparse panels (say a rare-event sanctions network at ~2%
density over 3-4 years) do not carry that per-period information, the
AR(1) persistence runs up against a unit root, and the forecast variance
explodes (Step 5 shows exactly that failure mode). A simulated panel
lets us show forecasting working before we show it failing.

```{r fit}
library(lame)
set.seed(2026)

n_fc <- 30
T_fc <- 6
rho_true   <- 0.7
beta_bar   <- 0.5
beta_t_true <- numeric(T_fc)
beta_t_true[1] <- 0.9
for (t in 2:T_fc) {
	beta_t_true[t] <- beta_bar + rho_true * (beta_t_true[t - 1] - beta_bar) +
		rnorm(1, 0, 0.15)
}

Xdyad <- lapply(seq_len(T_fc), function(t) {
	x <- matrix(rnorm(n_fc * n_fc), n_fc, n_fc)
	array(x, dim = c(n_fc, n_fc, 1),
	      dimnames = list(NULL, NULL, "trade"))
})
a_fc <- rnorm(n_fc, 0, 0.4)
b_fc <- rnorm(n_fc, 0, 0.4)
Y <- lapply(seq_len(T_fc), function(t) {
	eta <- -0.5 + beta_t_true[t] * Xdyad[[t]][, , 1] + outer(a_fc, b_fc, "+")
	Yt  <- matrix(rbinom(n_fc * n_fc, 1, pnorm(eta)), n_fc, n_fc)
	diag(Yt) <- NA
	rownames(Yt) <- colnames(Yt) <- sprintf("a%02d", seq_len(n_fc))
	Yt
})
names(Y) <- paste0("t", seq_len(T_fc))

round(beta_t_true, 3)                                   # the truth we recover
sapply(Y, function(y) round(mean(y, na.rm = TRUE), 3))  # per-period density

fit <- lame(
	Y, Xdyad = Xdyad,
	family = "binary", R = 0,
	dynamic_beta = "dyad",      # the trade coefficient is AR(1)
	dynamic_beta_kind = "ar1",  # mean-reverting; switch to "rw1" for drift
	nscan = 800, burn = 200, odens = 5,
	verbose = FALSE
)

# the recovered per-period coefficient tracks the simulated path
round(coef(fit)["trade_dyad", ], 3)
```

The recovered `trade_dyad` path tracks the simulated truth, and (as we
confirm in Step 5) the posterior on $\rho_\beta$ sits comfortably below
1, so the forecast is well-posed.

A quick diagnostic: `summary(fit)` prints the per-block posterior-mean ρ_β
and fires a stationarity warning when, for any block, the 5th
percentile of the posterior on ρ_β is ≥ 0.97 **and** the IQR is < 0.1
(i.e. at least 95% of the mass sits near a unit root with little
spread). The forecast-time warning fires on a complementary trigger:
the upper 97.5% credible bound on ρ_β reaches 0.99. If either warning
fires, refit with `dynamic_beta_kind = "rw1"`. RW1 is unit-root by
construction and is the right prior for permanent-drift coefficients.

## Step 2: Forecast `h` periods ahead

```{r forecast}
set.seed(1)   # predict(h=) propagates the AR(1) state stochastically; seed for reproducibility
# 3-step-ahead forecast on the link scale (linear predictor)
fc_link <- predict(fit, h = 3, type = "link")
length(fc_link)         # 3 (one matrix per future period)
dim(fc_link[[1]])       # n x n  (30 x 30 here; unipartite)

# response-scale forecast applies the family inverse link per draw
fc_resp <- predict(fit, h = 3, type = "response")
# binary: each entry is a posterior-mean predicted probability
range(fc_resp[[1]], na.rm = TRUE)

# by_draw = TRUE returns the full [n, n, h, n_draws] array
fc_full <- predict(fit, h = 3, type = "response", by_draw = TRUE)
dim(fc_full)            # n x n x 3 x n_draws

# interval = "credible" returns a list of length-3 (lower / median / upper)
# matrices per period at the requested quantiles (default 95%)
fc_ci <- predict(fit, h = 3, type = "response", interval = "credible")
str(fc_ci[[1]])         # list($lower, $median, $upper), each n x n
```

The typical cell of `fc_ci[[1]]` carries an informative `lower` /
`median` / `upper` triple -- a genuine sub-interval of [0, 1] rather
than the whole range, because the coefficient path is mean-reverting
(`rho_beta` stays comfortably below 1). A small fraction of cells do
reach the boundary, but the vast majority stay well inside it. The
interval widens with the horizon `h` as forecast uncertainty
accumulates; on a stationary fit it rarely collapses toward the
degenerate [0, 1] you see pervasively near a unit root (Step 5 shows
that failure mode and the warning that catches it).

The forecast propagates `(β_t, a_t, b_t, U_t, V_t)` forward by drawing
from the joint posterior of the dynamic state-space hyperparameters
`(ρ_β, σ_β, ρ_ab, σ_ab, ρ_uv, σ_uv)` for each posterior draw. The same
forecasting machinery works for any family; the binary example above
is just for illustration. For `family = "normal"`, `type = "response"`
returns the forecasted continuous outcome; for `family = "poisson"`,
the response-scale forecast multiplies `exp(η)` by the future-period
exposure (pass via `newexposure = ...`; see also the `period_exposure`
note in `vignettes/dynamic_effects.Rmd`).

### Poisson exposure offsets via `period_exposure`

For `family = "poisson"`, pass `period_exposure = e` to `lame()` (a
non-negative numeric vector of length `T`) when the per-period rate
should be scaled by a known exposure (e.g. weeks observed, population
size, number of edge-formation opportunities). The sampler treats
`log(period_exposure[t])` as a fixed offset on the linear predictor:
the Poisson mean for period `t` becomes
`period_exposure[t] * exp(η_t)` instead of `exp(η_t)`. Setting a
non-trivial (any value not equal to 1) `period_exposure` for a
non-Poisson family is rejected with an error.
When you call `predict(fit, h = K)` later without `newexposure`, the
response-scale forecast reuses the last observed exposure for every
future period; pass `newexposure = rep(e_new, K)` to forecast under a
different exposure path.

## Step 3: Counterfactual forecasts

Pass `newdata = list_of_h_future_X_arrays` to combine alternative
future covariates with the forecast. Each array is `n x n x p` and the
slice order along the 3rd dimension must match the original
`dimnames(Xdyad[[1]])[[3]]` from the fit (here a single slice,
`trade`).

```{r counterfactual}
set.seed(1)   # forecast draws are stochastic; seed so the printed delta summary reproduces
# baseline scenario: freeze the trade covariate at the last observed period
last_X <- Xdyad[[length(Xdyad)]]
X_future <- list(last_X, last_X, last_X)
fc_cf <- predict(fit, h = 3, type = "response", newdata = X_future)

# counterfactual: shift every dyad's trade covariate up by one unit
# going forward. Because `trade` is a standardised (mean-zero) covariate,
# an additive +1 shift -- "one standard deviation more trade for every
# pair" -- is the interpretable counterfactual; a multiplicative scaling
# would be meaningless on a centred covariate (see caution 2 below).
X_future_up <- lapply(X_future, function(x) {
	x[, , 1] <- x[, , 1] + 1
	x
})
fc_up <- predict(fit, h = 3, type = "response", newdata = X_future_up)

# delta is the per-dyad change in posterior-mean cooperation probability
# at h = 3, attributable solely to the +1 trade shift
delta_h3 <- fc_up[[3]] - fc_cf[[3]]
summary(as.vector(delta_h3))
```

The trade coefficient is positive and mean-reverting, so a one-unit
increase in `trade` raises the posterior-mean cooperation probability at
`h = 3` on average (the mean and median of `delta_h3` are both positive).
The per-dyad shift varies in size, and a minority of dyads even show a
small *negative* shift: the three-step-ahead forecast of the coefficient
is itself uncertain (its posterior at `h = 3` is centred well above zero
but has nonzero mass below it), and the marginal probit effect
$\phi(\eta)\,\beta$ is smallest for dyads whose baseline linear predictor
$\eta = a_i + b_j + \beta x_{ij}$ already sits far from the 0.5
probability boundary. For substantive interpretation, summarise
`delta_h3` for the dyads that matter rather than reporting the aggregate
`summary()`.

Two cautions on counterfactual covariates. (1) Because `dynamic_beta`
draws the coefficient path forward from its posterior, the same
covariate change can produce different magnitudes at different horizons,
because the multiplier on `trade` is itself uncertain and drifting. (2)
Multiplicative counterfactuals on covariates that take negative or zero
values (such as this centred `trade` covariate) do not make substantive
sense; prefer additive shifts there, which is why we used `+ 1` rather
than `* 2` above.

## Step 4: Visualise the coefficient path

```{r autoplot, fig.width = 7, fig.height = 4, fig.alt="In-sample posterior coefficient paths from autoplot.lame: each panel shows the posterior median line and a 95 percent credible interval ribbon across the fitted training periods (t1-t6) for one regression coefficient. This plots the estimated in-sample path, not a forecast."}
library(ggplot2)
autoplot(fit, probs = c(0.025, 0.5, 0.975)) +
	labs(title = "Posterior coefficient paths",
	     subtitle = "median + 95% interval",
	     x = "Time period", y = "Coefficient value")
```

The ribbon shows the posterior 95% credible interval around the
posterior median per coefficient and period; faceted by coefficient.

## Step 5: Forecast diagnostics

`predict(fit, h = ...)` fires a one-per-fit warning if the upper
97.5% credible bound on ρ_β reaches 0.99 (i.e. the posterior puts
non-trivial mass near the unit root). On the mean-reverting fit above
that bound sits around 0.98, so **the warning does not fire** and the
forecasts are well-posed. The two regimes still have qualitatively
different forecast-variance behavior, and getting this distinction
right matters for picking your horizon:

- **AR(1), `dynamic_beta_kind = "ar1"`** with $|\rho| < 1$: the
  conditional $h$-step variance of $\beta_{T+h}$ given the training
  window is $\sigma_\beta^2 \cdot (1 - \rho^{2h}) / (1 - \rho^2)$. It
  *saturates* (the increment from $h$ to $h+1$ shrinks geometrically
  to zero) and the limit is the stationary variance
  $\sigma_\beta^2 / (1 - \rho^2)$. That stationary level itself blows
  up as $\rho \to 1$, so AR(1) is still informative at long horizons
  only when $\rho$ is comfortably bounded away from 1.
- **RW1, `dynamic_beta_kind = "rw1"`**: $\rho = 1$ by construction and
  the $h$-step variance is exactly $\sigma_\beta^2 \cdot h$, *linear*
  growth in $h$, never saturating.

In words: AR(1) eventually forgets the training window and falls back
to a stationary distribution; RW1 keeps adding innovation variance
every period. AR(1) forecasts can stay tight for moderate $h$ when
$\rho$ is small; RW1 forecasts get linearly wider every step
regardless. When the AR(1) stationarity warning fires and the
underlying process is genuinely unit-root, refit with
`dynamic_beta_kind = "rw1"`; if the AR(1) posterior on $\rho$ is
near-but-not-at 1 and the science says "should mean-revert eventually,"
stick with AR(1) but cap your reported horizon at $h \approx 3$.

A note on the printed warning text. As of the post-v4 fix, the cli
message branches on `dynamic_beta_kind`: under AR(1) it reports that
the $h$-step forecast variance *saturates* at the stationary level
$\sigma_\beta^2 / (1 - \rho_\beta^2)$ (which itself explodes as
$\rho_\beta \to 1$), and under RW1 it reports that the $h$-step
variance *grows linearly* in $h$. This is the strictly-correct
distinction — AR(1) saturates geometrically, RW1 grows exactly linearly
— and is the version to cite in any methods write-up.

For a longer-horizon diagnostic, use the exact rolling-origin
leave-future-out CV:

```{r lfo}
# refit on Y[1:(t-1)] for each t in periods and score Y[[t]].
# Skip the very first leave-out (would leave a 1-period training window
# which is too short for dynamic_beta); use the last 2 origins.
set.seed(1)  # the internal h=1 forecast draws inside lfo() are stochastic; seed for reproducibility
T_fit <- length(fit$YPM)
lfo_periods <- tail(seq_len(T_fit), max(1L, min(2L, T_fit - 2L)))
lfo_res <- lfo(fit, periods = lfo_periods, refit = TRUE,
               nscan = 200, burn = 50, odens = 5, verbose = FALSE)
print(lfo_res)
```

The `nscan = 200, burn = 50, odens = 5` settings inside `lfo()` are
chosen so the vignette builds in well under a minute on a laptop. They
give `200 / 5 = 40` stored draws per leave-out refit (burn-in is run and discarded, not subtracted from `nscan`), enough to
return a sensible point estimate of per-period elpd for *demonstration*,
but tighter than you should use when the LFO output drives an inference.
For real analyses, pass `nscan = 5000, burn = 1000, odens = 25` (or
whatever matches your `lame()` baseline) and verify the per-period elpd
is stable across two independent runs before relying on it.

On this mean-reverting fit the two leave-out folds return per-period
elpd of roughly -560 to -620 over 870 scored dyads (about -0.65 to -0.7
nats/dyad), and the two folds are close to each other -- the model
forecasts the held-out period about as well at origin 5 as at origin 6,
which is what you want to see when there is no late regime change. If `elpd` instead
dropped sharply at the last fold, that would be the empirical signal of
a regime change near the end of your training window; reach for
`detect_change_point(fit)` (a heuristic Bayes-factor diagnostic) to
localise it.

### Probability-integral-transform (PIT) calibration

`forecast_pit()` complements `lfo()` by asking a sharper question:
*given the held-out outcomes, is the h-step posterior-predictive
distribution itself well calibrated?* For continuous families the PIT
is the Gaussian CDF evaluated at the observation; for binary /
Poisson / ordinal it is the Czado-Gneiting-Held randomised PIT
(Czado, Gneiting & Held 2009). A well-calibrated forecast produces
PIT values that are Uniform(0, 1).

```{r pit}
# fit on the first five periods, hold out the sixth
fit_train <- lame(
	Y[1:5], Xdyad = Xdyad[1:5],
	family = "binary", R = 0,
	dynamic_beta = "dyad",
	nscan = 400, burn = 100, odens = 5,
	verbose = FALSE
)
set.seed(1)   # the randomised PIT draws are stochastic; seed for reproducibility
pit <- forecast_pit(fit_train, y_future = list(Y[[6]]))
print(pit)
```

(`fit_train` is fit on only five periods, so its $\rho_\beta$ posterior
sits closer to the unit root than the six-period `fit` above and the
internal forecast prints the Step-5 unit-root note. That note is about
the *shorter training fit*, not the main model; the calibration result
below is still valid.)

Read the two summaries together. `pit$cover_95` -- the fraction of
held-out cells whose observed outcome falls in the central 95%
posterior-predictive interval, target 0.95 -- is the **stable** summary;
here it lands at about 0.93, close to nominal, so the forecast intervals
have roughly the right width. `pit$ks_p` tests whether the full PIT
distribution is Uniform(0, 1); a value below 0.05 flags
mis-calibration. For a **binary** outcome the PIT is the randomised
Czado-Gneiting-Held version, and on a single held-out period its KS
statistic is noisy -- it hovers near the 0.05 threshold from run to run
(which is why we seed the chunk). Treat `cover_95` as the headline and
`ks_p` as corroborating evidence; do not over-read a single borderline
KS p-value on one held-out period. For a sharper test, hold out two or
more periods and pair `plot(pit)` with `pit$pit` to see whether any
mis-calibration is in the tails (over- vs under-dispersion) or the
centre (location bias). On a sparse, short, near-unit-root panel both
summaries degrade together -- `cover_95` drops well below 0.95 and
`ks_p` collapses toward 0 -- the calibration-side signature of the
forecast-variance explosion the Step 5 warning flags.

## Comparing two fits with `loo_compare()`

When you have two candidate specifications, say a static-`beta`
baseline and a `dynamic_beta = "dyad"` variant, `loo::loo_compare()`
is the one-liner that ranks them on out-of-sample predictive
performance. Both fits need `save_log_lik = TRUE` so `loo()` can read
the stored pointwise log-likelihood matrix off `fit$log_lik`:

```{r loo-compare, eval = FALSE}
fit_static <- lame(
	Y, Xdyad = Xdyad, family = "binary", R = 0,
	nscan = 2000, burn = 500, odens = 5,
	save_log_lik = TRUE,
	verbose = FALSE
)
fit_dyn <- lame(
	Y, Xdyad = Xdyad, family = "binary", R = 0,
	dynamic_beta = "dyad",
	nscan = 2000, burn = 500, odens = 5,
	save_log_lik = TRUE,
	verbose = FALSE
)

# rank the two fits: the row with elpd_diff = 0 is the best,
# subsequent rows show elpd_diff and its SE relative to it. Pass a NAMED
# list so the rows are labelled by model name rather than "model1"/"model2".
cmp <- loo::loo_compare(list(fit_static = loo::loo(fit_static),
                             fit_dyn    = loo::loo(fit_dyn)))
cmp
#>            elpd_diff se_diff
#> fit_dyn     0.0       0.0
#> fit_static -8.4       3.1
```

The `eval = FALSE` block above keeps the vignette build fast; the
printed `elpd_diff` of -8.4 and `se_diff` of 3.1 are illustrative
values you should expect to see on this scale of panel — they are not
literally what running the chunk reproduces. (Passing the bare
`loo()` objects rather than a named list still works but labels the rows
`model1` / `model2`.)

**How to read this output.** The top row is the preferred model
(`fit_dyn` here); `elpd_diff` is always zero for it. Each subsequent
row reports the *difference* in expected log pointwise predictive
density relative to the top model, together with the standard error
of that difference. In the example above, `fit_dyn` beats `fit_static`
by 8.4 elpd units with an SE of 3.1, so `|elpd_diff| / se_diff ~ 2.7`:
the dynamic-beta specification is preferred at roughly the
"2-sigma" threshold. A sociologist writing this up would say
**"the time-varying coefficient model is preferred over the static
specification by about 8 elpd units (SE 3.1), consistent with
nontrivial drift in the effect over the panel window."** If
`|elpd_diff|` were closer to its SE — say `-2.0 (SE 3.0)` — the two
specifications would be predictively indistinguishable and the
simpler static model would be the defensible choice.

A common rule of thumb: if `|elpd_diff|` exceeds roughly
`2 * se_diff`, the better-ranked model is preferred with reasonable
confidence; smaller gaps are within noise and the two specifications
are predictively indistinguishable.

Because `family = "binary"` is one of the six families with an exact
closed-form log-likelihood (`normal`, `binary`, `cbin`, `tobit`,
`poisson`, `ordinal`; see the **log-lik scale** discussion in the
[Dynamic Effects vignette](dynamic_effects.html#loo-waic-via-save_log_lik-true)),
the two `elpd_loo` values here are on the response scale and the
absolute numbers are directly comparable to a `loo()` result from a
probit-link `brms` fit on the same data, modulo the `lame` Gibbs
sampler's R-hat / ESS being satisfactory. For the rank likelihoods
`frn` and `rrl`, the default `observed_exact` falls back to an
augmented-Z normal approximation and is only valid for *relative*
comparison within `lame`; opt in to `log_lik_method = "observed_ghk"`
for the exact marginal on those two families.

### Memory-conscious long runs

If the in-memory log-likelihood matrix from `save_log_lik = TRUE`
would exceed RAM, pass `save_log_lik = "chunked"` instead. The
on-disk chunks are recovered transparently by `loo()`:

```{r chunked, eval = FALSE}
fit_chk <- lame(
	Y, Xdyad = Xdyad, family = "binary", R = 0,
	dynamic_beta = "dyad",
	nscan = 2000, burn = 500, odens = 5,
	save_log_lik = "chunked",          # per-column-chunk binary files
	log_lik_chunk_size = 5000L,
	verbose = FALSE
)
loo::loo(fit_chk)                       # reads the chunks transparently
```

## See also

- `vignettes/dynamic_effects.Rmd`: the full reference on
  `dynamic_beta` / `dynamic_ab` / `dynamic_uv` and the decision tree
  for picking `dynamic_beta_kind`.
- `?predict.lame`: argument reference for `h`, `newdata`, `type`,
  `by_draw`.
- `?autoplot.lame`: ggplot ribbon plot for dynamic coefficients.
- `?lfo`: exact rolling-origin LFO CV.
