---
title: "Getting Started with lame"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>",
	fig.width = 6,
	fig.height = 4,
	dpi = 150,
	out.width = "100%",
	fig.align = "center"
)
```

## What Is lame?

The `lame` package fits **L**ongitudinal **A**dditive and **M**ultiplicative **E**ffects models to network data. If you have a network (who trades with whom, who is friends with whom, who sanctions whom) observed at one or more time points, `lame` lets you estimate covariate effects while properly accounting for the dependencies that make network data special.

The core idea: every actor has a tendency to send ties ($a_i$), receive ties ($b_j$), and a latent position ($u_i$, $v_j$) that captures who connects with whom. Covariates shift the baseline probability of ties. And all of this can optionally evolve over time.

## A 5-Minute Example

Let's fit a model to a small longitudinal binary network in under 20 lines of code.

```{r quick-start, message=FALSE}
library(lame)
set.seed(6886)

# simulate 3 time periods of a 15-node directed network
n <- 15; T_periods <- 3
Y_list <- lapply(1:T_periods, function(t) {
	Y <- matrix(rbinom(n * n, 1, 0.15), n, n)
	diag(Y) <- NA
	rownames(Y) <- colnames(Y) <- paste0("N", 1:n)
	Y
})

# fit a longitudinal AME model
fit <- lame(
	Y = Y_list,
	R = 2,                # 2D latent space
	family = "binary",    # probit for 0/1 networks
	burn = 100,           # burn-in (use 1000+ for real work)
	nscan = 2500,         # post-burn-in samples (use 5000+)
	odens = 25,           # thinning
	verbose = FALSE,
	plot = FALSE
)

summary(fit)
```

That's it. You now have posterior estimates for regression coefficients, variance components, and latent positions pooled across all time periods.

**A note on MCMC settings.** The three key parameters are `burn` (iterations discarded as burn-in), `nscan` (post-burn-in iterations kept for inference), and `odens` (thinning: keep every `odens`-th sample to reduce autocorrelation). With the settings above, we store `nscan / odens` = 2500 / 25 = 100 posterior samples. For a publication, aim for at least 200 stored samples with adequate effective sample sizes.

## What Can You Do with a Fitted Model?

`lame` objects work with all the standard R methods you'd expect:

```{r s3-methods}
# regression coefficients (posterior means)
coef(fit)

# 95% credible intervals
confint(fit)

# predicted probabilities for every dyad at every time point
Y_hat <- predict(fit, type = "response")
cat("Predicted probability range:",
		round(range(unlist(Y_hat), na.rm = TRUE), 3), "\n")

# residuals
resid_list <- residuals(fit)
cat("Residual SD:", round(sd(unlist(resid_list), na.rm = TRUE), 3), "\n")
```

The intercept-only model estimates a single coefficient. Since we simulated ties with probability 0.15, the intercept should be near `qnorm(0.15)` $\approx$ -1.04 on the probit scale. The predicted probabilities range from near zero (for dyads where both the sender and receiver random effects are strongly negative) to higher values (where the random effects reinforce a tie).

## Checking Your Model

Two essential diagnostics:

**1. Did the MCMC converge?** The trace plots should look like fuzzy caterpillars, not random walks with trends.

```{r trace-plot, fig.height=5}
trace_plot(fit, params = "beta")
```

With an intercept-only model, there is just one regression parameter to check. The trace should show stable mixing around a value near `r round(qnorm(0.15), 2)` (since `qnorm(0.15)` corresponds to the 15% density we simulated).

**2. Does the model fit the data?** The GOF plots compare observed network statistics to what the model predicts. Observed values should fall within the simulated distributions.

```{r gof-plot, fig.width=8, fig.height=5}
gof_plot(fit)
```

If the observed statistics (points or vertical lines) fall within the posterior predictive bands or histograms, the model is reproducing the network's structural features.

## Visualizing Network Structure

The latent space shows which actors play similar roles in the network:

```{r uv-plot, fig.width=7, fig.height=6}
uv_plot(fit)
```

And the additive effects show who's unusually active or popular:

```{r ab-plot, fig.height=4}
ab_plot(fit, effect = "sender")
```

## Cross-Sectional Models

If you have a single network (not a time series), use `ame()` instead of `lame()`:

```{r cross-sectional, message=FALSE}
fit_cs <- ame(
	Y = Y_list[[1]],     # just one time period
	R = 2,
	family = "binary",
	burn = 100,
	nscan = 2500,
	odens = 25,
	verbose = FALSE
)

coef(fit_cs)
```

The output and methods are the same. The difference is that `lame()` pools information across time periods, giving you more precise estimates when the structure is stable.

## Dynamic Effects

When you believe the network structure is changing over time (alliances shifting, friendships evolving), you can let the latent positions and additive effects drift via AR(1) processes:

```{r dynamic, message=FALSE}
fit_dyn <- lame(
	Y = Y_list,
	R = 2,
	dynamic_ab = TRUE,    # time-varying sociality/popularity
	dynamic_uv = TRUE,    # time-varying latent positions
	family = "binary",
	burn = 100,
	nscan = 2500,
	odens = 25,
	verbose = FALSE,
	plot = FALSE
)

summary(fit_dyn)
```

The model estimates how persistent the latent positions are ($\rho_{uv}$) and how persistent the additive effects are ($\rho_{ab}$). Values near 1 mean slow evolution; values near 0 mean rapid change. Since we simulated independent networks, the data carry no temporal signal. The estimated persistence parameters reflect the prior rather than genuine dynamics. With only 3 time periods, there is little information to update the prior. See the [dynamic effects vignette](dynamic_effects.html) for a more thorough treatment with 5 time periods and discussion of when dynamic effects add value.

## Bipartite Networks

For two-mode networks (students and courses, countries and treaties), pass a rectangular matrix and set `mode = "bipartite"`:

```{r bipartite, message=FALSE}
nA <- 10; nB <- 8
Y_bip <- lapply(1:3, function(t) {
	Y <- matrix(rbinom(nA * nB, 1, 0.2), nA, nB)
	rownames(Y) <- paste0("R", 1:nA)
	colnames(Y) <- paste0("C", 1:nB)
	Y
})

fit_bip <- lame(
	Y = Y_bip,
	mode = "bipartite",
	R = 2,
	family = "binary",
	burn = 100, nscan = 2500, odens = 25,
	verbose = FALSE, plot = FALSE
)

summary(fit_bip)
```

Notice that the bipartite summary reports `cab = 0.000` and `rho = 0.000` — these parameters (sender-receiver covariance and dyadic correlation) are not estimated in bipartite models because the two node types are structurally different. See the [bipartite vignette](bipartite.html) for a full walkthrough.

## Supported Data Types

| Family | Data | Example |
|--------|------|---------|
| `"normal"` | Continuous | Trade volumes, survey ratings |
| `"binary"` | 0/1 | Friendships, alliances, sanctions |
| `"ordinal"` | Ordered categories | Conflict intensity (none/threat/action) |
| `"poisson"` | Counts | Number of co-sponsored bills |
| `"tobit"` | Non-negative continuous | Trade with zeros |
| `"cbin"` | Censored binary | Friendships with nomination limits |
| `"frn"` | Fixed rank nomination | "Name your top 5 friends" |
| `"rrl"` | Row-ranked likelihood | Ranked preferences |

## Where to Go Next

| I want to... | Read this |
|--------------|-----------|
| Understand the full workflow with real data | [lame overview](lame-overview.html) |
| Learn about cross-sectional models in depth | [Your first AME model](cross_sec_ame.html) |
| Model two-mode networks | [Bipartite networks](bipartite.html) |
| Let the network structure evolve over time | [Dynamic effects](dynamic_effects.html) |

## References

Hoff, PD (2021). Additive and Multiplicative Effects Network Models. *Statistical Science*, 36, 34--50.

Minhas, S., Dorff, C., Gallop, M. B., Foster, M., Liu, H., Tellez, J., & Ward, M. D. (2022). Taking dyads seriously. *Political Science Research and Methods*, 10(4), 703--721.
