---
title: "Getting Started with lame"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 150,
  out.width = "100%",
  fig.align = "center"
)
```

## What Is lame?

The `lame` package fits **L**ongitudinal **A**dditive and **M**ultiplicative **E**ffects models to network data. If you have a network (who trades with whom, who is friends with whom, who sanctions whom) observed at one or more time points, `lame` lets you estimate covariate effects while properly accounting for the dependencies that make network data special.

The core idea: every actor has a tendency to send ties ($a_i$), receive ties ($b_j$), and a latent position ($u_i$, $v_j$) that captures who connects with whom. Covariates shift the baseline probability of ties. And all of this can optionally evolve over time.

## A 5-Minute Example

Let's fit a model to a small longitudinal binary network in under 20 lines of code.

```{r quick-start, message=FALSE}
library(lame)
set.seed(6886)

# Simulate 3 time periods of a 15-node directed network
n <- 15; T_periods <- 3
Y_list <- lapply(1:T_periods, function(t) {
  Y <- matrix(rbinom(n * n, 1, 0.15), n, n)
  diag(Y) <- NA
  rownames(Y) <- colnames(Y) <- paste0("N", 1:n)
  Y
})

# Fit a longitudinal AME model
fit <- lame(
  Y = Y_list,
  R = 2,                # 2D latent space
  family = "binary",    # probit for 0/1 networks
  burn = 100,           # burn-in (use 1000+ for real work)
  nscan = 500,          # post-burn-in samples (use 5000+)
  odens = 25,           # thinning
  verbose = FALSE,
  plot = FALSE
)

summary(fit)
```

That's it. You now have posterior estimates for regression coefficients, variance components, and latent positions pooled across all time periods.

## What Can You Do with a Fitted Model?

`lame` objects work with all the standard R methods you'd expect:

```{r s3-methods}
# Regression coefficients (posterior means)
coef(fit)

# 95% credible intervals
confint(fit)

# Predicted probabilities for every dyad at every time point
Y_hat <- predict(fit, type = "response")
cat("Predicted probability range:",
    round(range(unlist(Y_hat), na.rm = TRUE), 3), "\n")

# Residuals
resid_list <- residuals(fit)
cat("Residual SD:", round(sd(unlist(resid_list), na.rm = TRUE), 3), "\n")
```

## Checking Your Model

Two essential diagnostics:

**1. Did the MCMC converge?** The trace plots should look like fuzzy caterpillars, not random walks with trends.

```{r trace-plot, fig.height=5}
trace_plot(fit, params = "beta")
```

**2. Does the model fit the data?** The GOF plots compare observed network statistics to what the model predicts. Observed values should fall within the simulated distributions.

```{r gof-plot, fig.width=8, fig.height=5}
gof_plot(fit)
```

## Visualizing Network Structure

The latent space shows which actors play similar roles in the network:

```{r uv-plot, fig.width=7, fig.height=6}
uv_plot(fit)
```

And the additive effects show who's unusually active or popular:

```{r ab-plot, fig.height=4}
ab_plot(fit, effect = "sender")
```

## Cross-Sectional Models

If you have a single network (not a time series), use `ame()` instead of `lame()`:

```{r cross-sectional, message=FALSE}
fit_cs <- ame(
  Y = Y_list[[1]],     # just one time period
  R = 2,
  family = "binary",
  burn = 100,
  nscan = 500,
  odens = 25,
  verbose = FALSE
)

coef(fit_cs)
```

The output and methods are the same. The difference is that `lame()` pools information across time periods, giving you more precise estimates when the structure is stable.

## Dynamic Effects

When you believe the network structure is changing over time (alliances shifting, friendships evolving), you can let the latent positions and additive effects drift via AR(1) processes:

```{r dynamic, message=FALSE}
fit_dyn <- lame(
  Y = Y_list,
  R = 2,
  dynamic_ab = TRUE,    # time-varying sociality/popularity
  dynamic_uv = TRUE,    # time-varying latent positions
  family = "binary",
  burn = 100,
  nscan = 500,
  odens = 25,
  verbose = FALSE,
  plot = FALSE
)

summary(fit_dyn)
```

The model estimates how persistent the latent positions are ($\rho_{uv}$): values near 1 mean slow evolution, values near 0 mean rapid change. See the [dynamic effects vignette](dynamic_effects.html) for a deeper dive.

## Bipartite Networks

For two-mode networks (students and courses, countries and treaties), pass a rectangular matrix and set `mode = "bipartite"`:

```{r bipartite, message=FALSE}
nA <- 10; nB <- 8
Y_bip <- lapply(1:3, function(t) {
  Y <- matrix(rbinom(nA * nB, 1, 0.2), nA, nB)
  rownames(Y) <- paste0("R", 1:nA)
  colnames(Y) <- paste0("C", 1:nB)
  Y
})

fit_bip <- lame(
  Y = Y_bip,
  mode = "bipartite",
  R = 2,
  family = "binary",
  burn = 100, nscan = 500, odens = 25,
  verbose = FALSE, plot = FALSE
)

summary(fit_bip)
```

See the [bipartite vignette](bipartite.html) for a full walkthrough.

## Supported Data Types

| Family | Data | Example |
|--------|------|---------|
| `"normal"` | Continuous | Trade volumes, survey ratings |
| `"binary"` | 0/1 | Friendships, alliances, sanctions |
| `"ordinal"` | Ordered categories | Conflict intensity (none/threat/action) |
| `"poisson"` | Counts | Number of co-sponsored bills |
| `"tobit"` | Non-negative continuous | Trade with zeros |
| `"cbin"` | Censored binary | Friendships with nomination limits |
| `"frn"` | Fixed rank nomination | "Name your top 5 friends" |
| `"rrl"` | Row-ranked likelihood | Ranked preferences |

## Where to Go Next

| I want to... | Read this |
|--------------|-----------|
| Understand the full workflow with real data | [lame overview](lame-overview.html) |
| Learn about cross-sectional models in depth | [Your first AME model](cross_sec_ame.html) |
| Model two-mode networks | [Bipartite networks](bipartite.html) |
| Let the network structure evolve over time | [Dynamic effects](dynamic_effects.html) |

## References

Hoff, PD (2021). Additive and Multiplicative Effects Network Models. *Statistical Science*, 36, 34--50.

Minhas, S., Dorff, C., Gallop, M. B., Foster, M., Liu, H., Tellez, J., & Ward, M. D. (2022). Taking dyads seriously. *Political Science Research and Methods*, 10(4), 703--721.
