---
title: "lame Overview"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lame Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  fig.width = 6,
  fig.height = 4,
  dpi = 200,
  out.width = "100%",
  fig.align = "center",
  dev = "png"
)
```

## Package Overview

The `lame` package provides tools for fitting **L**ongitudinal **A**dditive and **M**ultiplicative **E**ffects models to network data observed over time. If you study relationships between actors (countries trading with each other, legislators co-sponsoring bills, students forming friendships across semesters) and you have repeated observations of those relationships, `lame` is designed for you.

The modeling approach builds on the Additive and Multiplicative Effects (AME) framework developed by Peter Hoff, whose [`amen`](https://pdhoff.github.io/amen/) package ([GitHub](https://github.com/pdhoff/amen)) provides the foundational implementation for cross-sectional network analysis. The `lame` package extends this framework in several directions:

- **Longitudinal and replicated data**: While `amen` focuses on single networks (with `ame()`) or basic replicated designs (with `ame_rep()`), `lame` is built from the ground up for panel data, networks observed across multiple time periods, potentially with actors entering and leaving the sample.
- **Dynamic effects**: Actors' positions in latent space and their baseline activity levels can evolve over time via AR(1) processes (`dynamic_uv` and `dynamic_ab`), letting you model how network structure changes rather than assuming it is fixed.
- **C++ acceleration**: Core sampling routines are implemented in Rcpp/RcppArmadillo for faster computation, which matters when you have many time periods or actors.
- **Bipartite networks**: Explicit support for two-mode networks (e.g., countries and treaties, users and products) with separate latent spaces for row and column nodes.
- **ggplot2-based diagnostics**: Built-in visualization functions (`trace_plot`, `gof_plot`, `ab_plot`, `uv_plot`) produce publication-ready ggplot2 figures.
- **Standard S3 methods**: `coef()`, `confint()`, `fitted()`, `residuals()`, `predict()`, and `simulate()` work as you would expect from any R model object.

The package is part of the `netify-verse`, so it works seamlessly with the [`netify`](https://netify-dev.github.io/netify/) package for data preparation.

## Application: International Sanctions

To see how `lame` works in practice, we analyze data from the Threat and Imposition of Sanctions (TIES) dataset (Morgan et al. 2014). The data cover 35 countries across four years (1993--1995 and 2000). The core question is straightforward: *which country-level and dyad-level factors predict whether one country sanctions another, and are there latent patterns of sanctioning behavior beyond what these covariates explain?*

The dataset includes:

- **Y**: A list of directed binary adjacency matrices (one per year). Entry $y_{ij,t} = 1$ means country $i$ imposed sanctions on country $j$ in year $t$.
- **Xdyad**: Dyadic covariates (geographic distance and the number of shared intergovernmental organizations (IGOs)) that capture relationship-level predictors.
- **Xrow / Xcol**: Sender and receiver covariates (GDP and population) capturing country-level characteristics. We use the same variables for both senders and receivers here, but different covariates can be specified for each role.

```{r setup, include=FALSE}
library(lame)
library(ggplot2)

# Load the TIES data
data("vignette_data")

```

### Fitting the Model

We fit a binary probit AME model with sender random effects (`rvar`), receiver random effects (`cvar`), dyadic correlation (`dcor`), and a 2-dimensional latent space (`R = 2`). The random effects capture the fact that some countries are simply more active sanctioners (sender effects) or more frequent targets (receiver effects), while the latent space picks up residual patterns, groups of countries that sanction similar targets or are targeted by the same senders.

```{r, message=FALSE, warning=FALSE}
# Fit the AME model for binary data
# Note: burn and nscan are kept small for fast vignette building.
# For real analyses, use burn >= 1000 and nscan >= 5000.
fit <- lame(
  Y = Y,
  Xdyad = Xdyad,           # dyadic covariates
  Xrow = Xrow,             # sender covariates
  Xcol = Xcol,             # receiver covariates
  family = "binary",       # Binary probit model
  rvar = TRUE,             # sender random effects
  cvar = TRUE,             # receiver random effects
  dcor = TRUE,             # Dyadic correlation
  R = 2,                   # Multiplicative effects dimension
  symmetric = FALSE,       # Directed network
  burn = 100,              # Burn-in iterations
  nscan = 500,             # Post-burn-in iterations
  odens = 25,              # Output density (thinning)
  verbose = FALSE,           # Suppress iteration output
  plot = FALSE             # Suppress real-time plots
)

```

### Interpreting the Results

```{r}
summary(fit)
```

The regression coefficients tell us about the structural drivers of sanctions. Shared IGO membership is positively associated with sanctions, as countries embedded in the same international institutions are more likely to sanction each other, probably because these institutions provide both the information and the institutional mechanisms to coordinate economic pressure. Distance has a negative coefficient, meaning geographically closer countries are more likely to sanction one another.

The variance components at the bottom of the summary capture network-level heterogeneity. Sender variance (`va`) and receiver variance (`vb`) indicate how much countries differ in their baseline propensity to impose or receive sanctions. The dyadic correlation (`rho`) captures reciprocity, the tendency for sanctions to be mutual.

### Checking Convergence

Because `lame` uses MCMC, we need to verify that the sampler has converged before trusting the results. The `trace_plot` function shows the sampled values over iterations (left panels) and the corresponding posterior densities (right panels) for each parameter.

```{r, fig.width=8, fig.height=8, dpi=100, dev="png", fig.alt="MCMC trace plots and density plots showing convergence and posterior distributions of regression coefficients"}
trace_plot(fit, params = "beta")
```

What to look for: the trace plots should look like "fuzzy caterpillars," bouncing around a stable mean without long-term trends or getting stuck in one region. The density plots should be smooth and unimodal. With only `r nrow(fit$BETA)` post-burn-in samples (due to the small `nscan` we used here), the chains have not fully converged. In a real analysis, you would run much longer chains and check that increasing the run length doesn't substantially change the posterior summaries.

### Goodness of Fit

The `gof_plot` function compares observed network statistics to their posterior predictive distributions. This tells us whether the model can reproduce key structural features of the data, not just the dyad-level relationships, but emergent properties like how much actors vary in their degree (heterogeneity), whether friends-of-friends tend to be friends (transitivity), and whether ties tend to be reciprocated (dyadic dependence).

```{r, fig.width=8, fig.height=6, dpi=100, dev="png", fig.alt="Goodness of fit plots comparing observed network statistics to posterior predictive distributions across time periods"}
gof_plot(fit)
```

For a longitudinal model, `gof_plot` shows these statistics across time. The observed values (points) are overlaid on the posterior predictive intervals (shaded bands). When the observed values fall inside the bands, the model is capturing that aspect of the network well. When they fall outside, it signals that the model is missing something. For instance, if the model consistently under-predicts transitivity, it might suggest a higher-dimensional latent space is needed.

### Latent Space

The multiplicative effects capture patterns of association that go beyond what the covariates and additive effects explain. The `uv_plot` function visualizes these latent positions. Countries positioned near each other in the sender space (triangles) tend to sanction similar targets; countries near each other in the receiver space (circles) tend to be sanctioned by the same senders. Countries placed on opposite sides of the plot tend to have dissimilar sanctioning patterns.

```{r, fig.width=8, fig.height=6, dpi=100, dev="png", fig.alt="Network visualization showing multiplicative effects latent space positions of countries in the sanctions network"}

# Network plot showing multiplicative effects
uv_plot(fit) +
  ggtitle("Sanctions Network - Multiplicative Effects") +
  theme_minimal() +
  theme(
    legend.position='none',
    axis.text=element_blank(),
    axis.title=element_blank()
  )

```

The circular layout places sender positions (triangles) on the outer ring and receiver positions (circles) on an inner ring, scaled by magnitude. Countries with larger effects are placed further from the center, indicating they play a more distinctive role in the network. Countries clustered together (for example, Western allies like the USA and Canada) share similar sanctioning profiles.


## Extracting Latent Positions

If you need the estimated latent positions in a tidy format (for custom plots, merging with external data, or exporting to other tools), the `latent_positions()` function returns a data frame:

```{r}
lp <- latent_positions(fit)
head(lp)
```

Each row gives one actor's position on one latent dimension at one time point. For dynamic models (`dynamic_uv = TRUE`), the `time` column tracks how positions evolve. You can optionally apply Procrustes alignment to remove rotational indeterminacy across time periods:

```{r}
aligned <- procrustes_align(fit)
str(aligned$U)  # aligned 3D array
```

## What's Next?

This vignette covered the core workflow: fitting, convergence, GOF, and visualization. For more specialized topics:

- **Single networks (no time series)?** See [Your First AME Model](cross_sec_ame.html) for a detailed cross-sectional walkthrough
- **Two types of nodes?** See [Bipartite Networks](bipartite.html)
- **Evolving network structure?** See [Dynamic Effects](dynamic_effects.html)
- **Just want the quick version?** See [Getting Started](lame.html)

**References**:

Hoff, PD (2021) Additive and Multiplicative Effects Network Models. Statistical Science 36, 34--50.

Morgan, T. Clifton, Bapat, N., & Kobayashi, Y. (2014). Threat and Imposition of Economic Sanctions 1945-2005. Conflict Management and Peace Science 31(5): 541-558.

Minhas, S., Dorff, C., Gallop, M. B., Foster, M., Liu, H., Tellez, J., & Ward, M. D. (2022). Taking dyads seriously. Political Science Research and Methods, 10(4), 703--721. doi:10.1017/psrm.2021.56
