---
title: "Introduction to LAME"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to LAME}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 150,
  out.width = "100%",
  fig.align = "center"
)
```

## Overview

The `lame` package implements Additive and Multiplicative Effects (AME)
models for both cross-sectional and longitudinal network data. It extends
the AME framework (Hoff 2021) to handle temporal dependencies through
dynamic latent factors and additive effects, with a C++ backend for
efficient computation.

### Key Features

- **Cross-sectional and longitudinal** network analysis via `ame()` and `lame()`
- **Dynamic effects**: time-varying additive and multiplicative effects with AR(1) processes
- **Bipartite networks**: two-mode networks with separate row and column latent spaces
- **Multiple families**: normal, binary, ordinal, poisson, tobit, censored binary,
  fixed rank nomination, and row-ranked likelihood
- **S3 methods**: `coef()`, `vcov()`, `confint()`, `residuals()`, `predict()`,
  `summary()`, `print()`

## Quick Start

```{r setup, message=FALSE}
library(lame)
set.seed(6886)
```

### Simulating Network Data

We begin by generating a small directed binary network observed over 3 time periods.

```{r simulate-data}
n <- 15
T_periods <- 3

Y_list <- lapply(1:T_periods, function(t) {
  Y <- matrix(rbinom(n * n, 1, 0.15), n, n)
  diag(Y) <- NA
  rownames(Y) <- colnames(Y) <- paste0("N", 1:n)
  Y
})

cat("Network dimensions:", n, "x", n, "\n")
cat("Time periods:", T_periods, "\n")
cat("Densities:", round(sapply(Y_list, mean, na.rm = TRUE), 3), "\n")
```

### Fitting a Longitudinal Model

The `lame()` function fits the longitudinal AME model using MCMC:

```{r fit-model, message=FALSE, warning=FALSE}
fit <- lame(
  Y = Y_list,
  R = 2,               # 2-dimensional latent space
  family = "binary",   # binary (probit) networks
  rvar = TRUE,         # sender random effects
  cvar = TRUE,         # receiver random effects
  burn = 100,          # burn-in (use 1000+ in practice)
  nscan = 500,         # post-burn-in iterations (use 5000+ in practice)
  odens = 25,          # thinning interval
  print = FALSE,
  plot = FALSE
)
```

### Inspecting the Fit

```{r summary}
summary(fit)
```

## S3 Methods

The package provides standard R methods for extracting results from fitted models.

### Regression Coefficients

```{r coef-methods}
# posterior means
coef(fit)

# posterior covariance matrix of regression coefficients
vcov(fit)

# 95% Bayesian credible intervals (not confidence intervals)
confint(fit)

# narrower 90% intervals
confint(fit, level = 0.90)
```

### Predictions and Residuals

```{r predict-residuals}
# predicted probabilities (response scale) -- returns list for lame objects
Y_hat <- predict(fit, type = "response")
cat("Predicted probability range:", round(range(unlist(Y_hat), na.rm = TRUE), 3), "\n")

# residuals (observed - fitted) -- returns list for lame objects
resid_list <- residuals(fit)
cat("Residual SD:", round(sd(unlist(resid_list), na.rm = TRUE), 3), "\n")
```

## Cross-Sectional Analysis

For a single network observed at one time point, use `ame()`:

```{r cross-sectional, message=FALSE, warning=FALSE}
fit_cs <- ame(
  Y = Y_list[[1]],
  R = 2,
  family = "binary",
  burn = 100,
  nscan = 500,
  odens = 25,
  print = FALSE
)

summary(fit_cs)
coef(fit_cs)
```

## Dynamic Effects

The `lame()` function supports time-varying sender/receiver effects and
latent positions via AR(1) processes.

```{r dynamic, message=FALSE, warning=FALSE}
fit_dyn <- lame(
  Y = Y_list,
  R = 2,
  dynamic_ab = TRUE,   # time-varying sender/receiver effects
  dynamic_uv = TRUE,   # time-varying latent positions

  family = "binary",
  burn = 100,
  nscan = 500,
  odens = 25,
  print = FALSE,
  plot = FALSE
)

summary(fit_dyn)
```

## Visualizing Results

### MCMC Diagnostics

Trace plots help assess convergence of the MCMC sampler:

```{r trace-plot, fig.height=5}
trace_plot(fit, params = "beta")
```

### Goodness of Fit

GOF plots compare observed network statistics to their posterior
predictive distributions:

```{r gof-plot, fig.width=8, fig.height=5}
gof_plot(fit)
```

### Latent Space

The `uv_plot()` function visualizes the estimated latent positions:

```{r uv-plot, fig.width=7, fig.height=6}
uv_plot(fit)
```

### Additive Effects

The `ab_plot()` function shows sender and receiver random effects:

```{r ab-plot, fig.height=4}
ab_plot(fit, effect = "sender")
```

## Model Families

The package supports 8 distributional families:

| Family | Description | Use case |
|--------|-------------|----------|
| `"normal"` | Gaussian | Continuous valued networks |
| `"binary"` | Probit | Binary networks (0/1) |
| `"ordinal"` | Ordinal probit | Ordered categorical ties |
| `"poisson"` | Poisson | Count networks |
| `"tobit"` | Censored normal | Non-negative continuous |
| `"cbin"` | Censored binary | Binary with censoring |
| `"frn"` | Fixed rank nomination | Fixed number of nominations |
| `"rrl"` | Row-ranked likelihood | Ranked preferences |

## Bipartite Networks

For two-mode networks with different row and column node sets, set
`mode = "bipartite"`:

```{r bipartite, message=FALSE, warning=FALSE}
# simulate a small bipartite network
nA <- 10; nB <- 8
Y_bip <- lapply(1:3, function(t) {
  Y <- matrix(rbinom(nA * nB, 1, 0.2), nA, nB)
  rownames(Y) <- paste0("R", 1:nA)
  colnames(Y) <- paste0("C", 1:nB)
  Y
})

fit_bip <- lame(
  Y = Y_bip,
  mode = "bipartite",
  R = 2,
  family = "binary",
  burn = 100,
  nscan = 500,
  odens = 25,
  print = FALSE,
  plot = FALSE
)

summary(fit_bip)
```

## References

Hoff, PD (2021). Additive and Multiplicative Effects Network Models.
*Statistical Science*, 36, 34--50.

Minhas, S., Dorff, C., Gallop, M. B., Foster, M., Liu, H., Tellez, J.,
& Ward, M. D. (2022). Taking dyads seriously. *Political Science
Research and Methods*, 10(4), 703--721.
