---
title: "lame Overview"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lame Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "",
	fig.width = 6,
	fig.height = 4,
	dpi = 200,
	out.width = "100%",
	fig.align = "center",
	dev = "png"
)
```

## Package Overview

The `lame` package provides tools for fitting **L**ongitudinal **A**dditive and **M**ultiplicative **E**ffects models to network data observed over time. If you study relationships between actors (countries trading with each other, legislators co-sponsoring bills, students forming friendships across semesters) and you have repeated observations of those relationships, `lame` is designed for you.

The modeling approach builds on the Additive and Multiplicative Effects (AME) framework developed by Peter Hoff, whose [`amen`](https://pdhoff.github.io/amen/) package ([GitHub](https://github.com/pdhoff/amen)) provides the foundational implementation for cross-sectional network analysis. The `lame` package extends this framework in several directions:

- **Longitudinal and replicated data**: While `amen` focuses on single networks (with `ame()`) or basic replicated designs (with `ame_rep()`), `lame` is built from the ground up for panel data, networks observed across multiple time periods, potentially with actors entering and leaving the sample.
- **Dynamic effects**: Actors' positions in latent space and their baseline activity levels can evolve over time via AR(1) processes (`dynamic_uv` and `dynamic_ab`), letting you model how network structure changes rather than assuming it is fixed.
- **C++ acceleration**: Core sampling routines are implemented in Rcpp/RcppArmadillo for faster computation, which matters when you have many time periods or actors.
- **Bipartite networks**: Explicit support for two-mode networks (e.g., countries and treaties, users and products) with separate latent spaces for row and column nodes.
- **ggplot2-based diagnostics**: Built-in visualization functions (`trace_plot`, `gof_plot`, `ab_plot`, `uv_plot`) produce publication-ready ggplot2 figures.
- **Standard S3 methods**: `coef()`, `confint()`, `fitted()`, `residuals()`, `predict()`, and `simulate()` work as you would expect from any R model object.

The package is part of the `netify-verse`, so it works seamlessly with the [`netify`](https://netify-dev.github.io/netify/) package for data preparation.

## Application: International Sanctions

To see how `lame` works in practice, we analyze data from the Threat and Imposition of Sanctions (TIES) dataset (Morgan et al. 2014). The data cover 35 countries across four years (1993--1995 and 2000). The core question is straightforward: *which country-level and dyad-level factors predict whether one country sanctions another, and are there latent patterns of sanctioning behavior beyond what these covariates explain?*

The dataset includes:

- **Y**: A list of directed binary adjacency matrices (one per year). Entry $y_{ij,t} = 1$ means country $i$ imposed sanctions on country $j$ in year $t$.
- **Xdyad**: Dyadic covariates (geographic distance and the number of shared intergovernmental organizations (IGOs)) that capture relationship-level predictors.
- **Xrow / Xcol**: Sender and receiver covariates (log GDP and log population) capturing country-level characteristics. We use the same variables for both senders and receivers here, but different covariates can be specified for each role.

```{r setup}
library(lame)
library(ggplot2)
set.seed(6886)

# load the TIES data
data("vignette_data")
```

### Fitting the Model

We fit a binary probit AME model with sender random effects (`rvar`), receiver random effects (`cvar`), dyadic correlation (`dcor`), and a 2-dimensional latent space (`R = 2`). The random effects capture the fact that some countries are simply more active sanctioners (sender effects) or more frequent targets (receiver effects), while the latent space picks up residual patterns, groups of countries that sanction similar targets or are targeted by the same senders.

```{r, message=FALSE, warning=FALSE}
# fit the AME model for binary data
# note: for real analyses, use burn >= 2000 and nscan >= 10000.
# sparse networks like this one require long chains.
fit <- lame(
	Y = Y,
	Xdyad = Xdyad,           # dyadic covariates
	Xrow = Xrow,             # sender covariates
	Xcol = Xcol,             # receiver covariates
	family = "binary",       # Binary probit model
	rvar = TRUE,             # sender random effects
	cvar = TRUE,             # receiver random effects
	dcor = TRUE,             # Dyadic correlation
	R = 2,                   # Multiplicative effects dimension
	symmetric = FALSE,       # Directed network
	burn = 500,              # Burn-in iterations
	nscan = 2500,            # Post-burn-in iterations
	odens = 5,               # Thinning (keep every 5th sample)
	verbose = FALSE,         # Suppress iteration output
	plot = FALSE             # Suppress real-time plots
)

```

### Interpreting the Results

```{r}
summary(fit)
```

The coefficients table reports posterior means, standard errors, and 95% credible intervals for each covariate. With a binary probit model, the coefficients are on the latent scale: positive values increase the probability of a sanction, negative values decrease it.

The variance components reveal where heterogeneity lies. The sender variance (`va`) is much larger than the receiver variance (`vb`), meaning the main source of heterogeneity is in how actively countries impose sanctions, not in who gets targeted. A few countries are prolific sanctioners while most rarely sanction at all. The dyadic correlation (`rho`) captures reciprocity: high values indicate that if country A sanctions country B, B is more likely to sanction A.

The intercept is large and negative, reflecting the sparsity of the network (density around 1--2%). On the probability scale, this corresponds to a near-zero baseline probability of a sanction. The sender random effects and latent positions then shift individual dyads away from this baseline.

### Checking Convergence

Because the model is estimated via MCMC, we need to verify that the sampler has converged before trusting the results. The `trace_plot` function shows the sampled values over iterations (top panels) and the corresponding posterior densities (bottom panels) for each parameter.

```{r, fig.width=8, fig.height=8, dpi=100, dev="png", fig.alt="MCMC trace plots and density plots showing convergence and posterior distributions of regression coefficients"}
trace_plot(fit, params = "beta")
```

What to look for: the trace plots (top) should look like "fuzzy caterpillars," bouncing around a stable mean without long-term trends or getting stuck in one region. The density plots (bottom) should be smooth and unimodal.

With `r nrow(fit$BETA)` stored samples, you may notice slow mixing in some parameters, particularly the intercept and row covariates. This is typical for very sparse binary networks: with only about 20 ties per year, the likelihood surface is flat and the sampler moves slowly. For a publication, you would run much longer chains (`burn >= 2000, nscan >= 10000`) and verify that effective sample sizes (available via `coda::effectiveSize()`) are adequate for all parameters.

### Goodness of Fit

The `gof_plot` function compares observed network statistics to their posterior predictive distributions. This tells us whether the model can reproduce key structural features of the data, not just the dyad-level relationships, but emergent properties like how much actors vary in their degree (heterogeneity), whether friends-of-friends tend to be friends (transitivity), and whether ties tend to be reciprocated (dyadic dependence).

```{r, fig.width=8, fig.height=6, dpi=100, dev="png", fig.alt="Goodness of fit plots comparing observed network statistics to posterior predictive distributions across time periods"}
gof_plot(fit)
```

For a longitudinal model, `gof_plot` shows these statistics across time. The observed values (points) are overlaid on the posterior predictive intervals (shaded bands). When the observed values fall inside the bands, the model is capturing that aspect of the network well. When they fall outside, it signals that the model is missing something. For instance, if the model consistently under-predicts transitivity, it might suggest a higher-dimensional latent space is needed.

### Latent Space

The multiplicative effects capture patterns of association that go beyond what the covariates and additive effects explain. The `uv_plot` function visualizes these latent positions. Countries positioned near each other in the sender space (triangles) tend to sanction similar targets; countries near each other in the receiver space (circles) tend to be sanctioned by the same senders. Countries placed on opposite sides of the plot tend to have dissimilar sanctioning patterns.

```{r, fig.width=8, fig.height=6, dpi=100, dev="png", fig.alt="Network visualization showing multiplicative effects latent space positions of countries in the sanctions network"}

# network plot showing multiplicative effects
uv_plot(fit) +
	ggtitle("Sanctions Network - Multiplicative Effects") +
	theme_bw() +
	theme(
		panel.border = element_blank(),
		legend.position = 'none',
		axis.text = element_blank(),
		axis.title = element_blank(),
		panel.grid = element_blank()
	)

```

The circular layout places sender positions (triangles) on the outer ring and receiver positions (circles) on an inner ring, scaled by magnitude. Countries with larger effects are placed further from the center, indicating they play a more distinctive role in the network. Countries clustered together share similar sanctioning profiles, while countries on opposite sides have dissimilar patterns.


## Extracting Latent Positions

If you need the estimated latent positions in a tidy format (for custom plots, merging with external data, or exporting to other tools), the `latent_positions()` function returns a data frame:

```{r}
lp <- latent_positions(fit)
head(lp)
```

Each row gives one actor's position on one latent dimension at one time point. For static models like this one, the positions are the same across all time periods. For dynamic models (`dynamic_uv = TRUE`), positions evolve over time and `latent_positions()` returns each time point separately (see the [dynamic effects vignette](dynamic_effects.html)).

When using dynamic latent positions, the latent space is only identified up to rotation at each time point. The `procrustes_align()` function aligns each time period's positions to the previous one, removing arbitrary rotations so that trajectories across time are interpretable:

```{r}
# procrustes alignment is useful for dynamic models.
# for a static model, positions are already aligned.
aligned <- procrustes_align(fit)
str(aligned$U)
```

## What's Next?

This vignette covered the core workflow: fitting, convergence, GOF, and visualization. For more specialized topics:

- **Single networks (no time series)?** See [Your First AME Model](cross_sec_ame.html) for a detailed cross-sectional walkthrough
- **Two types of nodes?** See [Bipartite Networks](bipartite.html)
- **Evolving network structure?** See [Dynamic Effects](dynamic_effects.html)
- **Just want the quick version?** See [Getting Started](lame.html)

**References**:

Hoff, PD (2021) Additive and Multiplicative Effects Network Models. Statistical Science 36, 34--50.

Morgan, T. Clifton, Bapat, N., & Kobayashi, Y. (2014). Threat and Imposition of Economic Sanctions 1945-2005. Conflict Management and Peace Science 31(5): 541-558.

Minhas, S., Dorff, C., Gallop, M. B., Foster, M., Liu, H., Tellez, J., & Ward, M. D. (2022). Taking dyads seriously. Political Science Research and Methods, 10(4), 703--721. doi:10.1017/psrm.2021.56
