---
title: "Your First AME Model"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Your First AME Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>",
	fig.align = "center"
)
```

## Why Network Models?

Imagine you're studying friendships in a high school. You have data on who nominated whom as a friend, plus information about each student (gender, race, grade). A natural first instinct is to run a logistic regression: does sharing the same gender predict friendship?

The problem is that friendships aren't independent observations. Some students are more social (they nominate lots of friends), some are more popular (they receive lots of nominations), and friendships tend to be reciprocated: if Alice names Bob, Bob is more likely to name Alice. A standard regression ignores all of this, and your standard errors will be wrong.

The **Additive and Multiplicative Effects (AME)** model handles these dependencies directly. It gives each actor a sender effect ($a_i$, how social they are), a receiver effect ($b_j$, how popular they are), and a position in a latent space ($u_i$, $v_j$) that captures who tends to connect with whom beyond what the covariates explain. The additive effects enter as $a_i + b_j$; the latent positions enter *multiplicatively* as the dot product $u_i'v_j$. Think of it as a regression that takes network structure seriously.

## The Data

We'll analyze a friendship network from the [Add Health study](https://addhealth.cpc.unc.edu/), a longitudinal study of adolescents in the United States. The `addhealthc3` dataset in `lame` contains a directed friendship nomination network along with student characteristics.

```{r load-data}
library(lame)
library(ggplot2)
set.seed(6886)

# load the Add Health friendship network
data(addhealthc3)

# convert valued network to binary (any nomination = friendship)
Y <- (addhealthc3$Y > 0) * 1
X_nodes <- addhealthc3$X

n <- nrow(Y)
cat("Students:", n, "\n")
cat("Friendships:", sum(Y, na.rm = TRUE), "\n")
cat("Network density:", round(mean(Y, na.rm = TRUE), 3), "\n")
```

Notice the `na.rm = TRUE` calls: network data often has missing entries (the diagonal is `NA` because self-ties are undefined). The model handles missing values internally via data augmentation, so `NA`s can stay in the matrix -- including dyads built from missing *covariates* (here `race` and `grade` have a handful), which `ame()` likewise treats as unobserved.

Before modeling: how much do students vary in their number of friends?

```{r viz-network, fig.width=10, fig.height=4, fig.alt="Two faceted histograms showing the distribution of nominations sent and nominations received per student in the Add Health friendship network, illustrating heterogeneity in sociality and popularity."}
out_degree <- rowSums(Y, na.rm = TRUE)
in_degree <- colSums(Y, na.rm = TRUE)

degree_df <- data.frame(
	Degree = c(out_degree, in_degree),
	Type = rep(c("Nominations sent", "Nominations received"), each = n)
)

ggplot(degree_df, aes(x = Degree)) +
	geom_histogram(binwidth = 1) +
	facet_wrap(~Type, ncol = 2) +
	labs(x = "Number of Ties", y = "Count") +
	theme_bw() +
	theme(
		panel.border = element_blank(),
		axis.ticks = element_blank(),
		legend.position = "top",
		strip.background = element_rect(fill = "black", color = "black"),
		strip.text = element_text(color = "white", hjust = 0)
	)
```

Some students are much more social than others, and some much more popular -- exactly what the sender and receiver random effects will capture.

## Building Covariates

The classic question in friendship networks is **homophily**: do birds of a feather flock together? We'll test whether students are more likely to be friends with others of the same gender, race, and grade.

We deliberately include both `same_grade` (binary) and `grade_diff` (continuous: grades apart) to illustrate a common modeling pitfall, collinearity: they measure nearly the same thing, and we'll see the consequences in the results below.

```{r create-covariates}
# `outer(x, x, FUN)` builds an n x n matrix whose (i,j) entry is FUN(x[i], x[j]).
# So `outer(female, female, "==")` returns TRUE where students i and j share the
# same gender. Multiplying by 1 converts TRUE/FALSE to 1/0.
#
# homophily indicators: 1 if same, 0 if different
same_female <- outer(X_nodes[,"female"], X_nodes[,"female"], "==") * 1
same_race <- outer(X_nodes[,"race"], X_nodes[,"race"], "==") * 1
same_grade <- outer(X_nodes[,"grade"], X_nodes[,"grade"], "==") * 1
# absolute grade difference; same shape, but a continuous covariate
grade_diff <- abs(outer(X_nodes[,"grade"], X_nodes[,"grade"], "-"))

# pack into a 3D array (n x n x p)
Xdyad <- array(NA, dim = c(n, n, 4))
Xdyad[,,1] <- same_female
Xdyad[,,2] <- same_race
Xdyad[,,3] <- same_grade
Xdyad[,,4] <- grade_diff
dimnames(Xdyad)[[3]] <- c('same_female', 'same_race', 'same_grade', 'grade_diff')
for(k in 1:4) diag(Xdyad[,,k]) <- NA

# nodal covariates (sender and receiver characteristics)
Xrow <- X_nodes[, c("female", "grade")]
Xcol <- X_nodes[, c("female", "grade")]
```

## Fitting the Model

Now the fun part. We fit a binary probit AME model with:

- **Covariates** for homophily effects
- **Sender and receiver random effects** (sociality / popularity)
- **Dyadic correlation** (reciprocity)
- **2-dimensional latent space** (residual clustering)

### A note for ERGM users: what AME assumes

If you're coming from `ergm` / statnet, the key difference is the dependence assumption. ERGM uses change statistics (e.g. `gwesp`, `triangle`, `kstar`) to encode *unconditional* higher-order dependence: changing one tie shifts the probability of other ties directly. AME instead assumes **conditional dyadic independence**: given the additive effects $(a_i, b_j)$, the latent positions $(u_i, v_j)$, the dyadic correlation $\rho$, and the covariates, all dyads are independent, and higher-order structure is captured *indirectly* by integrating over the latent effects.

Three practical consequences:

- **No degeneracy.** AME avoids the model-degeneracy failure modes of ERGMs with `triangle` or unconstrained `kstar` terms.
- **No change-statistic interpretation of $\beta$.** A coefficient on `same_grade` is a partial association on the probit-latent scale, not a log-odds change conditional on the rest of the network.
- **Some triangle structure is missed.** When transitive closure is the substantive target (`gwesp` is the leading example), ERGM is the right tool; for structural pattern + actor heterogeneity + covariate effects with calibrated uncertainty, AME is generally the more stable estimator.

```{r fit-model, message=FALSE, warning=FALSE}
fit <- ame(Y,
					Xdyad = Xdyad,
					Xrow = Xrow,
					Xcol = Xcol,
					R = 2,              # 2D latent space
					family = "binary",  # probit model for 0/1 data
					rvar = TRUE,        # sender random effects
					cvar = TRUE,        # receiver random effects
					dcor = TRUE,        # dyadic correlation (reciprocity)
					burn = 500,         # burn-in (lengthen for a final run)
					nscan = 2000,       # post-burn-in iterations
					odens = 25,         # thinning
					verbose = FALSE,
					gof = TRUE)
```

## What Did We Find?

```{r print-summary}
summary(fit)
```

Let's unpack the key results:

**Homophily effects.** The `grade_diff` coefficient is strongly negative and its credible interval excludes zero: students further apart in grade are much less likely to be friends. The `same_female` coefficient is positive (gender homophily) but its 95% credible interval includes zero -- read the sign as suggestive, not the effect as established. The `same_race` coefficient captures race homophily after controlling for grade and gender.

**The collinearity pitfall.** `same_grade` and `grade_diff` measure nearly the same thing -- a grade difference of 0 *is* the same-grade event -- so the model cannot cleanly *split* the grade signal between them. It still identifies the dominant effect: `grade_diff` is strongly negative and stable from run to run (about `r round(mean(fit$BETA[, "grade_diff_dyad"]), 2)`, never near zero). What suffers is the partition: `same_grade` is left to absorb a small, noisy residual bump on top of the linear grade-distance trend. The lesson is about *interpretation*, not estimate instability: include one or the other, not both, so the grade effect lands on a single, clean coefficient.

**Variance components.** The sender variance (`va`) and receiver variance (`vb`) quantify how much students differ in sociality and popularity. The dyadic correlation (`rho`) captures reciprocity: values near 1 mean that if A nominates B, B almost always nominates A back. Reciprocity is strong here.

For binary networks with a probit link, the coefficients are on the latent scale, and the textbook "$0.4 \times \beta$" rule for converting them to probability changes is only sharp when the baseline probability is near 0.5 -- which this sparse network is not. Rather than juggling correction factors, compute the quantity you actually want with `predict(fit, type = "response")` and contrast predicted probabilities under counterfactual covariate settings.

## Did the Sampler Converge?

The model is estimated via MCMC (Markov chain Monte Carlo), so trace plots are the first check that the sampler explored the posterior thoroughly.

```{r trace-plots, fig.width=10, fig.height=8, fig.alt="MCMC trace and density plots for each regression coefficient; stable traces fluctuate around a steady mean and the density plots are smooth and unimodal."}
trace_plot(fit, params = "beta", ncol = 3)
```

The regression coefficients are not the whole story, though -- the slowest-mixing term in these fits is usually the dyadic correlation `rho`, so it deserves its own trace:

```{r trace-rho, fig.width=8, fig.height=5, fig.alt="MCMC trace and density plot for the dyadic correlation (rho); the trace moves slowly relative to the regression coefficients, illustrating the slow mixing discussed in the text."}
trace_plot(fit, params = "variance", include = "Dyadic Correlation")
```

**What to look for:** Traces should fluctuate around a stable mean (long trends mean non-convergence) and densities should be smooth and unimodal. With `r nrow(fit$BETA)` post-burn-in samples most coefficients mix acceptably for a demo; the slowest at this seed is `rho` -- the reciprocity parameter that is chronically the slowest-mixing term in these fits (bulk ESS around 27 here, versus roughly 35--112 for the regression coefficients). Read low ESS as a *mixing* problem, not an *estimate* problem: point estimates such as `grade_diff`'s are stable from run to run even on short chains. But a slow-mixing parameter is exactly the one whose uncertainty you should not quote from a single short chain -- `rho` gets its real test in the multi-chain run below.

### Numerical convergence diagnostics: `posterior::as_draws()`

A trace plot is a sanity check, not a diagnostic. The Stan-era summary triple is **split-$\hat R$**, **bulk ESS**, and **tail ESS**; `lame` registers an `as_draws()` method, so the same `posterior`-package workflow you would run on a `stanfit` works here:

```{r as-draws-rhat-ess, eval = requireNamespace("posterior", quietly = TRUE), message = FALSE, warning = FALSE}
library(posterior)
draws <- posterior::as_draws(fit)              # draws_array [iter, chain, var]
posterior::summarise_draws(draws)              # mean, sd, q5, q95, rhat,
                                               # ess_bulk, ess_tail per param
```

The conventional thresholds: split-$\hat R$ < 1.01 for every monitored parameter; `ess_bulk` and `ess_tail` $\ge$ 400 per chain. This demo stores only `r nrow(fit$BETA)` draws (one short chain), so it cannot reach the ESS target -- the largest `ess_bulk` in the table is only about 112, and `rho` sits lowest at about 27. The $\hat R$ column, by contrast, looks deceptively clean: every parameter comes back at or below about 1.08, with `rho` near 1.01--1.02. That is the trap of single-chain diagnostics: split-$\hat R$ computed on one chain compares the two halves of that chain, so it flags within-chain non-stationarity but cannot detect two chains that have each settled in different parts of the posterior. The parameter that actually fails to converge on this network -- `rho` -- looks fine in this table and only reveals itself when separate chains disagree, which is exactly what the four-chain fit below exposes ($\hat R \approx 1.8$). One row will look degenerate in both tables: `ve` is the latent residual variance, fixed at 1 for probit identification, so its draws are constant and `posterior` correctly reports NA for $\hat R$ and ESS -- ignore that row.

### Between-chain $\hat R$: `ame_parallel(n_chains = 4)`

For real convergence assessment, run four chains from different seeds and let `summarise_draws()` compute Rhat across chains:

```{r ame-parallel-rhat, eval = requireNamespace("posterior", quietly = TRUE)}
# reduced chains/iterations for a fast build; use n_chains = 4 and
# nscan >= 2000 for a convergence check you will report.
fit_mc <- ame_parallel(
	Y, Xdyad = Xdyad, Xrow = Xrow, Xcol = Xcol,
	family = "binary", R = 2,
	burn = 200, nscan = 800, odens = 10,
	n_chains = 2, cores = 1,            # cores = 1 keeps the chunk portable
	combine_method = "pool",            # pooled fit; chain_indicator is preserved
	verbose = FALSE
)
posterior::summarise_draws(posterior::as_draws(fit_mc))
```

This is the payoff of running more than one chain. In the single-chain table above, `rho` looked merely mediocre -- $\hat R \approx 1.02$, bulk ESS around 27 -- for the reason just described: one chain can only be compared with itself. With a second, independently seeded chain the failure is unmistakable: the convergence banner flags `rho` at R-hat = 1.801, and `summarise_draws()` puts it at 1.63 with a bulk ESS of 3.7. (The two numbers differ because the banner computes the classic Gelman-Rubin ratio while `posterior` uses rank-normalized split-$\hat R$; both are far past the 1.01 bar.) This is the same reciprocity parameter the trace plots flagged as chronically slow-mixing. Do not report `rho`'s credible interval from a run that looks like this -- increase `nscan` and the number of chains until its $\hat R$ clears the threshold. Every other parameter sits at $\hat R \le 1.02$ across chains, so there is no sign of between-chain disagreement; the covariate and variance-component estimates still need a longer multi-chain run to clear the ESS bar before you quote their uncertainty.

The pooled object carries a `chain_indicator` so `as_draws()` keeps chain identity separate; the same reshape feeds `bayesplot::mcmc_trace()` and `tidybayes::tidy_draws()` directly.

### Other quick checks

Two cheap diagnostics worth running once per fit. `lame` is a Gibbs / Metropolis-Hastings sampler, not HMC, so it has no divergences -- the closest analog is the per-block failure count exposed as `fit$mh_counters`; treat any block with more than 5% failures as suspicious. And `prior_summary(fit)` prints the hyperparameters *actually used* (defaults filled in), the cheapest way to catch a typo in a `prior = list(...)` override.

```{r mh-counters}
str(fit$mh_counters)
```

```{r prior-summary-cross}
prior_summary(fit)
```

### Model comparison: `loo()` and `loo_compare()`

If you refit with `save_log_lik = TRUE`, the fit carries a `[n_stored, n_obs]` pointwise log-likelihood matrix and `loo::loo(fit)` works directly via the registered S3 method:

```{r loo-cross, eval = requireNamespace("loo", quietly = TRUE), message = FALSE}
# reduced iterations for a fast build; use nscan >= 2000 for a reported result.
fit_ll <- ame(Y, Xdyad = Xdyad, Xrow = Xrow, Xcol = Xcol,
              family = "binary", R = 2,
              burn = 200, nscan = 800, odens = 10,
              save_log_lik = TRUE, verbose = FALSE)

# alternative model: drop the dyadic covariate
fit_ll_null <- ame(Y, Xrow = Xrow, Xcol = Xcol,
                   family = "binary", R = 2,
                   burn = 200, nscan = 800, odens = 10,
                   save_log_lik = TRUE, verbose = FALSE)

loo_dyad <- loo::loo(fit_ll)
loo_null <- loo::loo(fit_ll_null)

loo_dyad   # read the Pareto-k table before any elpd number

loo::loo_compare(list(with_dyad = loo_dyad, no_dyad = loo_null))
```

This time the diagnostics matter as much as the comparison. Start with the warning `loo()` raises and the Pareto-$k$ table it prints: 287 of the 992 dyads (28.9%) fall in the "bad" range ($0.47 < k \le 1$) and 11 more (1.1%) are "very bad" ($k > 1$), so for nearly a third of the observations the importance weights behind `elpd_loo` cannot be trusted. This failure mode is structural rather than a fluke: every dyad's likelihood depends on the actor-level random effects ($a_i$, $b_j$, $u_i'v_j$), so leaving one dyad out shifts the posterior enough to make the importance ratios heavy-tailed, and with only 80 stored draws there is little to stabilize them. Read the comparison in that light: `loo_compare()` sorts `with_dyad` first, but its 24.6-point advantage is smaller than its own standard error (`se_diff` 25.6) -- not decisive even at face value -- and with this many flagged dyads the honest conclusion is to not lean on `elpd_diff` at all. The workflow lesson: always read the Pareto-$k$ table *before* `elpd_diff`. For a comparison you can report, lengthen the run substantially (`nscan >= 2000`, more stored draws) and, for random-effects-dense network models like AME, prefer $K$-fold cross-validation over dyads -- or the held-out dyad evaluation shown below via `evaluate_heldout()` -- to PSIS-LOO.

For the `normal`, `binary`, `cbin`, `poisson`, and `ordinal` families the stored log-likelihood is the **exact family-specific Y density**, so `elpd_loo` is directly comparable to a `loo()` from a brms or rstanarm fit to the same family; see `?loo.ame` for the `frn` rank-likelihood caveat and `fit$log_lik_method`.

## Does the Model Fit the Data?

Goodness-of-fit (GOF) checks whether the model can reproduce structural features of the observed network -- degree heterogeneity, reciprocity, clustering -- beyond dyad-level prediction.

```{r gof-plot, fig.width=10, fig.height=6, fig.alt="Goodness-of-fit panels comparing observed network statistics (dashed orange vertical line) against the posterior predictive histograms (grey) for sender and receiver degree heterogeneity, dyadic dependence, triadic dependence, and transitivity. Observed and predicted differ on both colour and linetype so the cue survives grayscale printing and colour-blind viewing."}
gof_plot(fit)
```

In each panel the grey histogram is the posterior-predictive distribution from networks simulated out of the fitted model, and the **dashed orange vertical line is the observed value** (colour and linetype dual-encode the contrast so it survives grayscale printing and colour-blind viewing). See `?gof_plot` for the `statistics = ...` aliases and the exact column names stored in `fit$GOF`.

The observed values for Sender and Receiver Degree Heterogeneity and for Dyadic Dependence (essentially empirical reciprocity) fall inside their posterior-predictive histograms -- those three features are explicitly modeled by the random effects, the variance components, and $\rho$. The two triad-level statistics tell a different story: observed Triadic Dependence sits in the upper tail, and observed Transitivity at the far right edge. AME under-predicts triangle closure on this network, the expected signature of conditional dyadic independence: the latent space soaks up some triangle structure through clustering of the $u_i$, but there is no explicit closure term. If transitivity is your substantive target, fit an ERGM; if it is a diagnostic concern only, increasing `R` sometimes helps modestly.

You can also compute GOF after the fact using the `gof()` function, which accepts custom statistics. The built-in `trans.dep` panel is a *correlation-style* clustering measure; ERGM users will typically want the raw clustering coefficient (`sna::gtrans`, the scalar summary that `gwesp(decay = 0, fixed = TRUE)` targets), which takes a few lines as a custom stat:

```{r custom-gof}
# classic clustering coefficient: transitive triples / two-paths
trans_ratio <- function(Y) {
	Yb <- Y; Yb[is.na(Yb)] <- 0            # NA -> 0 for matrix multiply
	YY <- Yb %*% Yb
	triangles <- sum(YY * Yb)              # transitive triples
	two_path  <- sum(YY) - sum(diag(YY))   # two-paths (potential triangles)
	c(trans_ratio = triangles / max(two_path, 1))
}

gof_custom <- gof(fit, custom_gof = trans_ratio, nsim = 50, verbose = FALSE)
obs <- gof_custom[1, "trans_ratio"]
sim <- gof_custom[-1, "trans_ratio"]

# posterior-predictive p-value (right tail): near 0 = under-predicted
c(observed = round(obs, 3), sim_mean = round(mean(sim), 3),
  pp_p_right = round(mean(sim >= obs), 3))
```

Read this the way you would read `ergm::gof()`: `pp_p_right` is the fraction of simulated networks at or above the observed value, so a value near 0 means the model under-predicts the statistic. On Add Health it comes back close to zero (around 0.08 in our run) -- the model produces triangles, but does not concentrate them per two-path the way the data do, the same structural gap the built-in panels showed. See `?gof` for adding further custom statistics (raw triangle counts, degree assortativity, and so on).

## Visualizing the Latent Space

The multiplicative effects ($u_i'v_j$) place each student in a 2D latent space: students near each other in sender space (triangles) nominate similar friends, and students near each other in receiver space (circles) are nominated by similar students. Clusters likely share some unobserved characteristic (same social group, same extracurriculars) that drives friendship choices beyond gender, race, and grade.

```{r latent-space, fig.width=8, fig.height=6, fig.alt="Biplot of the estimated 2D latent space: triangles mark each student's sender position and circles mark receiver positions on the two latent dimensions; students plotted near each other have similar friendship patterns beyond what the covariates explain."}
uv_plot(fit, layout = "biplot", show.edges = FALSE, label.nodes = FALSE)
```

## Who Are the Most Social / Popular Students?

The additive effects decompose individual heterogeneity into sender effects (sociality) and receiver effects (popularity):

```{r individual-effects, fig.width=10, fig.height=8, fig.alt="Two stacked lollipop charts showing the posterior-mean sender effect (sociality) and receiver effect (popularity) for each student as a point connected by a stem to a dashed zero line, sorted from the most negative effect on the left to the most positive on the right; positive values sit above the line and negative below."}
p1 <- ab_plot(fit, effect = "sender", sorted = TRUE,
							title = "Sender Effects (Sociality)")
p2 <- ab_plot(fit, effect = "receiver", sorted = TRUE,
							title = "Receiver Effects (Popularity)")

library(patchwork)
p1 / p2
```

Students with large positive sender effects nominate more friends than expected given their covariates; those with large positive receiver effects receive more nominations.

## Making Predictions

The model gives you predicted probabilities for every possible tie, useful for link prediction and for understanding dyad-level fit. One caution for sparse networks: predicting "no tie" for every dyad already gets most predictions right, so raw accuracy flatters the model -- judge it against the modal-class baseline.

```{r predictions}
pred_resp <- predict(fit, type = "response")

# how well does the model classify?
Y_vec <- as.vector(Y)
pred_vec <- as.vector(pred_resp)
keep <- !is.na(Y_vec)

# simple classification at threshold 0.5
pred_binary <- (pred_vec[keep] > 0.5) * 1
confusion <- table(Actual = Y_vec[keep], Predicted = pred_binary)
knitr::kable(confusion, caption = "Confusion Matrix (threshold = 0.5)")

accuracy <- sum(diag(confusion)) / sum(confusion)
baseline <- max(mean(Y_vec[keep]), 1 - mean(Y_vec[keep]))  # always-predict-modal-class
cat("\nAccuracy:", round(accuracy, 3),
    " | Baseline (predict modal class):", round(baseline, 3),
    " | Lift:", round(accuracy - baseline, 3), "\n")
```

At density ~13% the modal-class baseline is ~0.87, so the relevant question is whether the model's accuracy is materially above 0.87, not whether it crosses 0.5.

For link prediction, AUC-style metrics evaluate the model's ability to rank true ties above non-ties across all thresholds. For sparse networks PR-AUC is the more informative summary: a model that always predicts "no tie" already has AUROC near 0.5 but PR-AUC near the network density, so the gap between PR-AUC and density is the metric of substantive interest.

```{r auc-eval, eval = requireNamespace("pROC", quietly = TRUE) && requireNamespace("precrec", quietly = TRUE)}
# pROC: ROC curve and AUROC
# install.packages("pROC")
roc_obj <- pROC::roc(response = Y_vec[keep], predictor = pred_vec[keep],
                     quiet = TRUE)
auroc   <- as.numeric(pROC::auc(roc_obj))

# precrec: both ROC and Precision-Recall AUCs (the latter is more
# informative when the positive class is rare, as in friendship networks)
# install.packages("precrec")
ev <- precrec::evalmod(scores = pred_vec[keep], labels = Y_vec[keep])
precrec::auc(ev)   # returns AUROC and PR-AUC side-by-side

cat("AUROC:", round(auroc, 3), "\n",
    "Baseline AUROC (random ranker):", 0.5, "\n",
    "Baseline PR-AUC (density of Y):", round(mean(Y_vec[keep]), 3), "\n")
```

Here that gap is wide: PR-AUC 0.603 against a density baseline of 0.128 (roughly five times the base-rate floor), and AUROC 0.874 against 0.5 for a random ranker. Both are in-sample numbers, though -- the held-out estimates in the next section are the ones to quote for any predictive claim.

### Held-out link prediction

For a held-out evaluation, mask a random subset of dyads to `NA` before fitting (the sampler excludes them from the likelihood via data augmentation), then score `predict(fit, type = "response")` on the masked indices. Stratify the split on `Y` so the test set contains both classes.

```{r heldout}
# 80/20 stratified mask of off-diagonal dyads
set.seed(6886)
off_diag    <- which(row(Y) != col(Y) & !is.na(Y))
pos_idx     <- off_diag[Y[off_diag] == 1]
neg_idx     <- off_diag[Y[off_diag] == 0]
test_idx    <- c(sample(pos_idx, round(0.2 * length(pos_idx))),
                 sample(neg_idx, round(0.2 * length(neg_idx))))

Y_train               <- Y
Y_train[test_idx]     <- NA      # mask test dyads from the likelihood

fit_train <- ame(Y_train, Xdyad = Xdyad, Xrow = Xrow, Xcol = Xcol,
                 R = 2, family = "binary",
                 rvar = TRUE, cvar = TRUE, dcor = TRUE,
                 burn = 200, nscan = 800, odens = 10,   # reduced for build speed
                 verbose = FALSE, gof = FALSE)

pred_train <- predict(fit_train, type = "response")

# `evaluate_heldout()` scores predictions on a logical TEST mask:
# columns n_eval, auroc, auprc, brier, logloss
test_mask           <- matrix(FALSE, nrow(Y), ncol(Y),
                              dimnames = dimnames(Y))
test_mask[test_idx] <- TRUE
evaluate_heldout(y_obs = Y, y_pred = pred_train,
                 mask = test_mask, family = "binary")
cat("Test dyads:", length(test_idx),
    " | positives:", sum(Y[test_idx] == 1), "\n")
```

On the 198 held-out dyads (25 of them true friendships) the model posts AUROC 0.836 and PR-AUC 0.551 -- only a modest drop from the in-sample 0.874 / 0.603, the reassuring signal that the fit generalizes rather than memorizing the training dyads. PR-AUC is the number to weigh: a random ranker's PR-AUC equals the positive rate (about 0.13), so the model's precision runs roughly four times chance on dyads it never saw. With only 25 positive test dyads and a single short chain, read these as a sanity check rather than a benchmark -- report the positive count alongside them, and see `?evaluate_heldout` for the metric definitions and options.

## Simulating from the Model

You can generate new networks from the fitted posterior. A **posterior
predictive check** asks: does a network simulated from posterior draws
look like the observed one on a feature you care about (here, overall
density)? Observed values inside the simulated distribution mean the
model reproduces that feature; values in a tail mean it is missing
something.

```{r simulation, fig.width=8, fig.height=4, fig.alt="Posterior predictive check histogram of simulated network densities (grey bars) with a dashed Okabe-Ito orange vertical line marking the observed density; the encoding is dual on colour and linetype so the cue survives greyscale printing and colour-blind viewing. Overlap indicates the model reproduces the observed density."}
sims <- simulate(fit, nsim = 100)

# compare simulated vs observed density
sim_densities <- sapply(sims$Y, function(y) mean(y, na.rm = TRUE))
obs_density <- mean(Y, na.rm = TRUE)

ggplot(data.frame(density = sim_densities), aes(x = density)) +
	geom_histogram(bins = 20, fill = "grey60") +
	# dual-encode the observed density as colour AND linetype so the cue
	# survives greyscale printing and colour-blind viewing (Okabe-Ito orange).
	geom_vline(xintercept = obs_density,
	           color = "#D55E00", linewidth = 1, linetype = "dashed") +
	labs(title = "Posterior Predictive Check: Network Density",
			subtitle = "Dashed orange line = observed; grey histogram = simulated from model",
			x = "Network Density (Mean Tie Probability)", y = "Replicate Count") +
	theme_bw() +
	theme(panel.border = element_blank(), axis.ticks = element_blank(),
	      legend.position = "top")
```

## Practical Tips

**Choosing R (latent dimensions).** Start with R = 0 (no latent space), then try R = 1 and R = 2 and compare GOF. `ame()` / `lame()` warn when `R > floor(n/3)` because at that rank the multiplicative effects absorb structure that belongs to the additive effects; in practice `R = 2` or `R = 3` is the right default.

**MCMC settings.** For exploratory work, `burn = 500, nscan = 2000, odens = 25` is fine. For a final run, use at least `burn = 2000, nscan = 10000, odens = 25` and check convergence with `trace_plot()`.

**Variance components.** Include `rvar = TRUE` and `cvar = TRUE` unless you have a specific reason not to. Include `dcor = TRUE` for directed networks where reciprocity is plausible.

## What's Next?

- **Multiple time periods?** See the [lame overview](lame-overview.html) for longitudinal models
- **Two types of nodes?** See the [bipartite vignette](bipartite.html)
- **Evolving network structure?** See the [dynamic effects vignette](dynamic_effects.html)
- **Other data types?** See `?ame` for the supported `family` options

## References

Hoff, PD (2021). Additive and Multiplicative Effects Network Models. *Statistical Science* 36, 34--50.

Minhas, S., Dorff, C., Gallop, M. B., Foster, M., Liu, H., Tellez, J., & Ward, M. D. (2022). Taking dyads seriously. *Political Science Research and Methods*, 10(4), 703--721.
