---
title: "lame Overview"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lame Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>",
	fig.width = 6,
	fig.height = 4,
	dpi = 96,
	out.width = "100%",
	fig.align = "center",
	dev = "png"
)
```

## Package Overview

The `lame` package provides tools for fitting **L**ongitudinal **A**dditive and **M**ultiplicative **E**ffects models to network data observed over time. If you study relationships between actors (countries trading with each other, legislators co-sponsoring bills, students forming friendships across semesters) and you have repeated observations of those relationships, `lame` is designed for you.

The modeling approach builds on the Additive and Multiplicative Effects (AME) framework developed by Peter Hoff, whose [`amen`](https://pdhoff.github.io/amen/) package ([GitHub](https://github.com/pdhoff/amen)) provides the foundational implementation for cross-sectional network analysis. The `lame` package extends this framework in several directions:

- **Longitudinal and replicated data**: While `amen` focuses on single networks (with `ame()`) or basic replicated designs (with `ame_rep()`), `lame` is built from the ground up for panel data, networks observed across multiple time periods, potentially with actors entering and leaving the sample.
- **Dynamic effects**: Actors' positions in latent space and their baseline activity levels can evolve over time via AR(1) processes (`dynamic_uv` and `dynamic_ab`), letting you model how network structure changes rather than assuming it is fixed.
- **C++ acceleration**: Core sampling routines are implemented in Rcpp/RcppArmadillo for faster computation, which matters when you have many time periods or actors.
- **Bipartite networks**: Explicit support for two-mode networks (e.g., countries and treaties, users and products) with separate latent spaces for row and column nodes.
- **ggplot2-based diagnostics**: Built-in visualization functions (`trace_plot`, `gof_plot`, `ab_plot`, `uv_plot`) return standard ggplot2 figures you can theme and extend.
- **Standard S3 methods**: `coef()`, `confint()`, `fitted()`, `residuals()`, `predict()`, and `simulate()` work as you would expect from any R model object.

The package works with [`netify`](https://netify-dev.github.io/netify/) for data preparation.

### Migrating from `amen`

Both `amen` and `lame` export an `ame()` function. If you load both packages in the same session, `lame` prints a `.onAttach` startup message so you know unqualified `ame(...)` calls dispatch to whichever package was attached last; use `lame::ame(...)` or `amen::ame(...)` to be explicit. The table below maps the breaking differences for old `amen` scripts:

| Topic | `amen::ame()` | `lame::ame()` / `lame::lame()` | Old code breaks? |
|---|---|---|---|
| Family strings | `"nrm"`, `"bin"`, `"ord"`, `"cbin"`, `"frn"` | `"normal"`, `"binary"`, `"ordinal"`, `"cbin"`, `"frn"`, `"poisson"` | No -- amen short forms are accepted as aliases (with a message recommending the full name). The ALS fitters support only normal/binary/poisson and reject the rest with a pointer to MCMC. |
| Bipartite rank | single `R` | `R` (square) or separate `R_row` / `R_col` (bipartite); single `R` still accepted | No (old `R` calls work; `R_row` / `R_col` is new) |
| Verbosity flag | `print = TRUE/FALSE` | `verbose = TRUE/FALSE` | No (deprecation warning) |
| `fit$BETA` shape (static) | `[n_stored, p]` matrix | `[n_stored, p]` matrix | No (identical) |
| `fit$BETA` shape (`dynamic_beta`) | not available | `[n_stored, p, T]` 3-D array | New feature; use `coef(fit)` (returns `[p, T]` for dynamic fits) -- a bare `apply(fit$BETA, 2, mean)` silently averages over time. See [Dynamic Effects](dynamic_effects.html). |
| Longitudinal data | `ame_rep()` for replicates | `lame()` with a list of `T` matrices and optional `dynamic_uv` / `dynamic_ab` / `dynamic_beta` | Refit via `lame()` |
| GOF output | `fit$GOF` matrix | `fit$GOF` (cross-sectional) or list-of-matrices per period (longitudinal) | Yes for longitudinal scripts |
| S3 methods | `summary()`, `plot()` | adds `coef()`, `confint()`, `fitted()`, `residuals()`, `predict()`, `simulate()`, `autoplot()`, `tidy()` | No (additive) |
| `loo` / `waic` | not available | `save_log_lik = TRUE` plus `loo::loo(fit)` / `loo::waic(fit)` S3 methods; see [Dynamic Effects](dynamic_effects.html) | N/A (new feature) |
| `broom::tidy()` | not available | `tidy.ame()` / `tidy.lame()` registered against `generics::tidy` | N/A (new feature) |
| `broom::glance()` | not available | `glance.ame()` / `glance.lame()` registered against `generics::glance`; returns `nobs`, `n_actors`, `n_periods`, `n_stored`, `family`, `mode`, `R`, `dynamic_uv`, `dynamic_ab`, `dynamic_beta`, `elpd_loo`. Pairs with `tidy()` for `modelsummary::modelsummary(fit)`. | N/A (new feature) |

The most common porting fix is `family = "nrm"` → `family = "normal"` and `family = "bin"` → `family = "binary"`. Everything else either still works unchanged or surfaces a deprecation warning rather than an error.

## Application: Dutch College Friendships

To see how `lame` works in practice, we analyze the **Dutch college friendship network** (van de Bunt, van Duijn, and Snijders 1999), a panel of 32 students who arrived as strangers and rated each other's friendship at seven time points. (We drop the cold-start first wave, when the students barely knew each other and almost no ties exist; six waves remain.) The substantive question is whether students sort by gender, smoking status, and academic program, and whether residual clustering remains once those covariates are absorbed.

The dataset includes:

- **Y**: directed friendship ratings on a -1 to 4 scale (-1 a negative/troubled relationship, 0 no or uncertain tie, 1-4 increasing friendship). We binarize any positive rating to a tie (`(Y > 0) * 1`), so the rare -1 ratings fold in with 0 as "no tie".
- **X**: three node-level attributes: `male`, `smoker`, `program`.
- We construct three dyadic **homophily indicators** from the node attributes: `same_male`, `same_smoker`, `same_program`.

```{r setup}
library(lame)
library(ggplot2)
set.seed(6886)

data("dutchcollege")

n <- nrow(dutchcollege$Y)
T_all <- dim(dutchcollege$Y)[3]
actor_names <- sprintf("S%02d", seq_len(n))

# binarize and drop the cold-start wave (t = 1 has almost no ties)
Y <- lapply(2:T_all, function(t) {
	Yt <- (dutchcollege$Y[, , t] > 0) * 1
	diag(Yt) <- NA
	rownames(Yt) <- colnames(Yt) <- actor_names
	Yt
})
names(Y) <- paste0("t", 2:T_all)

# nodal covariates passed identically as sender and receiver
X_node <- dutchcollege$X
rownames(X_node) <- actor_names
Xrow <- lapply(seq_along(Y), function(t) X_node)
Xcol <- Xrow

# dyadic homophily indicators
same_male    <- outer(X_node[, "male"],    X_node[, "male"],    "==") * 1
same_smoker  <- outer(X_node[, "smoker"],  X_node[, "smoker"],  "==") * 1
same_program <- outer(X_node[, "program"], X_node[, "program"], "==") * 1
Xdyad_one <- array(0, dim = c(n, n, 3),
                   dimnames = list(actor_names, actor_names,
                                   c("same_male", "same_smoker", "same_program")))
Xdyad_one[, , 1] <- same_male
Xdyad_one[, , 2] <- same_smoker
Xdyad_one[, , 3] <- same_program
Xdyad <- lapply(seq_along(Y), function(t) Xdyad_one)
```

The friendship network's average density rises across the panel as students settle in, though the trajectory is not monotone (density bounces between roughly 0.40 and 0.63 across the six waves):

```{r}
sapply(Y, function(y) round(mean(y, na.rm = TRUE), 2))
```

### Fitting the Model

We fit a binary probit AME model with sender random effects (`rvar`), receiver random effects (`cvar`), dyadic correlation (`dcor`), and a one-dimensional latent space (`R = 1`). The random effects capture sender heterogeneity (some students simply nominate more friends) and receiver heterogeneity (some students are nominated more often), while the latent space picks up residual clustering: students who befriend similar people for reasons that the observed covariates do not capture. We use `R = 1` because with only 32 students a two-dimensional latent space competes with the three dyadic homophily indicators for the same signal and the chain mixes poorly on the dyadic coefficients; `R = 1` resolves that cleanly and the substantive results are nearly identical.

```{r, message=FALSE, warning=FALSE}
fit <- lame(
	Y = Y,
	Xdyad = Xdyad,           # dyadic homophily indicators
	Xrow = Xrow,             # sender covariates
	Xcol = Xcol,             # receiver covariates
	family = "binary",       # binary probit model
	rvar = TRUE,             # sender random effects
	cvar = TRUE,             # receiver random effects
	dcor = TRUE,             # dyadic correlation (reciprocity)
	R = 1,                   # 1-D multiplicative latent space
	symmetric = FALSE,       # friendships are directed
	burn = 500,              # burn-in
	nscan = 4000,            # post-burn-in iterations
	odens = 10,              # thinning -> 400 stored draws
	posterior_opts = list(save_UV = TRUE),  # keep U/V draws: gives
	                                        # latent_positions() real
	                                        # posterior SDs below
	save_log_lik = TRUE,     # pointwise log-lik: feeds loo() at the end
	verbose = FALSE,
	plot = FALSE
)
```

### Interpreting the Results

```{r}
summary(fit)
```

A lot lands at once, so let's walk through it.

**Intercept.** The probit intercept is the latent linear predictor when *every covariate is zero* and every random effect sits at its mean. That is not the same as the average student: the covariates here are uncentered, and `program` in particular is coded 2-4 and is never zero, so "all covariates zero" extrapolates well outside the observed data. The familiar `qnorm(density)` rule of thumb describes the predictor at the *average* covariate profile, not at the all-zero corner. The two reconcile exactly once you add back the mean covariate contribution:

```{r intercept-decomp}
bhat        <- colMeans(fit$BETA)
slice_means <- apply(fit$X[[1]], 3, mean, na.rm = TRUE)   # mean of each design column
centroid_lp <- sum(bhat * slice_means[names(bhat)])       # predictor at the average profile
c(intercept     = unname(bhat["intercept"]),
  centroid_lp   = centroid_lp,
  qnorm_density = qnorm(mean(unlist(Y), na.rm = TRUE)))
```

The intercept itself is strongly negative, but the predictor at the average covariate profile (`centroid_lp`) sits close to `qnorm(density)` -- the uncentered covariates account for nearly the entire gap, leaving only a small shift from the random-effect variance. The intercept absorbs the mean covariate contribution; it is not a baseline probability. If you want the intercept to read directly as a baseline, center the covariates before fitting. Either way, read predicted probabilities on the response scale off `predict(fit, type = "response")`.

**Homophily.** The three dyadic coefficients (`same_male`, `same_smoker`, `same_program`) measure whether students of the same type are more likely to be friends, holding the sender, receiver, and latent positions fixed. Positive values mean homophily; credible intervals that exclude zero indicate that the direction of the association is well-identified. Program homophily tends to be the strongest of the three in this network, consistent with the substantive expectation that students bond around shared coursework.

**Nodal covariates.** `male_row` / `male_col` measure whether male students are unusually active senders or popular receivers, and similarly for `smoker` and `program`. With only 32 students, the credible intervals on individual nodal effects are wide; treat the sign as suggestive, the magnitude with caution.

**Variance components.** `va` (sender variance) and `vb` (receiver variance) quantify how much students differ in sociability and popularity beyond what the covariates explain. The dyadic correlation `rho` captures reciprocity: values away from zero indicate that ties tend to be mutual. In friendship networks `rho > 0` is the rule rather than the exception.

### Checking Convergence

The model is estimated via MCMC, so trace plots and posterior-density plots are useful checks. The `trace_plot` function shows the sampled values over iterations (top panels) and the corresponding posterior densities (bottom panels) for each parameter.

```{r overview-trace, fig.width=8, fig.height=8, dpi=100, dev="png", fig.alt="MCMC trace plots (top) and marginal posterior density plots (bottom) for each regression coefficient of the Dutch college fit; well-mixed fuzzy-caterpillar traces around a stable mean and smooth unimodal densities together indicate the chain has converged."}
trace_plot(fit, params = "beta")
```

Trace plots should bounce around a stable mean without long-term trends or sticky regions, and density plots should be smooth and unimodal. Numerically, common MCMC checks are **split-$\hat R$ < 1.01** for monitored parameters and **bulk / tail ESS $\ge$ 400 per chain**.

With `r nrow(fit$BETA)` stored samples, the regression coefficients mix well at this seed. The variance components mix more slowly than the regression coefficients -- compare their $\hat R$ and ESS columns in the `summarise_draws` table below. Which parameter mixes worst is seed-dependent in a short chain, so treat this fit as a compact worked example rather than a final diagnostic run.

### Bayesian-ecosystem diagnostics: `posterior::as_draws()`

`lame` registers `as_draws()` methods so any fit drops straight into the Stan-era diagnostic ecosystem (`posterior`, `bayesplot`, `tidybayes`). The reshape lifts the `BETA` and `VC` blocks into a `draws_array` indexed by iteration / chain / variable, which means `posterior::summarise_draws()` gives you per-parameter posterior mean, median, sd, MAD, 5%/95% quantiles, **split-$\hat R$**, and **bulk / tail ESS** in one call:

```{r as-draws-demo, eval = requireNamespace("posterior", quietly = TRUE)}
library(posterior)
draws <- posterior::as_draws(fit)              # draws_array [iter, chain, var]
posterior::summarise_draws(draws)              # rhat, ess_bulk, ess_tail per param
```

A single-chain fit like this one has `chain = 1`, so $\hat R$ reduces to within-chain split-$\hat R$, which is informative about within-chain non-stationarity but not about between-chain disagreement. For $\hat R$ across independently initialised chains, see the multi-chain section in the dynamic-effects vignette.

### Goodness of Fit

The `gof_plot` function compares observed network statistics to their posterior predictive distributions. This tells us whether the model can reproduce key structural features of the data, not just the dyad-level relationships, but emergent properties like how much actors vary in their degree (heterogeneity), whether friends-of-friends tend to be friends (transitivity), and whether ties tend to be reciprocated (dyadic dependence).

```{r overview-gof, fig.width=8, fig.height=6, dpi=100, dev="png", fig.alt="Longitudinal goodness-of-fit panels for the Dutch college fit: each facet plots one network statistic (Sender Degree Heterogeneity, Receiver Degree Heterogeneity, Dyadic Dependence, Triadic Dependence, Transitivity) across the six time periods. The observed series is a solid orange line with filled orange points; the posterior-predictive median is a dashed dark line; the 95 percent credible interval is a grey ribbon. The three elements differ on both colour and linetype, so the comparison survives grayscale printing and colour-blind viewing. Observed series falling outside the band flag structural features the model is missing."}
gof_plot(fit)
```

For a longitudinal model, `gof_plot` shows these statistics across time. The figure encodes three series on each panel using both colour and linetype redundantly: the **observed series is a solid orange line with filled orange points**, the **posterior-predictive median is a dashed dark line**, and the **95% credible interval is a grey ribbon** (the width is controlled by the `credible.level` argument). When the observed series falls inside the ribbon, the model is reproducing that aspect of the network; when it falls outside, the model is missing something. Falling outside is itself a substantive finding about the data-generating process, not a failure to report.

The panel-strip headers are the human-readable labels (`Sender Degree Heterogeneity`, etc.). `fit$GOF` is a named list whose elements are `sd.rowmean`, `sd.colmean`, `dyad.dep`, `cycle.dep` (cyclic triads), and `trans.dep`; user-facing aliases for `gof_plot(..., statistics = ...)` are `sd.row`, `sd.col`, `dyad.dep`, `triad.dep`, `trans.dep`.

Reading the figure: the Dutch college fit reproduces the average level of reciprocity and most of the cyclic-triad structure but under-predicts sender degree heterogeneity, transitivity, and (at the panel ends) receiver heterogeneity. The observed Sender Degree Heterogeneity series sits above the upper edge of the 95% band at five of the six periods (inside the band only at $t = 1$); some students nominate many more friends than the additive sender effect $a_i$ plus latent $u_i$ can absorb. Transitivity is under-predicted at every period -- the observed value sits above the upper edge of the band at $t = 1$ through $t = 6$ -- so triangle closure is systematically too low under the latent space. Receiver Degree Heterogeneity is above the band at $t = 1, 2$, inside at $t = 3$ and $t = 5$, and below the band at $t = 4$ and $t = 6$. The Triadic Dependence panel (cyclic triads, `cycle.dep`) sits inside the band at five of the six periods (outside only at $t = 1$). Dyadic Dependence (reciprocity) is the most variable: above the band at $t = 1$ and $5$ (and marginally at $t = 2$), below at $t = 3$ and $t = 6$, inside only at $t = 4$.

This is the standard signature of a latent-space model on a friendship network. AME assumes conditional dyadic independence: given the additive and multiplicative effects, dyads are independent, so triangle closure has to be soaked up by clustering of the latent positions $u_i$. ERGMs handle triangle closure directly via terms like `gwesp` or `triangle`. The two model classes are complementary. ERGM is the right answer for triangle-closure questions; AME is the right answer when the analyst wants stable, non-degenerate estimates with explicit actor heterogeneity and a coherent posterior to forecast or simulate from.

#### Posterior-predictive temporal-trend test: `gof_temporal()`

The figure read above is qualitative. For a scalar posterior-predictive p-value on the temporal *trend* in a chosen network statistic, `gof_temporal()` complements `gof_plot()`. It computes the chosen statistic at each observed period, fits an OLS slope on period index, draws posterior-predictive replicate panels via `simulate(fit)`, recomputes the slope in each replicate, and returns a two-sided Gelman-style posterior-predictive p-value defined as `2 * min(p_up, 1 - p_up)`, so it runs from 0 (observed slope in the extreme tail) up to 1 (observed slope dead-centre). A static fit on truly trending data yields `p_pp` near 0 (the observed slope is unreachable); a fit that captures the trend yields a `p_pp` well away from 0 (the observed slope is central in the predictive distribution).

```{r overview-gof-temporal}
# density rises across the panel on net (the network gets denser as
# students settle in, though the trajectory is not monotone), so it is the
# natural temporal-trend target.
gt_density <- gof_temporal(fit, stat = "density", n_rep = 100, seed = 6886)
gt_density   # prints stat, observed slope, n_rep, and the p_pp

# reciprocity is the natural second target for a directed network.
gt_recip <- gof_temporal(fit, stat = "reciprocity", n_rep = 100, seed = 6886)
gt_recip
```

A `p_pp` of 0 on `density` says the static fit cannot reproduce the empirical net densification slope (about +0.025 per period in our run); that is the expected diagnostic for a fit without `dynamic_beta` or time-varying covariates and is consistent with the per-period mass shift in `gof_plot()`. The reciprocity check tells the same story in the opposite direction: the static fit also rejects the observed reciprocity slope (`p_pp` near 0), because reciprocity actually drifts *downward* across the panel (slope about $-0.03$ per period) and the static $\rho$ cannot track that drift. Both checks are pointing at the same structural mismatch: model parameters that are constant across periods cannot reproduce period-to-period drift in either marginal quantity. The companion diagnostic `detect_change_point()` reports a heuristic tail-ratio score comparing the posterior path of each `dynamic_beta` coefficient to its AR(1) prior null. It requires a fit with `dynamic_beta` active and is therefore not applicable to the static fit shown here; the worked example lives in the [Dynamic Effects](dynamic_effects.html) vignette.

### Latent Space

The multiplicative effects capture association patterns beyond what the
covariates and additive effects explain. We fit `R = 1`, so each student's
sender position ($u_i$) and receiver position ($v_i$) is a single number --
and a one-dimensional space should be *plotted* as one dimension. (A 2-D
scatter of a 1-D space piles every point onto a line and looks degenerate
even when the positions are informative. Why not just fit `R = 2`? On these
32 students the second latent dimension competes with the three homophily
dummies for the same signal: refitting with `R = 2` collapses the
`same_male` and `same_smoker` effective sample sizes to about 15, versus
healthy mixing at `R = 1`.)

The honest display for a 1-D space is a ranked dot plot: every student's
sender and receiver position with a 95% interval, ordered by sender
position. The intervals come from the saved U/V draws
(`posterior_opts = list(save_UV = TRUE)` in the fit above).

```{r overview-uv, fig.width=7, fig.height=6.5, dpi=100, dev="png", fig.alt="Ranked dot plot of the one-dimensional latent positions for all 32 students: for each student, an orange point marks the posterior-mean sender position and a blue point the receiver position, each with a horizontal 95 percent interval; students are ordered by sender position, and a dashed vertical line marks zero."}
lp1 <- subset(latent_positions(fit), dimension == 1)
ordv <- with(subset(lp1, type == "U"), actor[order(value)])
lp1$actor <- factor(lp1$actor, levels = ordv)

ggplot(lp1, aes(x = value, y = actor, color = type)) +
	geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
	geom_errorbar(aes(xmin = value - 2 * posterior_sd,
	                  xmax = value + 2 * posterior_sd),
	              width = 0, alpha = 0.5, orientation = "y") +
	geom_point(size = 1.8) +
	scale_color_manual(values = c(U = "#D55E00", V = "#0072B2"),
	                   labels = c(U = "sender (u)", V = "receiver (v)")) +
	labs(x = "Latent position (dimension 1)", y = NULL, color = NULL) +
	theme_bw() +
	theme(panel.border = element_blank(), panel.grid.major.y = element_blank(),
	      axis.ticks = element_blank(), legend.position = "top")
```

Three things to read off this plot. First, the dimension is genuinely
informative: the extremes (around $+1.9$ at the top versus $-1.4$ at the
bottom) are separated by roughly six times the width of a typical 95%
interval, so the ordering is signal, not noise. Second, sender and receiver
positions are strongly related (their correlation is about 0.7): students
near the top both seek out and attract the same friendship cluster. Third,
the exceptions are the interesting actors -- students whose blue and orange
points diverge are attractive to a cluster they do not themselves reach
toward (or vice versa), a pattern no covariate in the model encodes.
Because this fit is static, these positions are pooled across all six
waves; if you suspect students *moved* through this space as the year
progressed, that is exactly what `dynamic_uv = TRUE` estimates (see the
[dynamic effects vignette](dynamic_effects.html)).


## Extracting Latent Positions

If you need the estimated latent positions in a tidy format (for custom plots, merging with external data, or exporting to other tools), the `latent_positions()` function returns a data frame:

```{r}
lp <- latent_positions(fit)
head(lp)
```

Each row gives one actor's position on one latent dimension at one time point; a `type` column distinguishes the sender position (`"U"`) from the receiver position (`"V"`), so a directed fit returns two rows per actor per dimension. The `posterior_sd` column is populated because the fit saved its U/V draws (`posterior_opts = list(save_UV = TRUE)`); without that option it is `NA` with a one-time message telling you how to refit. For a static fit like this one the positions do not change over time, so `latent_positions()` returns a single slice (labelled `time = 1`) that applies to every period rather than repeating it. For dynamic models (`dynamic_uv = TRUE`), positions evolve over time and `latent_positions()` returns each time point separately (see the [dynamic effects vignette](dynamic_effects.html)).

When using dynamic latent positions, the latent space is only identified up to rotation at each time point. The `procrustes_align()` function aligns each time period's positions to the previous one, removing arbitrary rotations so that trajectories across time are interpretable:

```{r}
# procrustes alignment is useful for dynamic models.
# for a static model, positions are already aligned.
aligned <- procrustes_align(fit)
str(aligned$U)
```

## Tidy-data round-trip via netify

`lame` uses the `netify` package for tidy network data. Build a netify object
from an edgelist and pass it directly to `lame()` or `ame()`. If you want to
inspect the pieces first, `netify::to_lame()` returns the `Y`, `Xdyad`,
`Xrow`, and `Xcol` objects that `lame` will use internally. Once the fit is
built, `tidy()`, `autoplot()`, and `prediction_draws_long()` give you
broom-/ggplot-/marginaleffects-ready output that works with the rest of the
tidyverse.

### Entry: tidy edgelist → `lame()`

```{r tidyverse-entry, eval = requireNamespace("netify", quietly = TRUE), message = FALSE}
library(netify)

# a tidy edgelist is often how network data arrives. Here we melt the
# Dutch-college wave list back into one (from, to, year, tie) row per
# ordered dyad, carrying a dyadic covariate along for the ride.
edges_long <- do.call(rbind, lapply(seq_along(Y), function(t) {
	Yt  <- Y[[t]]
	idx <- which(!is.na(Yt), arr.ind = TRUE)
	data.frame(
		from = rownames(Yt)[idx[, 1]],
		to   = colnames(Yt)[idx[, 2]],
		year = names(Y)[t],
		tie  = as.integer(Yt[idx]),
		same_program = Xdyad[[t]][, , "same_program"][idx]
	)
}))
head(edges_long)

netlet <- netify::netify(
	edges_long,
	actor1 = "from", actor2 = "to", time = "year",
	weight = "tie", dyad_vars = "same_program",
	symmetric = FALSE, mode = "unipartite",
	missing_to_zero = FALSE,
	output_format = "longit_array"
)

# a netify object drops straight into lame(); reduced iterations here so the
# vignette builds quickly (use burn >= 500, nscan >= 4000 for real analyses).
fit_netlet <- lame(
	netlet,
	family = "binary", R = 1,
	burn = 50, nscan = 200, odens = 5,
	verbose = FALSE
)

# to inspect what lame() will see, to_lame() returns the pieces directly
pieces <- netify::to_lame(netlet, lame = TRUE)
names(pieces)
```

Two things in that chunk deserve to be named. First, `fit_netlet` is a
live fit taken straight off the netify object -- no reshaping, no
hand-built lists -- confirming that a netify object drops into `lame()`
as-is. Second, `names(pieces)` lists the inputs `lame()` actually
consumes -- `Y`, `Xdyad`, `Xrow`, `Xcol` (the latter two are `NULL` here
because only a dyadic covariate rode along on the edgelist) -- plus the
`mode` / `family` metadata netify carries. The payoff of the round-trip
is the alignment: `netify()` took the flat
`(from, to, year, tie, same_program)` table and rebuilt, for every
period, a tie matrix and a covariate slice keyed to the same actor
order -- exactly the `rownames`-and-`array()` bookkeeping done by hand
in the setup chunk at the top of this vignette. When your data arrives
as an edgelist, as panel network data usually does, that bookkeeping is
the reason to reach for the round-trip: the edgelist is the only object
you touch, and the per-period matrices never have to be assembled by
hand.

### Exit: `tidy()`, `autoplot()`, `prediction_draws_long()`

The rest of the round-trip runs against the `fit` object built from the
Dutch-college data above:

```{r tidy-fit}
# tidy(fit) -> term / estimate / std.error / statistic / p.value /
# conf.low / conf.high; one row per coefficient (or per
# coefficient x period for dynamic_beta fits). The generic ships with
# lame so this runs without broom installed; when broom is on the path
# `broom::tidy(fit)` dispatches to the same method via
# generics::tidy registration.
tidy(fit)
```

### Coefficient tables: `glance()` + `tidy()` -> `modelsummary` / `gt` / `kableExtra`

`lame` ships both `tidy.lame()` and `glance.lame()`, registered
against `generics::tidy` / `generics::glance` at load time, so any
broom-aware table package consumes a `lame` fit directly. The most
compact route is `modelsummary::modelsummary(fit)`, which uses `tidy()`
for the coefficient panel and `glance()` for the goodness-of-fit
footer:

```{r modelsummary-fit, eval = requireNamespace("modelsummary", quietly = TRUE) && requireNamespace("broom", quietly = TRUE)}
# glance() returns the one-row model summary modelsummary uses for its
# lower panel: nobs, n_actors, n_periods, n_stored, family, mode, R,
# dynamic_uv / dynamic_ab / dynamic_beta, elpd_loo (NA unless save_log_lik=TRUE).
broom::glance(fit)

# pass the fit straight to modelsummary -- tidy() drives the upper
# (coefficients) panel, glance() drives the lower (GOF) panel.
modelsummary::modelsummary(
	list("Dutch college" = fit),
	statistic = "conf.int",
	gof_map = c("nobs", "n_actors", "n_periods", "n_stored",
	            "family", "R", "dynamic_uv", "dynamic_ab")
)
```

Multiple fits compose into a side-by-side table the same way they would
for `lm()` / `glm()` -- pass a named list, one column per fit. Below we
compare the full specification against a no-latent-space variant (`R =
0`) to see how much of the coefficient story is being soaked up by
multiplicative effects:

```{r fit-nouv, message = FALSE, warning = FALSE}
fit_noUV <- lame(
	Y = Y, Xdyad = Xdyad, Xrow = Xrow, Xcol = Xcol,
	family = "binary", rvar = TRUE, cvar = TRUE, dcor = TRUE,
	R = 0,                          # no latent space
	symmetric = FALSE,
	burn = 500, nscan = 4000, odens = 10,
	save_log_lik = TRUE,            # reused by loo_compare() below
	verbose = FALSE, plot = FALSE
)
```

```{r modelsummary-multi, eval = requireNamespace("modelsummary", quietly = TRUE) && requireNamespace("broom", quietly = TRUE), message = FALSE, warning = FALSE}
modelsummary::modelsummary(
	list("AME (R = 1)" = fit, "Additive only (R = 0)" = fit_noUV),
	statistic = "conf.int",
	gof_map = c("nobs", "n_actors", "n_periods", "n_stored",
	            "family", "R", "dynamic_uv", "dynamic_ab")
)
```

Reading down the `R = 1` column versus the `R = 0` column shows how the
latent space can redistribute signal: when residual clustering is real,
homophily coefficients move toward zero in the AME column because
structure that was loaded onto `same_program` (etc.) is instead absorbed
by the latent positions. On this small panel (32 students, short chains)
the movement is modest and not uniform across the three homophily terms,
with widely overlapping credible intervals, so treat the comparison as
illustrating the mechanism rather than as a precise decomposition; the
direction is what to look for, not the exact shift.

The same `tidy()` frame follows the broom column contract, so it drops
into `gt`, `kableExtra`, or `flextable` directly; ALS fits carry their
own `tidy()`/`glance()` methods and compose into the same tables.

```{r autoplot-fit, fig.width = 6, fig.height = 4, fig.alt="Horizontal coefficient plot returned by autoplot.lame: each row is a regression coefficient, with a point at the posterior mean and a horizontal segment marking the 95 percent credible interval; a dashed reference line at zero indicates where an effect is indistinguishable from null."}
# autoplot.lame returns a horizontal coefplot for static fits and a
# ribbon-per-period for dynamic_beta fits, so the same call works for
# both. ggplot layers compose on top.
autoplot(fit) +
	ggtitle("Dutch college friendship: posterior coefficients")
```

The coefplot is the visual form of the `tidy(fit)` table above. The
three homophily terms sit clearly to the right of the dashed zero
line -- `same_program` largest (0.58), `same_male` next (0.45),
`same_smoker` smallest (0.16) but still clear of zero -- so
same-program pairs are the most likely to befriend, holding everything
else fixed. The nodal terms are mixed: `male_row`, `male_col`, and
`smoker_row` sit clear of zero (male students are both more active
senders and more popular receivers; smokers nominate fewer friends),
`program_row` stops just short of it (the lower bound of its interval
grazes zero), and `smoker_col` and `program_col`
straddle the line -- the wide-interval caution from the coefficient
section applies to those. One glance reproduces the
coefficient-section story. The same generic dispatches on
`which = "uv"` (latent positions -- shown as the ranked dot plot in the
Latent Space section above) and `which = "ab"` (sender / receiver
random effects):

```{r autoplot-ab, fig.width = 6, fig.height = 4, fig.alt="Lollipop plot of sender random effects: one stem per student, length equal to the posterior-mean sender effect a_i; long positive stems mark students who nominate far more friends than their covariates predict."}
# which = "ab" delegates to ab_plot() (sender side by default; use
# ab_plot(fit, effect = "receiver") for the column side). Each lollipop
# is a student's sender random effect a_i: the longest positive stems
# are students who nominate far more friends than their covariates
# predict -- the heterogeneity that va summarises as one number.
autoplot(fit, which = "ab")
```

```{r pdl, message = FALSE}
# prediction_draws_long() returns a long-format data frame with
# .chain / .iteration / .draw / period / period_label / i / j /
# actor_i / actor_j / .value -- ready for tidybayes / marginaleffects.
pdl <- prediction_draws_long(fit, type = "response", n_draws = 50)
# the diagonal (i == j) is a self-tie, undefined in a friendship network, but
# the linear predictor is still numerically defined there -- drop it so the
# per-dyad summaries below cover only real (off-diagonal) pairs.
pdl <- pdl[pdl$i != pdl$j, ]
head(pdl, 3)
dim(pdl)
```

Because the `.chain` / `.iteration` / `.draw` / `.value` columns follow
the `tidybayes` / `marginaleffects` convention, this frame composes
directly with `ggdist::stat_pointinterval()`,
`tidybayes::point_interval()`, a `dplyr::group_by()` +
`summarise()` per-dyad summary, or
`marginaleffects::posterior_draws()` -- no reshaping step needed.

## Model comparison: `loo::loo_compare()` as a one-liner

When two specifications are on the table -- say the additive-effects-only
model against one that adds a multiplicative latent space -- the standard
out-of-sample yardstick is the expected log pointwise predictive
density (`elpd_loo`) returned by `loo::loo()`. Both fits need
`save_log_lik = TRUE` so `loo()` can read the pointwise
log-likelihood matrix off `fit$log_lik`; `loo::loo_compare()` then
ranks them in a single call:

```{r loo-compare-demo, eval = requireNamespace("loo", quietly = TRUE), message = FALSE, warning = FALSE}
# does the 1-D multiplicative latent space earn its keep over additive
# effects alone? both fits already carry save_log_lik = TRUE, so this
# reuses them -- no refitting. Pass a NAMED list so the rows are
# labelled by model rather than model1/model2.
cmp <- loo::loo_compare(list(no_latent = loo::loo(fit_noUV),
                             latent_R1 = loo::loo(fit)))
cmp
```

```{r loo-compare-read, echo = FALSE, eval = requireNamespace("loo", quietly = TRUE)}
best  <- rownames(cmp)[1]
worse <- rownames(cmp)[2]
dval  <- abs(round(cmp[2, "elpd_diff"], 1))
sval  <- round(cmp[2, "se_diff"], 1)
```

**How to read the output.** The top row is the preferred model and
always carries `elpd_diff = 0`. Each subsequent row reports the
*difference* in expected log pointwise predictive density relative to
the top model, with the standard error of that difference. Here
`` `r best` `` is preferred, beating `` `r worse` `` by about
`` `r dval` `` elpd units (SE `` `r sval` ``); a ratio of
`|elpd_diff|/se_diff` near or above 2 is a reasonably confident
preference, so the latent space is doing real predictive work on this
network. When `|elpd_diff|` is instead close to its SE the two
specifications are predictively indistinguishable and the simpler
model is the defensible choice. `loo()` may still flag a few dyads
with Pareto $\hat k > 0.7$ (the PSIS approximation straining); if that
warning appears for more than a handful of dyads, lengthen the chain or
use `loo::loo_moment_match()` before relying on `elpd_diff` for a
reported result.

This result sits comfortably beside the earlier side-by-side
`modelsummary` table, where the `R = 1` and `R = 0` coefficient
columns were nearly identical. The two are not in tension: the latent
space barely moves the fixed-effect point estimates yet buys roughly
`` `r dval` `` elpd units out of sample, because the predictive gain
is carried by the estimated latent positions themselves -- the
$u_i' v_j$ term in the linear predictor -- rather than by shifting
$\beta$. That is exactly what multiplicative effects are for: they
absorb residual dyadic structure (clustering, third-party dependence)
that the additive effects and covariates cannot represent, and that
structure shows up in held-out predictive density long before it
shows up in a coefficient table.

Because `family = "binary"` is one of the five families with an exact
closed-form pointwise log-likelihood (`normal`, `binary`, `cbin`,
`poisson`, `ordinal`), the absolute `elpd_loo` values here
are directly comparable to a `loo()` result from a probit-link `brms`
fit on the same data, modulo $\hat R$ and ESS being satisfactory on
the `lame` Gibbs sampler. For `frn`, see the
`log_lik_method = "observed_ghk"` note in
[Dynamic Effects](dynamic_effects.html#loo-waic-via-save_log_lik-true).

## Reproducibility: `sessionInfo()` and `renv.lock`

An analysis a reviewer (or your future self) can reproduce records the
seed and the package versions it was fit under. The pattern:

```{r repro-seed, eval = FALSE}
fit <- lame(..., seed = 6886)          # the sampler seed lives in lame()
saveRDS(list(fit = fit, session = sessionInfo()),
        file = "lame_fit_replication.rds")
```

For long-term reproducibility, commit an `renv.lock` so reviewers can
recreate the toolchain with `renv::restore()`. This vignette's own
session record:

```{r repro-sessioninfo}
sessionInfo()
```

## What's Next?

This vignette covered the core workflow: fitting, convergence, GOF, and visualization. For more specialized topics:

- **Single networks (no time series)?** See [Your First AME Model](cross_sec_ame.html) for a detailed cross-sectional walkthrough
- **Two types of nodes?** See [Bipartite Networks](bipartite.html)
- **Evolving network structure?** See [Dynamic Effects](dynamic_effects.html)
- **Just want the quick version?** See [Getting Started](lame.html)

**References**:

Hoff, PD (2021) Additive and Multiplicative Effects Network Models. Statistical Science 36, 34--50.

van de Bunt, G. G., van Duijn, M. A. J., & Snijders, T. A. B. (1999). Friendship networks through time: An actor-oriented dynamic statistical network model. *Computational & Mathematical Organization Theory*, 5(2), 167--192.

Minhas, S., Dorff, C., Gallop, M. B., Foster, M., Liu, H., Tellez, J., & Ward, M. D. (2022). Taking dyads seriously. Political Science Research and Methods, 10(4), 703--721. doi:10.1017/psrm.2021.56
